### 2018-03-28 README.Rmd
added analysis to git

---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# Exploring the effects of BCG vaccination in patients diagnosed with tuberculosis: observational study using the Enhanced Tuberculosis Surveillance system

## Abstract

### Background

Bacillus Calmette–Guérin (BCG) is one of the most widely-used vaccines worldwide. BCG primarily reduces the progression from infection to disease, however there is evidence that BCG may provide additional benefits. We aimed to investigate whether there is evidence in routinely-collected surveillance data that BCG vaccination impacts outcomes for tuberculosis (TB) cases in England.

### Methods

We obtained all TB notifications for 2009-2015 in England from the Enhanced Tuberculosis surveillance system. We considered five outcomes: All-cause mortality, death due to TB (in those who died), recurrent TB, pulmonary disease, and sputum smear status. We used logistic regression, with complete case analysis, to investigate each outcome with BCG vaccination, years since vaccination and age at vaccination, adjusting for potential confounders. All analyses were repeated using multiply imputed data.

### Results

We found evidence of an association between BCG vaccination and reduced all-cause mortality (aOR:0.76 (95%CI 0.64 to 0.89), P:0.001) and weak evidence of an association with reduced recurrent TB (aOR:0.90 (95%CI 0.81 to 1.00), P:0.056). Analyses using multiple imputation suggested that the benefits of vaccination for all-cause mortality were reduced after 10 years.

### Conclusions

We found that BCG vaccination was associated with reduced all-cause mortality in people with TB although this benefit was less pronounced more than 10 years after vaccination. There was weak evidence of an association with reduced recurrent TB.

## Repository structure

The repository is structured as follows:

- **output** Finalised paper, cover letter and checklists.
- **functions:** Functions for analysis, and document structure. Grouped into scripts by use case.
- **common:** Formatting styles that are  common to multiple documents, reference library etc.
- **sensitivity:** Various sensitivity analysis.
- **data**: Analysis output and summarised results. The raw data used for this analysis cannot be made public but see [here](https://www.samabbott.co.uk/tbinenglanddataclean/) for details of the extract and cleaning script used.
- **packrat:** R packages used for the analysis.

## Reproducing this research

### Manual install

- Intall R and Rstudio.

- Download the analysis folder from [https://github.com/seabbs/ExploreBCGOnOutcomes/archive/master.zip](https://github.com/seabbs/ExploreBCGOnOutcomes/archive/master.zip) or use `git clone`, as follows, in the command line (not the R terminal).

```{bash clone-package, eval = FALSE}
git clone https://github.com/seabbs/ExploreBCGOnOutcomes.git
```

- Once this has been downloaded click on the project file (`ExploreBCGOnOutcomes.Rproj`).

- Install the required packages using `packrat` as follows.

```{r get-packages, eval = FALSE}
install.packages("packrat")
packrat::restore()

```

- Load the analysis results by running `output/Paper.Rmd`. The sensitivity analysis results can be reproduced by setting variables in this document, as explained at the top of the script.

### Docker

This analysis can be run in a docker container based on the tidyverse docker image. To run the docker image run:

```{bash, eval = FALSE}
docker run -d -p 8787:8787 --name ExploreBCGOnOutcome -e USER=ExploreBCGOnOutcome -e PASSWORD=ExploreBCGOnOutcome seabbs/explorebcgonoutcome
```

The rstudio client can be found on port :8787 at your local machines ip. The default username:password is ExploreBCGOnOutcome:ExploreBCGOnOutcome, set the user with -e USER=username, and the password with - e PASSWORD=newpasswordhere.

### 2018-03-28 output/Paper.Rmd
added analysis to git

---
title: '**Title:** Exploring the effects of BCG vaccination in patients diagnosed with tuberculosis: observational study using the Enhanced Tuberculosis Surveillance system'
author: "Sam Abbott, Hannah Christensen, Maeve K Lalor, Dominik Zenner, Colin Campbell, Mary Ramsay, Ellen Brooks-Pollock"
csl: ../common/bmj.csl
output:
  word_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 8
    reference_docx: ../common/style.docx
  html_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 8
  pdf_document: default
  html_notebook: null
bibliography: ../common/library.bib
---

```{r setup, include=FALSE}
require(knitr)
```

```{r fig_cap, echo=FALSE}
#Load formatting functions
source('../functions/format_functions.R')

#set options for figure referencing
options(figcap.prefix = "Figure", figcap.sep = ":", figcap.prefix.highlight = "**")

#set options for table referencing
options(tabcap.prefix = "Table", tabcap.sep = ":", tabcap.prefix.highlight = "**")

##Supplementary Appendix: set options table referencing
options(suptabcap.prefix = "Supplementary table", suptab.delin = 'S', suptabcap.sep = ":", suptabcap.prefix.highlight = "**")
```

```{r run variables, include=FALSE}
##Create script of paper
#purl('Paper.Rmd', quiet=TRUE)
setwd('../output')
#Edit to save file that results are saved in - warning data files need to be added or this will fail
## main - main analysis (England only for data post 2008)
## drop_recurrent - drop all those flag for appearing twice in the ETS (drop second mention)
## restrict_UK_born - restricted to UK born only
## restrict_non_UK_born - restricted to Non-UK born only 
## elg_school_scheme - restricted to UK born that were eliblge for schools scheme cut is at 15 assuming elb closed in 2005
## Note restricting by UK birth status will mean that the paper cannot be knit and instead needs to be run as a script

folder <- 'main'
direct_data <- '~/data/tb_data/ETS/ExploreBCGOnOutcome/'

## should imputation be run as a sensitivity analysis?
## Turn off imputation for all analysis except for main
imput <- TRUE

##Create directories if they do not exist
ifelse(!dir.exists(paste0('../data/', folder)), dir.create(paste0('../data/', folder)), FALSE)
ifelse(!dir.exists(paste0(direct_data, folder)), dir.create(paste0(direct_data, folder)), FALSE)
```

```{r load packages, include=FALSE}
# Packages to load --------------------------------------------------------
source('../scripts/packages.R')
```

```{r load data, message=FALSE, warning=FALSE, include=FALSE}
# Load Cleaned data -------------------------------------------------------

#Load Cleaned data used by other analysis
## See https://www.samabbott.co.uk/tbinenglanddataclean/ for details of raw data cleaning and processing.

#Convert to tibble object
uk_not <- readRDS('~/data/tb_data/tbinenglanddataclean/clean_ets_2016.rds')

# sample from data for testing of imputation: needs to be commented out to run actual analysis
##uk_not<- uk_not[sample(nrow(uk_not), 40000), ]

## Change variable names in line with analysis (due to updated data)
uk_not <- uk_not %>% 
              dplyr::rename(yrsinceBCG = YrSinceBCG, Mortality = mortality, SuccTreat = succTreat, AgeatBCG = age_bcg) %>% 
                dplyr::mutate(TB_mortality = DeathDueTB)

#Restriction based on country of notification - to England
uk_not <- uk_not %>% dplyr::filter(country %in% c('England'))

## Restrict data to post the introduction of national BCG status
uk_not <- uk_not %>% filter(year > 2008)

## Define nested population for imputation
uk_not_nonested <- uk_not

## Define eligibility for the schools scheme
## Those that are UK born
## Assume everyone who should have been vaccinated in the scheme were vaccinated at the latest when they were 15
## Assume the scheme ran until 2005 (but not in 2005)
## Assume that only the UK born were eligible for the scheme
restrict_elg_school_scheme = function(df)
{
 df_elg <-  map(unique(df$year), function(x) {
    if (x > 2004)
    {
     df_elg <- df %>% 
                 dplyr::filter(year ==  x,age >= 14 + x - 2004) %>% 
                     filter(ukborn == 'UK Born')
    }else {
      df_elf <- df
    }
   return(df_elg)
  }) %>% bind_rows

 return(df_elg)
}

## Restiction function for data
  restrict_df = function(df, folder)
  {
    # Drop recurrent
    if (folder == 'drop_recurrent') {
          df_t <- df %>% dplyr::filter(is.na(recurrent))
    }else if (folder == 'restrict_UK_born') {
      #Restriction to UK born
      df_t <- df %>% dplyr::filter(ukborn %in% 'UK Born')
    }else if (folder == 'restrict_non_UK_born') {
      #Restiction to Non-uk-born and redefining cob with lower degrees of freedom
      df_t <- df %>% dplyr::filter(ukborn %in% 'Non-UK Born')
    }else if (folder == 'elg_school_scheme') {
      df_t <- df %>% restrict_elg_school_scheme
    }else {
        df_t <- df
      }
    return(df_t)
  }

## Restrict data based on sensitivty
uk_not <- restrict_df(uk_not, folder)

## Evaluation of population stats
elg_school_stats <- function(df)
{
  ## Year of vaccination prior to the schools scheme ending
  df %>% dplyr::filter(bcgvaccyr > 2004) %>% nrow -> vac_after_schools
  df %>% dplyr::filter(bcgvaccyr < 2005) %>% nrow -> vac_during_schools
  df %>% dplyr::filter(is.na(bcgvaccyr)) %>% nrow -> vac_yr_miss

  ##Age for elibigily to schools scheme
  df %>% restrict_elg_school_scheme %>%  nrow -> eli_cases

  df %>% dplyr::filter(ukborn == 'UK Born') %>% nrow -> uk_cases

  ##output
  output <- c(vac_after_schools, vac_during_schools, vac_yr_miss, eli_cases, uk_cases)
  names(output) <- c('vaccinated after school', 'vaccinated during school', 'year of vaccination missing', 'cases eligible for the schools scheme', 'uk born cases')

  return(output)
}

elg_school_stats_sum <- uk_not %>% elg_school_stats
```

```{r analysis variables, include=FALSE}
# Define Analysis Variables -----------------------------------------------
stew(file = paste0('../data/', folder, '/exposure_outcome_defs.rda'),{
  #Define outcomes to analysis - all outcomes selected are binary
outcomes <-  c('Mortality', 'TB_mortality', 'prevdiag','pulmextrapulm', 'sputsmear')

#Define Variables to analysis- BCG status is binary and Years since BCG is continous
variables <- c('bcgvacc', 'yrsinceBCG', 'ageatvac')

#Define Covariates
covariates <- c('age', 'sex', 'natquintile', 'ukborn', 'ethgrp', 'year')

#Define variables to fit to splines
splines <- ('age')

#Define outcome in words
outcomes_words <- c('All-cause mortality', 'Death due to TB (in those who died*)', 'Recurrent TB', 'Pulmonary TB', 'Sputum smear status - positive')
names(outcomes_words) <- outcomes

#Define Variables in words
variables_words <- c('BCG vaccination', 'Years since BCG vaccine', 'Age at vaccination')
names(variables_words) <- variables

#Define Covariates in words
covariates_words <- c('Age', 'Sex', 'IMD rank (with 1 as most deprived and 5 as least deprived)', 'UK born', 'Ethnic group', 'Calendar year')
names(covariates_words) <- covariates

#Define splines in words
splines_words <- c('age')
names(splines_words) <- splines
})

## Sensitivity analysis
if (folder %in% c('restrict_UK_born', 'restrict_non_UK_born', 'elg_school_scheme')) {
    source('../sensitivity/params_UK_born.R')
}

## TB death footer
tb_death_footer <- "Death due to TB in those who died and where cause of death was known"
```

```{r load functions, message=FALSE, warning=FALSE, include=FALSE}
# Load functions required for the analysis --------------------------------
source('../functions/main_functions.R')
source('../functions/analysis_functions.R')
source('../functions/summary_functions.R')
source('../functions/plot_functions.R')
source('../functions/demographic_functions.R')
```

```{r main analysis, message=FALSE, warning=FALSE, include=FALSE}
#Run the regression analysis - saved in script
source('../scripts/regression_analysis.R')
```

```{r Multiple imputation data, message=FALSE, echo=FALSE, results='asis', warning=FALSE}
## define variables to impute
imputed_var <- c(outcomes, covariates, variables, 'phec')
#Run imputation script - analysis for imputed dataset
if (imput) {
  source('../scripts/imp_data.R')
}
```

```{r explore-imp-missing-year-vac, echo = FALSE, eval = FALSE}
if (imput) {
     source('../sensitivity/explore_imputation_yrsinceBCG.R') 
}
```

```{r Multiple imputation analysis, message=FALSE, echo=FALSE, results='asis', warning=FALSE}
#Run imputation script - analysis for imputed dataset
if (imput) {
    source('../sensitivity/analysis_imp.R')
}
```

```{r Predict the number of cases prevented, message = FALSE, echo = FALSE, warning = FALSE}
if (folder %in% "main") {
#Run script to determine predicted impact of vaccination
source('../sensitivity/NoCasesPrevented.R')
}
```

**Authors:**

Sam Abbott, Bristol Medical School: Population Health Sciences, University of Bristol, Bristol, UK

Hannah Christensen, Bristol Medical School: Population Health Sciences, University of Bristol, Bristol, UK

Maeve K Lalor, TB section, National Infection Service, Public Health England, London, UK

Dominik Zenner, Respiratory Diseases Department, Centre for Infectious Disease Surveillance and Control (CIDSC), National Infections Service, Public Health England, London, UK

Colin Campbell, National Infection Service, Public Health England, London, UK

Mary E Ramsay, Immunisation, Hepatitis and Blood safety department, Public Health England, London, UK

Ellen Brooks-Pollock, Bristol Medical School: Population Health Sciences, University of Bristol, Bristol, UK

**Correspondence to:** Sam Abbott, Bristol Medical School: Population Health Sciences, University of Bristol, Bristol BS8 2BN, UK; sam.abbott@bristol.ac.uk; 01173310185

**WORD COUNT:** **Title** 19 (20), **Abstract** 220 (250), **Paper** 2740 (3500)

##### PAGEBREAK

**ABSTRACT**

**Background**

Bacillus Calmette–Guérin (BCG) is one of the most widely-used vaccines worldwide. BCG primarily reduces the progression from infection to disease, however there is evidence that BCG may provide additional benefits. We aimed to investigate whether there is evidence in routinely-collected surveillance data that BCG vaccination impacts outcomes for tuberculosis (TB) cases in England.

**Methods**

We obtained all TB notifications for 2009-2015 in England from the Enhanced Tuberculosis surveillance system. We considered five outcomes: All-cause mortality, death due to TB (in those who died), recurrent TB, pulmonary disease, and sputum smear status. We used logistic regression, with complete case analysis, to investigate each outcome with BCG vaccination, years since vaccination and age at vaccination, adjusting for potential confounders. All analyses were repeated using multiply imputed data.

**Results**

We found evidence of an association between BCG vaccination and reduced all-cause mortality (aOR:`r table_var_outs[[1]][2,5][[1]] %>% pretty_OR(CINote = "95%CI ")`, P:`r table_var_outs[[1]][1,6][[1]]`) and weak evidence of an association with reduced recurrent TB (aOR:`r table_var_outs[[1]][6,5][[1]] %>% pretty_OR(CINote = "95%CI ")`, P:`r table_var_outs[[1]][5,6][[1]]`). Analyses using multiple imputation suggested that the benefits of vaccination for all-cause mortality were reduced after 10 years.

**Conclusions**

We found that BCG vaccination was associated with reduced all-cause mortality in people with TB although this benefit was less pronounced more than 10 years after vaccination. There was weak evidence of an association with reduced recurrent TB.

**Keywords:** Tuberculosis, BCG, Surveillance

**What is the key question?**

Is there evidence in the Enhanced Tuberculosis (TB) Surveillance System that BCG vaccination improves TB outcomes?

**What is the bottom line?**

This analysis suggests that vaccination with BCG is associated with reduced all-cause mortality for TB patients and potentially a reduced chance of having a repeat episode.

**Why read on?**

BCG vaccination could reduce all-cause mortality in TB cases and may decrease the number of repeat episodes in low incidence countries.

##### PAGEBREAK

**INTRODUCTION**

Bacillus Calmette–Guérin (BCG) is one of the mostly widely-used vaccines and the only vaccine that protects against tuberculosis (TB) disease. BCG was first used in humans in 1921 and was introduced into the WHO Expanded Program on Immunization in 1974.[@The2004] BCG vaccination has been controversial due to its variable efficacy and possibility of causing a false positive result with the standard skin test for TB.[@Zwerling2011a] However, the lack of a more effective vaccine and the emergence of drug-resistant TB strains means that BCG remains the best available vaccination for TB.

BCG’s primary mode of action is to directly prevent the development of active, symptomatic disease. Its efficacy in adults is context specific, with estimates ranging between 0% and 78%.[@Mangtani2014a] Efficacy has been shown to be dependent on previous exposure, with unexposed individuals receiving the greatest benefit. Unlike in adults, BCG has consistently been shown to be highly protective against TB and TB meningitis in children.[@Rodrigues1993; @Colditz1994] For this reason the majority of countries that use BCG vaccinate at birth.[@Zwerling2011] Adult vaccination is no longer common, in the UK where universal BCG vaccination of adolescents was stopped in 2005 in favour of a targeted neonatal programme aimed at high risk children.

Vaccination policy has been primarily based on reducing the incidence of active TB and little attention has been given to any additional effects of BCG.[@Fine2005a; @Teo2006] There is some evidence that BCG vaccination induces innate immune responses which may provide non-specific protection,[@Kleinnijenhuis2012] TB patients with BCG scars were found to respond better to treatment with earlier sputum smear conversion,[@Jeremiah2010] and there is evidence to suggest that BCG vaccination is associated with reduced all-cause neonatal mortality[@Higgins] and both reduced TB[@Abubakar2013] and all-cause[@Rieckmann2016] mortality in the general population. Given that the immunology behind TB immunity is not well understood these findings suggest that BCG may play a more important role in improving TB outcomes than previously thought. We aimed to quantify the effects of BCG vaccination on outcomes for individuals with notified TB in England using routinely collected surveillance data to provide evidence for appropriate public health action and provision. Where we found an association, we additionally explored the role of years since vaccination, and age at vaccination.

#####PAGEBREAK

**METHOD**

**Enhanced Tuberculosis Surveillance (ETS) system**

We extracted all notifications from the Enhanced Tuberculosis Surveillance (ETS) system from January 1, 2009 to December 31, 2015. BCG vaccination status and year of vaccination have been collected since 2008. We considered five TB outcomes which were selected due to their association with increased case infectiousness or poor consequences for patients (`r tabRef('OutcomeDef')`).

```{r outcome-defs, warning = FALSE, message = FALSE, echo = FALSE}
OutcomeDefsTab <- data_frame(Outcome = c(
                                         'All-cause mortality',
                                         'Death due to TB (in those who died)',
                                         'Recurrent TB',
                                         'Pulmonary disease', 
                                         'Sputum smear status - positive'
                                         ),
                             Definition = c('All-cause mortality was defined using the overall outcome recorded in ETS, this is based on follow up at 12, 24, and 36 months (or until treatment completion). If a case dies within 12 months of completing treatment, from TB or from a cause related to TB then their overall outcome will be updated in the ETS. Those that were lost to follow up, or not evaluated were treated as missing.',
                                            'For cases with a known cause of death, death due to TB is defined as those that died from TB, or where TB had contributed to their death.',
                                           'TB cases who had recurrent episodes were identified in the dataset.',
                                            'Cases that present with pulmonary TB (regardless of extra-pulmonary TB status).',
                                            'Status of sputum sample tested for Acid-Fast Bacilli.'))

OutcomeDefsTab %>% 
  kable(align = 'll', caption = tabRef('OutcomeDef', 'Summary of outcome definitions and rationale for inclusion '))
```

**Exposure variables relating to BCG**

We included three exposure variables related to BCG: BCG status (vaccinated, yes/no), years since vaccination and age at vaccination.

BCG status was taken directly from the ETS. Years since BCG vaccination was defined as year of notification minus year of vaccination and categorised into two groups (0 to 10 and 11+ years), based on evidence that the average duration of BCG protection is 10-15 years.[@Abubakar2013] Age at vaccination is defined in the online supplementary information.

```{r plots of original var dist, warning = FALSE, echo = FALSE, message = FALSE, eval = FALSE}
## Plot of years since BCG vaccination distribution
ggplot(uk_not %>% filter(!is.na(ukborn), !is.na(yrsinceBCG)), aes(x = yr_bcg, group = ukborn, colour = ukborn, fill = ukborn)) +
  geom_bar() + 
    facet_grid(ukborn ~ yrsinceBCG, scales = 'free')

## Plot of age at vaccination
ggplot(uk_not %>% filter(!is.na(ukborn), !is.na(ageatvac)), aes(x = AgeatBCG, group = ukborn, colour = ukborn, fill = ukborn)) +
  geom_bar() + 
    facet_grid(ukborn ~ ageatvac, scales = 'free')
```

**Statistical Analysis**

R was used for all statistical analysis.@R The analysis was conducted in two stages. Firstly, we calculated proportions for all demographic and outcome variables, and compared vaccinated and unvaccinated TB cases using the $\chi ^2$ test. Secondly, we used logistic regression, with complete case analysis, to estimate the association between exposures and outcome variables, both with and without adjustment for confounders.

In the multivariable models, we adjusted for sex,[@Parslow2001; @Roth2006a; @Aaby2014] age,[@Teale1993] Index of Multiple Deprivation (2010) categorised into five groups for England (IMD rank),[@DCLG2011; @Bhatti1995] ethnicity,[@Parslow2001; @Abubakar2008] UK birth status,[@French2007; @Djuretic2002] and year of notification. As the relationship between age and outcomes was non-linear, we modelled age using a natural cubic spline with knots at the 25%, 50% and 75% quantiles.

We conducted sensitivity analyses to assess the robustness of the results, by dropping each confounding variable in turn and assessing the effect on the adjusted Odds Ratios (aORs) of the exposure variable. We repeated the analysis excluding duplicate recurrent cases, and restricting the study population to those eligible for the BCG schools scheme (defined as UK born cases that were aged 14 or over in 2004) to assess the comparability of the BCG vaccinated and unvaccinated populations. To mitigate the impact of missing data we used multiple imputation, with the MICE package.[@VanBuuren2011] We imputed 50 data sets (for 20 iterations) using all variables included in the analysis as predictors along with Public Health England centre. The model results were pooled using the small sample method,[@Barnard1999] and effect sizes compared with those from the main analysis.

**RESULTS**

**Description of the data**

There were `r format(nrow(uk_not),big.mark=",",scientific=FALSE)` TB notifications between 2009-2015 in England. Reporting of vaccination status and year of vaccination improved over time: `r round(sum(!is.na(uk_not[uk_not$year %in% as.character(2009:2012),'bcgvacc']))/nrow(uk_not[uk_not$year %in% as.character(2009:2012),])*100, digits=1)`% (`r sum(!is.na(uk_not[uk_not$year %in% as.character(2009:2012),'bcgvacc']))`/`r nrow(uk_not[uk_not$year %in% as.character(2009:2012),])`) of notifications included vaccination status for 2009 to 2012, increasing to `r round(sum(!is.na(uk_not[uk_not$year %in% as.character(2013:2015),'bcgvacc']))/nrow(uk_not[uk_not$year %in% as.character(2013:2015),])*100, digits=1)`% (`r sum(!is.na(uk_not[uk_not$year %in% as.character(2013:2015),'bcgvacc']))`/`r nrow(uk_not[uk_not$year %in% as.character(2013:2015),])`) from 2013 to 2015. The majority of cases that had a known vaccination status were vaccinated (`r round(length(uk_not$bcgvacc[(uk_not$bcgvacc %in% 'Yes' & !is.na(uk_not$bcgvacc))])/sum(!is.na(uk_not$bcgvacc))*100, digits=1)`%, `r length(uk_not$bcgvacc[(uk_not$bcgvacc %in% 'Yes' & !is.na(uk_not$bcgvacc))])`/`r sum(!is.na(uk_not$bcgvacc))`), and where age and year of vaccination was known, the majority of cases were vaccinated at birth (`r round(summary(uk_not$ageatvac)[1]/sum(!is.na(uk_not$ageatvac)), digits =1)*100`%, `r summary(uk_not$ageatvac)[1]`/`r sum(!is.na(uk_not$ageatvac))`).

```{r demographics breakdown, message=FALSE, warning=FALSE, echo=FALSE}
#Descriptive data analysis - saved in script
if (!(folder %in% c('restrict_UK_born', 'restrict_non_UK_born', 'elg_school_scheme'))) {
  source('../scripts/data_descript.R')
} 
```

Vaccinated cases were younger than unvaccinated cases on average (median age `r age_bcg_yes[2]` years (IQR `r age_bcg_yes[1]` to `r age_bcg_yes[3]`) compared to `r age_bcg_no[2]` years (IQR `r age_bcg_no[1]` to `r age_bcg_no[3]`)). A higher proportion of non-UK born cases were BCG vaccinated, (`r ukborn_prop_BCG[1,4] `%, `r ukborn_prop_BCG[1,2]`/`r ukborn_prop_BCG[1,6]`) compared to UK born cases (`r ukborn_prop_BCG[2,4] `%, `r ukborn_prop_BCG[2,2]`/`r ukborn_prop_BCG[2,6]`, P: `r ukborn_prop_BCG[2,9]`) and, of those vaccinated, a higher proportion of non-UK born cases were vaccinated at birth compared to UK born cases (`r ukborn_prop_vcbirth[2]` vs. `r ukborn_prop_vcbirth[1]` respectively, P: `r ukborn_test_vcbirth`). See online `r SuptabRef('outcomesum')` for the breakdown of outcome variables and `r SuptabRef('ConfounderStrat')` for the breakdown of confounding variables.

```{r particpants summary, echo = FALSE, message = FALSE, warning = FALSE}
bake(file = paste0('../data/', folder, '/participants.rds'), {

  ##  Filter data for only that were vaccinated (recorded as vaccinated)
uk_not_vac <- uk_not %>%  filter(bcgvacc %in% 'Yes')
out <- map2(variables, list(uk_not, uk_not_vac, uk_not_vac), function(variable, df){

  out <- outcomes %>% map_df(function(outcome, var = variable, df_out = df ){
    if (outcome %in% "TB_mortality") {
      df_out <- df_out %>% filter(Mortality %in% 'Yes')
    }
  #Find number with variable and outcome - with adjustments
uni <- sum(complete.cases(df_out[,c(outcome, var)]))
adj <- sum(complete.cases(df_out[,c(outcome, var, covariates)]))

#Present percentage missing from main data set
uni <- paste(uni, ' (', pretty_round(uni/nrow(df_out)*100,digits = 0), '%)',sep = '')
adj <- paste(adj, ' (', pretty_round(adj/nrow(df_out)*100,digits = 0), '%)',sep = '')

out  <- data_frame(univariable = uni, Adjusted = adj)
return(out)
})
  return(out)
})

participants_table <- bind_cols(data_frame(Outcome = outcomes_words), do.call(bind_cols, out))
colnames(participants_table) <- c('Outcomes', 'BCG', 'BCGAdj', 'YearOfVac', 'YearOfVacAdj', 'AgeAtVac', 'AgeAtVacAdj')

#Temp to assign variable for saving
tmp <- participants_table 
}) -> participants_table
```

```{r split out percentage odds, echo=FALSE, message=FALSE}
stew(file = paste0('../data/', folder, '/OR_for_text.rda'), {

pretty_OR_P <- function(OR, P) {
   pretty_OR(OR) %>% 
  str_replace("\\)", paste0(", P: ", P, ")"))
}

## OR Mortality
OR.mort <- pretty_OR_P(table_var_outs[[1]][2,3][[1]], table_var_outs[[1]][1,4][[1]])

## aOR difference Mortality
aOR.mort <- pretty_OR_P(table_var_outs[[1]][2,5][[1]], table_var_outs[[1]][1,6][[1]])

## Get OR for estimating impact of everyone being vaccinated
aOR.mort.split <- strsplit(table_var_outs[[1]][2,5][[1]], "") %>% unlist

aOR.prog <- list(OR = aOR.mort.split[1:4], 
                 LCI = aOR.mort.split[7:10], 
                 UCI = aOR.mort.split[15:18]) %>% 
  map(~paste(., collapse = "")) %>% 
  map_dfc(~as.numeric(.)) %>% 
  mutate(outcome = "Mortality")

## aOR recurrent TB
aOR.prev <- pretty_OR_P(table_var_outs[[1]][6,5][[1]], table_var_outs[[1]][5,6][[1]])

})

```

```{r link to predicted avg impact per year, message = FALSE, warning = FALSE, echo = FALSE}
if (folder %in% "main") {
  AllMortPred <- AvgYearlyPredictImpact %>% 
                filter(outcome %in% 'Mortality') %>%
                 pull(pretty_eff_year)
}
```

**All-cause mortality**

In the univariable analysis the odds of death from any cause were lower for BCG vaccinated TB cases compared to unvaccinated cases, with an OR of `r OR.mort` (`r tabRef('BCGoutsum')`, see `r SuptabRef('bcg_mortality_tab')` for the full table); an association remained after adjusting for confounders, but was attenuated with an aOR of `r aOR.mort`.  We estimate that if all unvaccinated cases had been vaccinated there would have been on average `r AllMortPred ` fewer deaths per year during the study period (out of `r unvac_mort_per_year` deaths per year on average in unvaccinated cases). Whilst there was evidence in univariable analyses to suggest all-cause mortality was higher in persons vaccinated more than 10 years prior to notification of TB and that all-cause mortality increased with increasing age group, these disappeared after adjusting for potential confounders (`r tabRef('YrsinceBCGoutsum')`, `r SuptabRef('ageatvacoutsum')`).

Similar results to the multivariable analysis were found using multiply imputed data for the association between vaccination status and all-cause mortality (aOR: `r sum_imp_bcg[1, 5][[1]]  %>%  pretty_OR`, P:  `r sum_imp_bcg[1, 6][[1]]`), but not for time since vaccination with a greatly increased risk of all-cause mortality estimated for those vaccinated more than 10 years before case notification, compared to those vaccinated more recently (aOR: `r sum_imp_yrsinceBCG[1, 5][[1]] %>%  pretty_OR`, (see online `r SuptabRef('sum_imp_bcg')`, `r SuptabRef('sum_imp_yrsinceBCG')`)). For age at vaccination results for the multivariable analysis using multiply imputed data were comparable to those found using complete case analysis, except that there was some evidence that vaccination in adolescence, compared to under 1, was associated with increased, rather than decreased, all-cause mortality (aOR: `r sum_imp_ageatvac[2, 6][[1]] %>%  pretty_OR`, `r SuptabRef('sum_imp_ageatvac')`).

**Deaths due to TB (in those who died)**

There was little evidence of any association between BCG vaccination and deaths due to TB (in those who died and where cause of death was known) in the univariable analysis (`r tabRef('BCGoutsum')`). The adjusted point estimate indicated an association between BCG vaccination and reduced deaths due to TB (in those who died) although the confidence intervals remained wide with a similar result found using multiply imputed data (see online `r SuptabRef('sum_imp_bcg')`). There were insufficient data to robustly estimate an association between deaths due to TB (in those who died) and years since vaccination or age at vaccination (`r tabRef('YrsinceBCGoutsum')`, `r SuptabRef('ageatvacoutsum')`).

**Recurrent TB**

In both the univariable and multivariable analysis there was some evidence that BCG vaccination was associated with reduced recurrent TB, although the strength of the evidence was weakened after adjusting for confounders (`r tabRef('BCGoutsum')`). In the adjusted analysis, the odds of recurrent TB were lower for BCG vaccinated cases compared to unvaccinated cases, with an aOR of `r aOR.prev`. The strength of the evidence for this association was comparable in the analysis using multiply imputed data (see online `r SuptabRef('sum_imp_bcg')`). There was little evidence in the adjusted analysis of any association between recurrent TB and years since vaccination (`r tabRef('YrsinceBCGoutsum')`) or age at vaccination (`r SuptabRef('ageatvacoutsum')`).

**Other Outcomes**

After adjusting for confounders there was little evidence for any association between BCG vaccination and pulmonary disease or positive sputum smear status (`r tabRef('BCGoutsum')`); similar results were found using multiply imputed data (see online `r SuptabRef('sum_imp_bcg')`).

```{r BCG vaccination summary table, message=FALSE, echo=FALSE, results='asis', warning=FALSE}
table_var_outs[[1]]$Outcome <- stringr::str_replace(as.character(table_var_outs[[1]]$Outcome), "[*]", "\u2021")

table_var_outs[[1]] %>% 
  combine_model_data_sum(DataSum = ModelDataSum[[1]]) %>% 
            pretty_sum_results_table(col_names = c('Outcome', 'BCG vaccinated', 'Cases*', 'Cases with outcome (%)', 'OR (95% CI)', 'P-value', 'Cases\u2020', 'Cases with outcome (%)', 'aOR (95% CI)', 'P-value'), 
                                     footer = paste0('OR (95% CI): unadjusted odds ratio with 95% confidence intervals, aOR (95% CI): adjusted odds ratios with 95% confidence intervals, ', sum_sample_size('BCG', VarAdj = 'BCGAdj'), ", \u2021 ", tb_death_footer), 
                                     label = 'BCGoutsum', 
                                     caption = 'Summary of associations between BCG vaccination and all outcomes')
```

```{r Years since Vaccination summary table, message=FALSE, echo=FALSE, results='asis', warning=FALSE}
## For deaths due to TB (in those who died) no data so set model outputs to null
table_var_outs[[2]][4, c(3,5)] <- "*Insufficient data*"
table_var_outs[[2]][3, c(4,6)] <- "-"

table_var_outs[[2]]$Outcome <- stringr::str_replace(as.character(table_var_outs[[2]]$Outcome), "[*]", "\u2021")

table_var_outs[[2]] %>% 
  combine_model_data_sum(DataSum = ModelDataSum[[2]]) %>% 
            pretty_sum_results_table(col_names = c('Outcome', 'Years since BCG', 'Cases*', 'Cases with outcome (%)', 'OR (95% CI)', 'P-value', 'Cases\u2020', 'Cases with outcome (%)', 'aOR (95% CI)', 'P-value'), 
                                     footer = paste0('OR (95% CI): unadjusted odds ratio with 95% confidence intervals, aOR (95% CI): adjusted odds ratios with 95% confidence intervals, ', sum_sample_size('YearOfVac', VarAdj = 'YearOfVacAdj', PerDescrip = '% of vaccinated cases'),
                                                     ", \u2021 ", tb_death_footer), 
                                     label = 'YrsinceBCGoutsum', 
                                     caption = 'Summary of associations between years since vaccination and all outcomes in individuals who were vaccinated. The baseline exposure is vaccination $\u2264 10$ years before diagnosis compared to vaccination $11+$ years before diagnosis. Deaths due to TB (in those who died) had insufficient data for effect sizes to be estimated in both the univariable and multivariable analysis.'
                                     )
```

```{r Covariate test, message=FALSE, echo=FALSE, results='asis', warning=FALSE, eval = FALSE}
#run analysis with each covaraite dropped - saved in script
  source('../sensitivity/Covariate_test.R')
```

```{r table testing cov bcgvacc, include = FALSE, eval = FALSE}
bcgvacc_cov_ef %>% kable
```

**Sensitivity analysis**

Dropping duplicate recurrent TB notifications increased the magnitude, and precision, of the effect sizes for recurrent TB, all-cause mortality, and deaths due to TB (in those who died) (see online `r SuptabRef('BCGoutsumsens')`). Restricting the analysis to only cases that were eligible for the BCG schools scheme reduced the sample size of the analysis (from an initial study size of `r nrow(uk_not)`, of which `r uk_not %>% filter(ukborn %in% 'UK Born') %>% nrow` were UK born, to `r elg_school_stats_sum['cases eligible for the schools scheme'] %>% as.numeric` cases that would have been eligible for the BCG schools scheme). With this reduced sample size, there was strong evidence in adjusted analyses of an association between BCG vaccination and reduced recurrent TB, and evidence of an association with decreased all-cause mortality (see online `r SuptabRef('BCGoutsumsens')`).

#####PAGEBREAK

**DISCUSSION**

```{r Summary stats for discussion, message = FALSE, echo = FALSE}
MissDataBCG <- paste0(round(sum(is.na(uk_not[uk_not$year %in%  as.character(2009:2015),'bcgvacc']))/nrow(uk_not[uk_not$year %in%  as.character(2009:2015),])*100, digits  = 0),'%  (',sum(is.na(uk_not[uk_not$year %in%  as.character(2009:2015),'bcgvacc'])),'/', nrow(uk_not[uk_not$year %in%  as.character(2009:2015),]), ')')

MissDataYearBCG <- paste0(round(sum(is.na(uk_not[(uk_not$year %in%  as.character(2009:2015) & uk_not$bcgvacc %in% 'Yes'),'ageatvac']))/nrow(uk_not[(uk_not$year %in% as.character(2009:2015) & uk_not$bcgvacc %in% 'Yes'),])*100, digits =  0), '%, (', sum(is.na(uk_not[(uk_not$year %in%  as.character(2009:2015) & uk_not$bcgvacc %in% 'Yes'),'ageatvac'])), '/', nrow(uk_not[(uk_not$year %in% c('2009', '2010', '2011', '2012', '2013', '2014') & uk_not$bcgvacc %in% 'Yes'),]), ')')
```

Using TB surveillance data collected in England we found that BCG vaccination, prior to the development of active TB, was associated with reduced all-cause mortality and fewer recurrent TB cases, although the evidence for this association was weaker. There was some suggestion that the association with all-cause mortality was due to reduced deaths due to TB (in those who died), though the study was underpowered to definitively assess this. We did not find evidence of an association between BCG status and positive smear status or pulmonary TB. Analysis with multiply imputed data indicated that notification 10+ years after vaccination was associated with increased all-cause mortality. In separate analyses, there was some evidence that vaccination at birth, compared to at any other age, was associated with reduced all-cause mortality, and increased deaths due to TB (in those who died).

This study used a large detailed dataset, with coverage across demographic groups, and standardized data collection from notifications and laboratories. The use of routine surveillance data means that this study would be readily repeatable with new data. The surveillance data contained multiple known risk factors, this allowed us to adjust for the confounding present in the univariable analysis, which attenuated the evidence for an association with BCG vaccination for all outcomes. However, there are important limitations to consider. The study was conducted within a population of active TB cases, therefore the association with all-cause mortality cannot be extrapolated to the general population. Additionally, vaccinated and unvaccinated populations may not be directly comparable because vaccination has been targeted at high-risk neonates in the UK since 2005. We mitigated this potential source for bias by conducting a sensitivity analysis including only those eligible for the universal school age scheme, and whilst the strength of associations were attenuated there remained some evidence of improved outcomes. Sensitivity analysis excluding recurrent cases indicated their inclusion may have biased our results towards the null.

Variable data completeness changed with time, with both BCG vaccination status and year of vaccination having a high percentage of missing data, which may not be missing completely at random. We therefore checked the robustness of our results with multiple imputation including regional variability, however an unknown missing not at random mechanism, or unmeasured confounding may still have introduced bias. We found a greatly increased risk of all-cause mortality for those vaccinated more than 10 years ago in the analysis with multiply imputed data, compared to the complete case analysis. This is likely to be driven by a missing not at random mechanism for years since vaccination, with older cases being both more likely to have been vaccinated more than 10 years previously and to also have an unknown year of vaccination.  The high percentage of missing data also means that we were likely to be underpowered to detect an effect of BCG vaccination on sputum smear status and deaths due to TB (in those who died), with years since vaccination, and age at vaccination likely to be underpowered for all outcomes. We were not able to adjust for either tuberculin skin test (TST) stringency, or the latitude effect, although we were able to adjust for UK birth status.[@Roy2014b] However, the bias induced by these confounders is likely to be towards the null, meaning that our effect estimates are likely to be conservative. Finally, BCG vaccination status may be subject to misclassification due to recall bias; validation studies of the recording of BCG status in the ETS would be required to assess this.

Little work has been done to assess the overall effect of BCG on outcomes for active TB cases although the possible non-specific effects of BCG are an area of active research.[@Higgins; @Kandasamy2016; @Pollard2017] Whilst multiple studies have investigated BCG's association with all-cause mortality, it has been difficult to assess whether the association continues beyond the first year of life.[@Pollard2017] The effect size of the association we identified between BCG and all-cause mortality in active TB cases was comparable to that found in a Danish case-cohort study in the general population (aHR: 0.58 (95% CI 0.39 to 0.85).[@Rieckmann2016] A recent systematic review also found that BCG vaccination was associated with reduced all-cause mortality in neonates, with an average relative risk of 0.70 (95% CI 0.49 to 1.01) from five clinical trials and 0.47 (95% CI 0.32 to 0.69) from nine observational studies at high risk of bias.[@Higgins] We found some weak evidence that BCG vaccination was associated with reduced deaths due to TB (in those who died), although our point estimate had large confidence intervals. Several meta-analyses have found evidence supporting this association,[@Colditz1994; @Abubakar2013] with one meta-analysis estimating a 71% (RR: 0.29 95% CI 0.16 to 0.53) reduction in deaths due to TB.[@Colditz1994] Our study was limited by the relative rarity of deaths due to TB in England, whereas other approaches benefited from the use of meta-analysis. This study could not determine the possible causal pathway for the association between BCG vaccination all-cause mortality, and recurrent TB. These are important to establish in order to understand the effect of BCG vaccination on TB outcomes.

We found that BCG vaccination was associated with reduced all-cause mortality, with some weaker evidence of an association with reduced recurrent TB. A plausible mechanism for this association is that BCG vaccination improves treatment outcomes,[@Jeremiah2010] which then results in decreased mortality, and reduced recurrent TB. However, these effects may also be independent and for all-cause mortality may not be directly related to active TB. In this case, a possible mechanism for the association between BCG vaccination and all-cause mortality is that BCG vaccination modulates the innate immune response, resulting in non-specific protection.[@Kleinnijenhuis2012] For low incidence countries, where the reduction in TB cases has been used as evidence to scale back vaccination programs,[@Zwerling2011] these results suggest that BCG vaccination may be more beneficial than previously thought. In countries that target vaccination at those considered to be at high risk of TB the results from this study could be used to help drive uptake by providing additional incentives for vaccination. The evidence we have presented should be considered in future cost-effectiveness studies of BCG vaccination programs.

Further work is required to determine whether years since vaccination and age at vaccination are associated with TB outcomes as this study was limited by low sample size, missing data for year of vaccination, and the relative rarity of some TB outcomes. However, due to the continuous collection of the surveillance data used in this analysis, this study could be repeated once additional data have been collected. The results from this study require validation in independent datasets and the analysis should be reproducible in other low incidence countries that have similarly developed surveillance systems. If validated in low incidence countries, similar studies in medium to high incidence countries should be conducted because any effect would have a greater impact in these settings.

**Acknowledgements**

The authors thank the TB section at Public Health England (PHE) for maintaining the Enhanced Tuberculosis Surveillance (ETS) system; all the healthcare workers involved in data collection for the ETS.

**Funding**

SEA, HC, and EBP are funded by the National Institute for Health Research Health Protection Research Unit (NIHR HPRU) in Evaluation of Interventions at University of Bristol in partnership with Public Health England (PHE). The views expressed are those of the author(s) and not necessarily those of the NHS, the NIHR, the Department of Health or Public Health England.

**Conflicts of interest**

HC reports receiving honoraria from Sanofi Pasteur, and consultancy fees from AstraZeneca, GSK and IMS Health, all paid to her employer.

**Accessibility of data and programming code**

The code for the analysis contained in this paper can be found at: [https://github.com/seabbs/ExploreBCGOnOutcomes](https://github.com/seabbs/ExploreBCGOnOutcomes)

##### PAGEBREAK

**REFERENCES**

<div id = 'refs'></div>

##### PAGEBREAK

## Online supplementary appendix: Exploring the effects of BCG vaccination in patients diagnosed with tuberculosis: observational study using the Enhanced Tuberculosis Surveillance system

Sam Abbott, Hannah Christensen, Maeve K Lalor, Dominik Zennor, Colin Campbell, Mary Ramsay, Ellen Brooks-Pollock

```{r participant summary table, echo=FALSE, message=FALSE, eval = FALSE}
participants_table %>% 
              pretty_sum_results_table(footer = paste0('(% of all cases for BCG vaccination, and % cases with known vaccination status for year of vaccination), * ', tb_death_footer), label = 'participant_table', caption = 'Summary of included notifications for each outcome and exposure variable, for both the univariable and multivariable data sets. Percentages in brackets summarise the percentage of the whole data set (2009-2015) that the selected participants represent. Year of vaccination summarises the data used for both age at vaccination and years of vaccination as both year and age at notification had no missing data, with the denominator being cases that had a recorded vaccination', cap_fun = SuptabRef)
```

**Outcome variables stratified by BCG vaccination status**

```{r outcome summary, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
bake(file  = paste0('../data/', folder, '/outcome_demo_tab_bcg.rds'), {
#Compile summary table
var <- c(outcomes)
var_words <- c(outcomes_words)
categories <- list(c('Yes', 'No', NA))
names(categories[[1]]) <- c('Vaccinated', 'Unvaccinated', 'Unknown vaccine status')
names(categories) <- 'bcgvacc'

  outcome_sum <- bind_cols(demographic_summary(var, var_words, uk_not), variable_summary(categories, var, var_words, uk_not))
}) -> outcome_sum

## Pretty results table
outcome_sum %>% 
    pretty_sum_results_table(col_names = c("Outcome", colnames(outcome_sum)[-1]),
                             footer = paste0('{% all cases}(% complete within vaccine status)[% *complete within category*], * ',  tb_death_footer), 
                             label = 'outcomesum', 
                             caption = 'Outcomes for individuals in England notified with tuberculosis between 2009-2015,  stratified by BCG vaccination status.', 
                             format.args = (list(strip.white = FALSE)),
                             cap_fun = SuptabRef)
```

**Confounding variables stratified by BCG vaccination status**

```{r demographic summary, echo=FALSE, message=FALSE}
bake(file = paste0('../data/', folder, '/demographic_tab_bcg.rds'), {
  #Compile summary table
var <- c(covariates)
var_words <- c(covariates_words)
categories <- list(c('Yes', 'No', NA))
names(categories[[1]]) <- c('Vaccinated', 'Unvaccinated', 'Unknown vaccine status')
names(categories) <- 'bcgvacc'

demo_sum <- bind_cols(demographic_summary(var, var_words, uk_not), variable_summary(categories, var, var_words, uk_not))
}) -> demo_sum_bcg

demo_sum_bcg %>% 
              pretty_sum_results_table(col_names = c("Confounder", colnames(demo_sum_bcg)[-1]),
                                       footer = paste0('{% all cases}(% complete within vaccine status)[% *complete within category*], * ',  tb_death_footer), 
                                       label = 'ConfounderStrat', 
                                       caption = 'Confounders for individuals in England notified with tuberculosis between 2009-2015,  stratified by BCG vaccination status.',
                                       cap_fun = SuptabRef)
```

##### PAGEBREAK

**Model output: All-cause mortality**

```{r bcg-mortality, message = FALSE, echo = FALSE, results = "asis"}
bcg_mortality_tab <- table_var_out[['bcgvacc']] %>% transpose()
bcg_mortality_tab  <- bcg_mortality_tab[["spline"]][[1]]

bcg_mortality_tab  %>%
  pretty_sum_results_table(col_names = c("Variable", "Total", "All-cause mortality", "OR (95% CI)", "P-value",
                                         "aOR (95% CI)", "P-value"),
                           footer = 'OR (95% CI): unadjusted odds ratio with 95% confidence intervals, aOR (95% CI): adjusted odds ratios with 95% confidence intervals', 
                           label = 'bcg_mortality_tab',
                           caption = 'Summary of logistic regression model output with BCG vaccination as the exposure and all-cause mortality as the outcome.',
                                     cap_fun = SuptabRef
)
```

##### PAGEBREAK

**Secondary exposure: Age at vaccination**

We calculated age at vaccination as year of vaccination minus year of birth. We categorized age at vaccination into $0$ to $< 1$, $1$ to $< 12$, $12$ to $< 16$ and 16+ years because the distribution was bi-model with modes at `r as.numeric(names(table(uk_not$AgeatBCG)[rev(order(table(uk_not$AgeatBCG)))]))[1]` and `r as.numeric(names(table(uk_not$AgeatBCG)[rev(order(table(uk_not$AgeatBCG)))]))[2]` years. This categorization captures the current UK policy of vaccination at birth, historic policy of vaccination at 13-15 years and catch up vaccination for high risk children.

```{r Age at vaccination summary table, message=FALSE, echo=FALSE, results='asis'}
table_var_outs[[3]]$Outcome <- stringr::str_replace(as.character(table_var_outs[[3]]$Outcome), "[*]", "\u2021")

table_var_outs[[3]] %>% 
  combine_model_data_sum(DataSum = ModelDataSum[[3]]) %>% 
            pretty_sum_results_table(col_names = c('Outcome', 'Age at BCG', 'Cases*', 'Cases with outcome (%)', 'OR (95% CI)', 'P-value', 'Cases\u2020', 'Cases with outcome (%)', 'aOR (95% CI)', 'P-value'), 
                                     footer = paste0('OR (95% CI): unadjusted odds ratio with 95% confidence intervals, aOR (95% CI): adjusted odds ratios with 95% confidence intervals, ', sum_sample_size('AgeAtVac', VarAdj = 'AgeAtVacAdj', PerDescrip = '% of vaccinated cases'),
                                                     " \u2021 ",  tb_death_footer), 
                                     label = 'ageatvacoutsum', 
                                     caption = 'Summary of associations between age at vaccination and all outcomes in individuals who were vaccinated - the baseline exposure is vaccination at birth compared to vaccination from 1 to < 12, 12 to < 16, and 16+ years of age.',
                                     cap_fun = SuptabRef
                                     )

```

##### PAGEBREAK

**Sensitivity analysis of the missing data using multiple imputation**

We found that repeating the analysis with an imputed data set had some effect on the results from the complete case analysis. There was a decrease in the accuracy of effect size estimates for BCG vaccination, some increase in p-values (`r SuptabRef('sum_imp_bcg')`). However, none of the estimated effects changed their direction, and there were no detectable systematic changes in the results.

For the secondary exposure variables (years since vaccination and age at vaccination, (`r SuptabRef('sum_imp_yrsinceBCG')` and `r SuptabRef('sum_imp_ageatvac')`), we found a change in direction of the point estimate between years since vaccination and all-cause mortality and recurrent TB, but similar results for age at vaccination and outcomes.

```{r Imp BCGvacc table, message=FALSE, echo=FALSE, results='asis', warning=FALSE}
  #BCG vaccination
if (imput) {
  sum_imp_bcg %>% 
              pretty_sum_results_table(footer = paste0('OR: odds ratio with 95% confidence intervals, aOR: adjusted odds ratio with 95% confidence intervals, fmi: fraction of missing information, * ', tb_death_footer), 
                                       col_names =  c('Outcome', 'OR (95% CI)', 'P-value', 'fmi', 'aOR (95% CI)', 'P-value', 'fmi'),
                                       label = 'sum_imp_bcg', 
                                       caption = 'Summary of associations between BCG vaccination and all outcomes, using pooled imputed data.', 
                                       cap_fun = SuptabRef)
}
```

```{r Imp yrsinceBCG table, message=FALSE, echo=FALSE, results='asis', warning=FALSE}
#BCG vaccination
if (imput) {

## Remove spurious results for deaths due to TB (in those who died) (no events in the data set to estimate on)
sum_imp_yrsinceBCG[2, c(2, 5)] <- '*Insufficient data*'

sum_imp_yrsinceBCG %>% 
              pretty_sum_results_table(footer = paste0('OR: odds ratio with 95% confidence intervals, aOR: adjusted odds ratio with 95% confidence intervals, fmi: fraction of missing information, * ', tb_death_footer),
                                       col_names =  c('Outcome', 'OR (95% CI)', 'P-value', 'fmi', 'aOR (95% CI)', 'P-value', 'fmi'),
                                       label = 'sum_imp_yrsinceBCG', 
                                       caption = 'Summary of associations between years since vaccination and all outcomes, using pooled imputed data. There was insufficient data to estimate an effect for deaths due to TB (in those who died)', 
                                       cap_fun = SuptabRef)
}
```

##### PAGEBREAK

```{r Imp ageatvac table, message=FALSE, echo=FALSE, results='asis', warning=FALSE}
  #BCG vaccination
if (imput) {
  sum_imp_ageatvac %>% 
              pretty_sum_results_table(footer = paste0('OR: odds ratio with 95% confidence intervals, aOR: adjusted odds ratio with 95% confidence intervals, fmi: fraction of missing information, * ', tb_death_footer),
                                       col_names =  c('Outcome', 'Age group (years)', 'OR (95% CI)', 'P-value', 'fmi', 'aOR (95% CI)', 'P-value', 'fmi'),
                                       label = 'sum_imp_ageatvac', 
                                       caption =   'Summary of associations between age at vaccination and all outcomes, using pooled imputed data(reference is vaccination at <1 year).', 
                                       cap_fun = SuptabRef)
}
```

##### PAGEBREAK

**Sensitivity analysis of the study population**

```{r BCG vaccination summary table - varying study population, message=FALSE, echo=FALSE, results='asis', warning=FALSE}
if (folder == 'main' && file.exists('../data/drop_recurrent/regression_analysis.rda') && file.exists('../data/elg_school_scheme/regression_analysis.rda')) { 
  #Define function that loads each analysis and formats the table ready to be combined
  read_rename_BCG_ass <- function(path, pop)
  {
    path <- paste0('../data/', path, '/regression_analysis.rda')
    load(path)
    tmp <- table_var_outs[[1]] %>% plyr::mutate(Population = NA) %>% add_row(Population = pop) %>% arrange(Population) %>% dplyr::select(Population, everything()) %>% map(function(.) ifelse(is.na(.), '', .)) %>% bind_cols
    return(tmp)
  }

  #Declare which results to append
  append_results <- c('drop_recurrent', 'elg_school_scheme')
  names(append_results) <- c('Recurrent cases dropped', 'Cases eligible for the \n schools scheme')

  #Append results
  bcg_ass_main <- map2_df(append_results, names(append_results), read_rename_BCG_ass) %>% bind_rows

  ## Make table
  bcg_ass_main %>% 
      pretty_sum_results_table(footer = paste0('OR: odds ratio with 95% confidence intervals, aOR: adjusted odds ratio with 95% confidence intervals, * ', tb_death_footer), 
                              col_names = c('Study population', 'Outcome', 'BCG', 'OR (95% CI)', 'P-value', 'aOR (95% CI)', 'P-value'),
                              label = 'BCGoutsumsens', 
                              caption =   paste0('Summary of associations between BCG vaccination and all outcomes; cases that have no recurrent flag in the ETS (', nrow(uk_not %>% filter(is.na(recurrent))), '), and cases that would have been eligible for the BCG schools scheme (', elg_school_stats_sum['cases eligible for the schools scheme'] %>% as.numeric, '). Those defined to be eligible for the schools scheme are the UK born, that were aged 14 or over in 2004'), 
                              cap_fun = SuptabRef)
}
```

#####PAGEBREAK

**Estimated power**

Power estimates were calculated in both the univariable and multivariable datasets for all outcomes. This represents an over estimate of the statistical power, as only a single exposure variable was accounted for, and age at vaccination has been simplified to a binary vaccinated at birth variable (`r SuptabRef('power_table')`).

```{r Estimate stat power of study, include=FALSE}
bake(file = paste0('../data/', folder, '/estimate_study_power.rds'), {
#source paper calc
source('../functions/Power_calculation.R')

#power_table
power_table = function(var, variables_words, outcomes, covariates, outcomes_words,data)
 {

if (var == 'VaccatBirth') 
   {
        data$VaccatBirth <- ifelse(!is.na(data$AgeatBCG), ifelse(data$AgeatBCG < 1, 'Yes', 'No'), NA) %>% factor
        var_words <- "Age since vaccination (vaccination at birth vs. older)"
   }else{
       var_words <- variables_words[var]
    }

  uni <- outcomes %>% map_df( function(.) power_calc(data, ., var, cov = '', outcomes_words)) %>% select(Outcome, Power)

  #Run for adjusted data
  adj <- outcomes %>% map_df( function(.) power_calc(data, ., var, cov = covariates, outcomes_words)) %>% select(Power)

  #Construct table
  power_table <- paste0(uni$Power,'%') %>% as.matrix %>% t
  power_table <- rbind(power_table, paste0(adj$Power, '%') %>% as.matrix %>% t)
  power_table <- bind_cols(as_data_frame(c('&nbsp;&nbsp;Univariable', '&nbsp;&nbsp;Multivariable')), as_data_frame(power_table))
  colnames(power_table) <- c('Analysis', outcomes_words)

  power_table[3,] <- c(var_words, rep('', length(outcomes_words)))
  power_table <- bind_rows(power_table[3,], power_table[-3,])
  return(power_table)
  }

  # Power for BCG vaccination -----------------------------------------------
  power_table_out <- power_table('bcgvacc', variables_words, outcomes, covariates, outcomes_words, uk_not)

  #Power for Years since vaccination

  power_table_out <- bind_rows(power_table_out, power_table('yrsinceBCG', variables_words, outcomes, covariates, outcomes_words, uk_not))

  # Power for Vaccination at birth -----------------------------------------------

  power_table_out <- bind_rows(power_table_out, power_table('VaccatBirth', variables_words, outcomes, covariates, outcomes_words, uk_not))

}) -> power_table

```

```{r power table, echo=FALSE, message=FALSE, eval = TRUE}
power_table %>% 
 pretty_sum_results_table(label = 'power_table', 
                           col_names = c('Exposure', colnames(.)[-1]),
                           caption = 'Summary of the estimated power for each analysis, for both the univariable and multivariable data sets, with alpha set as 0.05. Power estimates assume a single exposure variable, and age at vaccination has been simplified into a binary vaccinated at birth variable', 
                           cap_fun = SuptabRef
 )
```

```{r LRT test of spline, include=FALSE}
## Table of LRT tests
bake(file = paste0('../data/', folder, '/LRT_test_spline.rds'), {
tmp <- bind_cols(table_age_LRT[['linear spline']][[1]], table_age_LRT[['linear spline']][[3]][,-1], table_age_LRT[['linear spline']][[2]][,-1])

#rename columns
colnames(tmp) <- c('Outcomes', variables_words %>% map(function(.) . <- c(., 'Adjusted'))) %>% unlist
tmp1 <- tmp
}) -> LRT_test_spline
```

```{r LRT test table, include = FALSE, eval = FALSE}
kable(LRT_test_spline)
```

### 2018-03-28 output/cover_letter/cover_letter.Rmd
added analysis to git

---
title: "Cover letter"
output:
    word_document: 
        reference_docx: ../../common/style.docx
    html_notebook: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Dear Editor,

Please find attached our manuscript, entitled "Exploring the effects of BCG vaccination in patients diagnosed with tuberculosis: observational study using the Enhanced Tuberculosis Surveillance system".

We feel our manuscript will be of interest to you and the journal's readership because Bacillus Calmette-Guérin (BCG) is one of the most widely-used vaccines worldwide and our study explores possible additional benefits from vaccination. We found evidence to suggest that vaccination with BCG is associated with reduced all-cause mortality for TB patients and weaker evidence of an association with a reduced chance of having a repeat episode. Recognition of such benefits could lead to a re-evaluation of BCG policy, both in low and high incidence countries. Our study benefits from the use of data from the Enhanced Tuberculosis surveillance system, which contains data on demographics, microbiology and outcomes of active TB cases in England. We made use of logistic regression, with both complete case analysis and multiply imputed data, to explore the effects of BCG vaccination in this robust dataset.

This work was presented as a short oral abstract at the 48th Union world conference on lung health (2017). We do not have any issues relating to the journal policies. All authors have approved this version of the manuscript for submission and the content of the manuscript has not been published or submitted elsewhere.

With kind regards, on behalf of all the co-authors,

Sam Abbott

### 2018-03-28 functions/Power_calculation.R

@@ -1,0 +1,48 @@
+
+power_calc <- function(data, outcome, variable, cov = '', outcomes_words)
+{
+  # Restrict data set -------------------------------------------------------
+  #Compile model variables to select correct data
+  if (cov[1] != c(''))
+  {
+    model <- c(variable, cov)
+  }else{
+    model <- variable
+  }
+  
+  data <- data[,c(outcome, model)] %>% na.omit 
+  
+  #sample size
+  n <- data %>% nrow
+  
+  #Outcome rate given no variable
+  p_1 <- data[,variable] %>% flatten %>% map(function(.)ifelse(. == 1,1,0)) %>% unlist %>% as.logical
+  p_1 <- data[p_1,outcome] %>% flatten %>% map(function(.)ifelse(. == 2,1,0)) %>% unlist
+  p_1 <- sum(p_1)/length(p_1)
+  
+  #Outcome rate given variable
+  p_2 <- data[,variable] %>% flatten %>% map(function(.)ifelse(. == 2,1,0)) %>% unlist %>% as.logical
+  p_2 <- data[p_2,outcome] %>% flatten %>% map(function(.)ifelse(. == 2,1,0)) %>% unlist
+  p_2 <- sum(p_2)/length(p_2)
+  
+  #Outcome rate in sample
+  out_prop <- data[,variable] %>% flatten %>% map(function(.)ifelse(. == 2,1,0)) %>% unlist
+  out_prop <- sum(out_prop)/length(out_prop)
+  
+  #Calulate power
+  power <- powerLogisticBin(n, p_1, p_2, out_prop, alpha = 0.05)
+  
+  #pretty_round analysis
+  p_1 <- pretty_round(p_1, digits = 3)
+  p_2 <- pretty_round(p_2, digits = 3)
+  out_prop <- pretty_round(out_prop, digits = 3)
+  power <- pretty_round(power*100, digits = 0)
+  
+  out <- data_frame(Outcome = outcomes_words[outcome], analysis = ifelse(cov[1] == '', 'Univariate', 'Adjusted'), `Sample size`= n,
+                    `Pr(Y|X=0)` = p_1, `Pr(Y|X=1)` = p_2, `Pr(X=1)` = out_prop, Power = power)
+  return(out)
+}
+
+
+
+

### 2018-03-28 functions/analysis_functions.R

@@ -1,0 +1,210 @@
+# Define funcitons for analysis -------------------------------------------
+
+# Define Models -----------------------------------------------------------
+models_comb <- function(outcome, variable, covariates)
+{
+  univariable = c(list(c(outcome,variable)),covariates %>% map(function(x)
+    c(outcome, x)))
+  names(univariable) <- c(variable, covariates)
+  
+  adjusted = c(list(c(
+    outcome, variable, covariates
+  )), c(variable, covariates) %>% map(
+    function(x)
+      c(outcome, variable, covariates)[!(c(outcome, variable, covariates) %in% x)]
+  ))
+  names(adjusted) <- c('full', variable, covariates)
+  
+  
+  models <- tibble::frame_data(
+    ~outcome, ~variable, ~analysis,    ~models,
+    outcome, variable, 'univariable', univariable,
+    outcome, variable, 'adjusted',   adjusted
+  )
+  return(models)
+}
+
+# Flash models to allow of mulitple analysis levels (i.e linear, quadratic, splines)
+models_flash <- function(models, levels)
+{
+  #store temp models
+  temp_models <- models
+  row_models <- nrow(models)
+  #flash rows - crude but tricky otherwise
+  if (length(levels) > 1)
+  {
+    for (i in levels[-1])
+    {
+      models <- dplyr::bind_rows(models, temp_models)
+    }
+  }
+  
+  #define a column indicating analysis level
+  models <- models %>% mutate(level = unlist(map(levels, function(.) rep(.,row_models))))
+  models <- models[,c(1:3,6, 4:5)]
+}
+
+# Add restricted data -----------------------------------------------------
+models_data <- function(models, data, variable, covariates)
+{
+
+  univariable <- flatten(flatten(models[models$analysis %in% 'univariable', 'models' ])) %>% map(function(x) na.omit(data[, x]))
+  names(univariable) <- c(variable, covariates)
+  
+  adjusted  <-  flatten(flatten(models[models$analysis %in% 'adjusted', 'models' ]))[['full']] %>% list() %>%  map(function(x) na.omit(data[, x])) %>% rep((length(covariates) + 2))
+  names(adjusted) <- c('full', variable, covariates)
+  
+  
+  #bind into data frame
+  models <- models %>% mutate(data = list(
+    univariable,
+    adjusted))
+  return(models)
+}
+
+#Regressions for each outcome and variable ---------------------------
[…]

### 2018-03-28 functions/data_descript_functions.R

@@ -1,0 +1,60 @@
+
+#Distribution for age with BCG vaccination
+age_bcg_no <- quantile(uk_not$age[uk_not$bcgvacc %in% 'No'], c(1/4,1/2,3/4)) %>% as.numeric
+age_bcg_yes <- quantile(uk_not$age[uk_not$bcgvacc %in% 'Yes'], c(1/4,1/2,3/4)) %>% as.numeric
+
+#Test differences between 15-65
+temp_prop <- prop.test(c(length(uk_not$age[(uk_not$bcgvacc %in% 'No' & uk_not$age > 15 & uk_not$age < 65)]),length(uk_not$age[(uk_not$bcgvacc %in% 'Yes' & uk_not$age > 15 & uk_not$age < 65)])), c(length(uk_not$age[uk_not$bcgvacc %in% 'No']),length(uk_not$age[uk_not$bcgvacc %in% 'Yes'])))
+
+age_15_65_prop <-round(temp_prop$estimate*100, digits=1) %>% as.numeric
+age_15_65_test <- ifelse(temp_prop$p.value < 0.001, '< 0.001', as.character(round(temp_prop$p.value, digits=3)))
+
+#Test differences between <15
+temp_prop <- prop.test(c(length(uk_not$age[(uk_not$bcgvacc %in% 'No' & uk_not$age < 15)]),length(uk_not$age[(uk_not$bcgvacc %in% 'Yes' & uk_not$age < 15)])), c(length(uk_not$age[uk_not$bcgvacc %in% 'No']),length(uk_not$age[uk_not$bcgvacc %in% 'Yes'])))
+
+age_15_prop <-round(temp_prop$estimate*100, digits=1) %>% as.numeric
+age_15_test <- ifelse(temp_prop$p.value < 0.001, '< 0.001', as.character(round(temp_prop$p.value, digits=3)))
+
+#Test differences between >65
+temp_prop <- prop.test(c(length(uk_not$age[(uk_not$bcgvacc %in% 'No' & uk_not$age > 65)]),length(uk_not$age[(uk_not$bcgvacc %in% 'Yes' & uk_not$age > 65)])), c(length(uk_not$age[uk_not$bcgvacc %in% 'No']),length(uk_not$age[uk_not$bcgvacc %in% 'Yes'])))
+
+age_65_prop <-round(temp_prop$estimate*100, digits=1) %>% as.numeric
[…]

### 2018-03-28 functions/demographic_functions.R

@@ -1,0 +1,120 @@
+
+
+# Tables of Demographics --------------------------------------------------
+
+#demographic summary
+demographic_summary = function(var, var_words, data, PerDigits = 0)
+{
+    #construct data summary
+    data_summary <- c('All', var) %>% map(function(.){
+      if (. == 'All')
+      {
+        temp <- c('Total, all cases', nrow(data))
+      }else{
+        if (sum(class(unlist(data[,.])) %in% 'factor') == 1)
+        {
+          #Add main variable name
+          temp <- c(c(var_words)[.], sum(!is.na(data[,.])))
+          
+          #percentage calc
+          temp[2] <- paste(temp[2], ' (', pretty_round(as.numeric(temp[2])/nrow(data[,.])*100, digits = PerDigits), ')', collapse = '', sep = '')
+          
+          #Add each catagorical variable
+          for (i in 1:length(levels(unlist(data[,.]))))
+          {
+            temp <- rbind(temp, c(paste('&nbsp;&nbsp; ',levels(unlist(data[,.]))[i]), 
+                                  length(unlist(data[unlist(data[,.]) %in% levels(unlist(data[,.]))[i], .]))))
+            
+            
+            temp[i + 1,2] <- paste('&nbsp;&nbsp; ', temp[i + 1,2], ' [*', pretty_round(as.numeric(temp[i + 1,2])/sum(!is.na(data[,.]))*100, digits = PerDigits), '*]', collapse = '', sep = '')
+          }
+        }else{
+          #Account for non catagorical variables
+          temp <- c(c(var_words)[.], sum(!is.na(data[,.])))
+          
+          temp[2] <- paste(temp[2], ' (', pretty_round(as.numeric(temp[2])/nrow(data[,.])*100, digits = PerDigits), ')', collapse = '', sep  = '')
+          if (. != 'year')
+          {
+            #mean
+            var <- as.numeric(unlist(na.omit(data[,.])))
+            temp <- rbind(temp, c('&nbsp;&nbsp; Mean **[SD]**', paste('&nbsp;&nbsp; ', pretty_round(mean(var), digits = PerDigits), ' **[', pretty_round(sd(var),digits = PerDigits), ']**', collapse = '', sep = '')))
+            #median
[…]

### 2018-03-28 functions/format_functions.R

@@ -1,0 +1,249 @@
+
+# Paper fig and table ref -------------------------------------------------
+figRef <- local({
+  tag <- numeric()
+  created <- logical()
+  used <- logical()
+  function(label, caption, prefix = options("figcap.prefix"), 
+           sep = options("figcap.sep"), prefix.highlight = options("figcap.prefix.highlight")) {
+    i <- which(names(tag) == label)
+    if (length(i) == 0) {
+      i <- length(tag) + 1
+      tag <<- c(tag, i)
+      names(tag)[length(tag)] <<- label
+      used <<- c(used, FALSE)
+      names(used)[length(used)] <<- label
+      created <<- c(created, FALSE)
+      names(created)[length(created)] <<- label
+    }
+    if (!missing(caption)) {
+      created[label] <<- TRUE
+      paste0(prefix.highlight, prefix, " ", i, sep, prefix.highlight, 
+             " ", caption)
+    } else {
+      used[label] <<- TRUE
+      paste(prefix, tag[label])
+    }
+  }
+})
+
+tabRef <- local({
+  tag <- numeric()
+  created <- logical()
+  used <- logical()
+  function(label, caption, prefix = options("tabcap.prefix"), 
+           sep = options("tabcap.sep"), prefix.highlight = options("tabcap.prefix.highlight"), InSentance = FALSE) {
+    i <- which(names(tag) == label)
+    if (length(i) == 0) {
+      i <- length(tag) + 1
+      tag <<- c(tag, i)
+      names(tag)[length(tag)] <<- label
+      used <<- c(used, FALSE)
+      names(used)[length(used)] <<- label
+      created <<- c(created, FALSE)
+      names(created)[length(created)] <<- label
+    }
+    if (!missing(caption)) {
+      created[label] <<- TRUE
+      paste0(prefix.highlight, prefix, " ", i, sep, prefix.highlight, 
+             " ", caption)
+    } else {
+      used[label] <<- TRUE
+      if (!InSentance) {
+        prefix <- tolower(prefix)
+      }
+      paste(prefix, tag[label])
+    }
+  }
+})
+
+# Supp Info fig and tab ref -----------------------------------------------
+
+SupfigRef <- local({
+  tag <- numeric()
+  created <- logical()
+  used <- logical()
[…]

### 2018-03-28 functions/main_functions.R

@@ -1,0 +1,175 @@
+
+# Run regression analysis -------------------------------------------------
+main_analysis <- function(outcomes, variables, covariates, splines, data, BF_cor, BF_scale)
+{
+  #Construct data frame of all models and there required data sets
+  models <-  map2(rep(outcomes,length(variables)), sort(rep(variables,length(outcomes))),
+                  function(outcome, variable) models_comb(outcome, variable, covariates) %>%
+                    models_data(data, variable, covariates)) %>% do.call(rbind,.) %>% models_flash(c('linear', 'spline'))
+  
+  #Run Regressions and calculate anova for each model
+  models <- models %>% models_regress(splines, BF_cor, BF_scale)
+  
+  #Search for regression issues
+  model_warnings <- flatten(models[,'warnings']) %>% 
+                      map_dbl(function(.)if (!is.na(.)) . <- 1 else . <- 0) %>% 
+                         as.logical %>% 
+                            (function(x) {models[x,]})
+  
+  if (nrow(model_warnings) != 0 ) print(model_warnings)
+  
+  return(list(models,model_warnings))
+}
+
+# Summarise Analysis ------------------------------------------------------
+main_summary <- function(models, outcomes_words, variables_words, covariates_words, splines, data)
+{
+
+  #Construct Output summary tables - add demographic data
+  data_summary <- lapply(1:nrow(models), function(i){models[i,]}) %>% map( function(.) data_summary(., variables_words, covariates_words, outcomes_words, data))
+  models <- models %>% mutate(data_summary = data_summary)
+
+  #Summarise ORs for each model
+  models <-  lapply(1:nrow(models), function(i){models[i,]}) %>% map_df(models_summary)
+  
+  #Concatinate final output in a single table for each outcome and variable combination
+  models <- models %>% summary_tables(splines)
+  
+  #Summarise results for each variable against all outcomes
+  models <-  lapply(1:nrow(models), function(i){models[i,]}) %>% map_df(main_effect)
+
+  #Summarise the secondary effects
+  secondary_summary <-  lapply(1:nrow(models), function(i){models[i,]}) %>% map_df(function(.,df=data) secondary_effect(., df))
+  
+  #Summarise sensitivity analysis for each variable against all outcomes
[…]

### 2018-03-28 functions/plot_functions.R

@@ -1,0 +1,150 @@
+
+#Produce forest plot summarising variable effect
+forest_plot<- function(df, outcomes, outcomes_words, variables_words,level){
+  # d is a data frame with 4 columns
+  # d$x gives variable names
+  # d$y gives center point
+  # d$ylo gives lower limits
+  # d$yhi gives upper limits
+  
+  #Reorder outcomes and replace with words
+  df$outcome <- factor(df$outcome, levels = rev(outcomes))
+  df$Outcomes <- df$outcome
+  
+  #Rename variables
+  df$Analysis <- df$analysis
+  
+  plot <- ggplot(df, aes(x=Outcomes, y=OR, ymin=LCI, ymax=UCI, fill=Analysis, colour=Analysis, shape=Analysis))+
+    geom_pointrange(position = position_dodge(width = 0.2)) +
+    geom_hline(yintercept=1, lty=2) +
+    scale_x_discrete(labels = rev(outcomes_words)) +
+    coord_flip() +
+    theme_bw()
+  
+  return(plot)
+}
+
+#Produce plots of the models fitted against the data
+outcome_variable_plot <- function(df, outcome_words, variables_words, k)
+{
+  plot <- ggplot(df, aes(x=age, y=fit, ymin=LCI, ymax=UCI, fill=type, colour=type, shape=type)) +
+    geom_pointrange(position = "jitter") +
+    ylab('proportion') + xlab('age') +
+    ylim(0,1) +
+    theme_bw() +
+    facet_wrap(~level, ncol=2)
+  
+  return(plot)
+}
+
+#Produce forest plot summarising variable effect - including stepwise
+forest_plot_stepwise<- function(df, outcomes, outcomes_words, variables_words,level,stepwise){
+  # d is a data frame with 4 columns
+  # d$x gives variable names
+  # d$y gives center point
+  # d$ylo gives lower limits
+  # d$yhi gives upper limits
+  
+  #correct stepwise formating
+  stepwise <- stepwise %>% mutate(OR=as.numeric(OR), LCI=as.numeric(LCI), UCI=as.numeric(UCI),analysis=rep('stepwise', nrow(stepwise)))
+  #Filter dataframe
+  df <- df[, c('variable', 'outcome', 'OR', 'LCI', 'UCI', 'analysis')]
+  
+  #bind data frame together
[…]

### 2018-03-28 functions/summary_functions.R

@@ -1,0 +1,606 @@
+
+
+# Construct data tables for a single row of models
+data_summary = function(model, variables_words,covariates_words, outcomes_words, data_complete)
+{
+  
+  #Extract key values
+  #Pull out key data information
+  data <-  flatten(flatten(model[,'data']))[[1]]
+  level <-  flatten(flatten(model[,'level']))[[1]]
+  outcome <-  flatten(flatten(model[,'outcome']))[[1]]
+  variable <-  flatten(flatten(model[,'variable']))[[1]]
+  analysis <-  flatten(flatten(model[,'analysis']))[[1]]
+  
+  #Set outcome and positive outcome
+  x <- outcome
+  if (x == 'sputsmear')
+  {
+    Y <- "Positive"
+  }else if (x == 'pulmextrapulm') {
+    Y <- 'Pulmonary, with or without EP'
+    }else{
+    Y <- 'Yes'
+  }
+  
+  if (analysis == 'adjusted')
+  {
+    #construct data summary
+    data_summary <- c('Total cases', colnames(data)[-1]) %>% map(function(.){
+
+      ## Assign matching column name
+      match_col <- paste0(., 'match')
+      
+      ## Add total row
+      if (. == 'Total cases')
+      {
+        temp <- c(., nrow(data), paste0(nrow(data[data[[x]] %in% Y,]), ' (', pretty_round(nrow(data[data[[x]] %in% Y,])/nrow(data)*100, digits = 0), ')'), match_col)
+      }else if (sum(class(unlist(data[,.])) %in% 'factor'))
+      {
+        #Add main variable name
+        temp <- c(c(variables_words,covariates_words)[.], NA, NA, match_col)
+        
+        #Add each catagorical variable
+        for (i in 1:length(levels(unlist(data[,.]))))
+        {
+          temp <- rbind(temp, c(paste('&nbsp;&nbsp; ' ,levels(unlist(data[,.]))[i]), 
+                                length(unlist(data[unlist(data[,.]) %in% levels(unlist(data[,.]))[i], .])),
+                                length(unlist(data[(unlist(data[,x]) %in% Y & unlist(data[,.]) %in% levels(unlist(data[,.]))[i]) , x])),
+                                paste0(., levels(unlist(data[,.]))[i])
+          ))
+          temp[i + 1,3] <- paste(temp[i + 1,3], ' (', pretty_round(as.numeric(temp[i + 1,3])/as.numeric(temp[i + 1,2])*100, digits = 0), ')', collapse = '', sep = '')
+        }
[…]

### 2018-03-28 scripts/data_descript.R

@@ -1,0 +1,181 @@
+stew(file=paste0('../data/', folder, '/descriptive_data.rda'), {
+# Distributions by age ----------------------------------------------------
+
+#Distribution for age with BCG vaccination
+age_bcg_no <- quantile(uk_not$age[uk_not$bcgvacc %in% 'No'], c(1/4,1/2,3/4)) %>% as.numeric
+age_bcg_yes <- quantile(uk_not$age[uk_not$bcgvacc %in% 'Yes'], c(1/4,1/2,3/4)) %>% as.numeric
+
+#Distribution for vaccinated at birth
+age_vcbirth_yes <- quantile(uk_not$age[uk_not$ageatvac %in% '< 1'], c(1/4,1/2,3/4), na.rm=TRUE) %>% as.numeric
+age_vcbirth_no <- quantile(uk_not$age[!(uk_not$ageatvac %in% c('< 1', NA))], c(1/4,1/2,3/4), na.rm=TRUE) %>% as.numeric
+
+#Distribution for age with BCG vaccination
+age_ukborn_no <- quantile(uk_not$age[uk_not$ukborn %in% 'Non-UK Born'], c(1/4,1/2,3/4),na.rm=TRUE) %>% as.numeric
+age_ukborn_yes <- quantile(uk_not$age[uk_not$ukborn %in% 'UK Born'], c(1/4,1/2,3/4), na.rm=TRUE) %>% as.numeric
+
+
+
+# Test differences with age -----------------------------------------------
+
+#Test differences between 15-65
+df <- uk_not
+df$temp <- ifelse (uk_not$age < 65 & uk_not$age > 15, 'Yes', 'No') %>% factor
+df <- df[,c('bcgvacc', 'temp')] %>% na.omit
+prop_tab <- improved_table(df$temp, df$bcgvacc)
+
+age_15_65_prop <- paste0(prop_tab[2, 7:8], '%, ', prop_tab[2, 1:2],'/', prop_tab[3, 1:2])
+age_15_65_test <- prop_tab[2, 9]
+
+#Test differences between <15
+df <- uk_not
+df$temp <- ifelse (uk_not$age < 15, 'Yes', 'No') %>% factor
+df <- df[,c('bcgvacc', 'temp')] %>% na.omit
+prop_tab <- improved_table(df$temp, df$bcgvacc)
+
[…]

### 2018-03-28 scripts/imp_data.R

@@ -1,0 +1,121 @@
+# For nested variables need to add level to make not nested ----------------
+if(sum(imputed_var %in% 'TB_mortality') == 1 )
+{
+  uk_not_nonested$TB_mortality <- uk_not_nonested$TB_mortality %>% as.character
+  uk_not_nonested$TB_mortality <- ifelse(uk_not_nonested$Mortality %in% 'No', 'Still Alive', uk_not_nonested$DeathDueTB) %>%
+    factor
+}
+
+
+## Handle the nesting of yrsinceBCG in BCG vaccination (i.e assign NA to all yrsinceBCG cases where a case has not actually had BCG)
+if(sum(imputed_var %in% 'yrsinceBCG') == 1 )
+{
+  uk_not_nonested$yrsinceBCG <- uk_not_nonested$yrsinceBCG %>% as.character
+  uk_not_nonested$yrsinceBCG <- ifelse(uk_not_nonested$bcgvacc %in% 'No', 'Not vac', uk_not_nonested$yrsinceBCG) %>%
+    factor
+}
+## Handle the nesting of ageatvac in BCG vaccination (i.e assign NA to all ageatvac cases where a case has not actually had BCG)
+if(sum(imputed_var %in% 'ageatvac') == 1 )
+{
+  uk_not_nonested$ageatvac <- uk_not_nonested$ageatvac %>% as.character
+  uk_not_nonested$ageatvac <- ifelse(uk_not_nonested$bcgvacc %in% 'No', 'Not vac', uk_not_nonested$ageatvac) %>% 
+    factor
+}
+
+## add spline relationship to imputation model
+if (sum(imputed_var %in% 'age') == 1 ) {
+  uk_not_nonested <- uk_not_nonested %>% 
+                              mutate(age1  = ns(age, knots = c(summary(age)[c("1st Qu.","Median", "3rd Qu.")]))[,1],
+                                    age2 =  ns(age, knots = c(summary(age)[c("1st Qu.","Median", "3rd Qu.")]))[,2], 
+                                    age3 =  ns(age, knots = c(summary(age)[c("1st Qu.","Median", "3rd Qu.")]))[,3])
+  imputed_var <- c(imputed_var, 'age1', 'age2', 'age3')
+  imputed_var <- imputed_var[imputed_var != 'age']
+}
+
+## Maxit set to 50 iterations
+miss_per_var <- uk_not_nonested %>%  map_dbl(function(.) sum(is.na(.))/length(.)*100) %>% mean
[…]

### 2018-03-28 scripts/packages.R

@@ -1,0 +1,25 @@
+
+# packages used in this module --------------------------------------------
+package_names <- c('tidyverse', 'splines', 'devtools', 'parallel', 
+                   'pryr', 'cowplot', 'reshape2', 'knitr', 'lme4', 'plotly',
+                   'powerMediation', 'compiler', 'pander', 'mice', 'pomp', 
+                   'bookdown', 'matrixStats', 'haven', 'lattice', 'stringr',
+                   'forcats', 'epiR')
+
+pkgLoad <- function(x)
+{
+  if (!require(x,character.only = TRUE))
+  {
+    install.packages(x,dep=TRUE, repos='http://www.stats.bris.ac.uk/R/')
+    if(!require(x,character.only = TRUE)) stop("Package not found")
+  }
+  return(paste0('Loaded ', x, ' successfully'))
+}
+
+pck <- lapply(package_names, pkgLoad)
+print(pck)
+
+#library install via github
+library(modelr)
+library(purrr)
+library(broom)

### 2018-03-28 scripts/regression_analysis.R

@@ -1,0 +1,26 @@
+# Run and Summarise results-----------------------------------------------
+stew(file=paste0('../data/', folder, '/regression_analysis.rda'),{
+#enable compilation
+enableJIT(3)
+
+output <- regression_analysis(outcomes, variables, covariates, splines, outcomes_words, variables_words, covariates_words, splines_words, uk_not, BF_cor = FALSE, BF_scale = NULL)
+
+
+#Summary table of output per variable against a single outcome
+table_var_out <- output[[1]] 
+
+#Summary table of output per variable against all outcomes
+table_var_outs <- output[[2]]
+
+#Summary table of LRT for comparing age terms
+table_age_LRT <- output[[3]] %>% transpose()
+
+#Summary of model sample size
+ModelDataSum <- output[[4]]
+
+#switch off JIT
+enableJIT(0)
+
+#remove output
+rm('output')
+})

### 2018-03-28 sensitivity/NoCasesPrevented.R

@@ -1,0 +1,29 @@
+bake(file = paste0('../data/', folder, '/predicted_cases_prevented.rds'), {
+  
+  unvac_outcome_count <- uk_not %>% 
+    filter(bcgvacc %in% "No") %>% 
+    filter(year >= 2009) %>% 
+    select(Mortality) %>%
+    dplyr::count(Mortality) %>% 
+    dplyr::rename(Mortality_c = n) %>% 
+    dplyr::filter(Mortality %in% "Yes") %>% 
+    select(Mortality = Mortality_c) %>% 
+    gather(key = "outcome", value = "cases")
+  
+  AvgYearlyPredictImpact <- unvac_outcome_count %>% 
+    full_join(aOR.prog, by = "outcome") %>% 
+    gather(key = "measure", value = "effect", OR, LCI, UCI) %>% 
+    mutate(prev_cases = cases * (1 - effect)) %>% 
+    mutate(prev_cases_per_year = prev_cases / length(unique(uk_not$year))) %>% 
+    select(outcome, measure, prev_cases_per_year) %>% 
+    spread(key = "measure", value = "prev_cases_per_year") %>% 
+    group_by(outcome) %>% 
+    mutate(pretty_eff_year = ORFormat(c(OR, UCI, LCI), pref = " (95% CI ", Digits = 0))
+
+  unvac_mort_per_year <- round(unvac_outcome_count$cases[[1]] / length(unique(uk_not$year)), 0)
+  
+  tmp <- list(AvgYearlyPredictImpact, unvac_mort_per_year) 
+}) -> tmp
+
+AvgYearlyPredictImpact <- tmp[[1]]
+unvac_mort_per_year  <- tmp[[2]]

### 2018-03-28 sensitivity/analysis_imp.R

@@ -1,0 +1,156 @@
+## Run analysis for imputed data set, aggregating results into a single table
+  bake(file = paste0('../data/', folder, '/imputed_sum_tab.rds'), {
+    # wrapper function --------------------------------------------------------
+    
+    fit_imputation_wrapper <- function(imp_df)
+    {
+      # run and pool models -----------------------------------------------------
+      
+      # univariable --------------------------------------------------------------
+      model_outcome_uni = function(outcome, variable, LRT = FALSE){
+        
+        if (LRT) {
+          formula <- paste0(outcome, '~', '1')
+        }else{
+          formula <- paste0(outcome, '~', '1 +', variable)
+        }
+        
+        model <- with(imp_df, glm(as.formula(formula), family = binomial("logit")))
+        
+        model <- pool(model)
+        
+        return(model)
+      }
+      
+      model_bcgvacc_uni <- outcomes %>% map(function(.) model_outcome_uni(., variable = 'bcgvacc'))
+      model_ageatvac_uni <- outcomes %>% map(function(.) model_outcome_uni(., variable = 'ageatvac'))
+      model_yrsinceBCG_uni <- outcomes %>% map(function(.) model_outcome_uni(., variable = 'yrsinceBCG'))
+      
+      
+      # Adjusted ----------------------------------------------------------------
+      
+      model_outcome_adj = function(outcome, variable, LRT = FALSE){
+        #covariates with spline - assumes added spline to imputation  
+        cov <- covariates %>% map_chr(function(x){
+          if (splines == x) {
+            x <- paste0('age', seq(1,3)) %>% paste(collapse = ' + ') 
+          }else{
+            x <- x
+          }
+          return(x)
+        })
+        
+        if (LRT) {
+          formula <- paste0(outcome, '~', "1", paste(c('', cov), collapse = ' + '))
+        }else{
+          formula <- paste0(outcome, '~', variable, paste(c('', cov), collapse = ' + '))
+        }
+        
+        
[…]

### 2018-03-28 sensitivity/explore_imputation_yrsinceBCG.R

@@ -1,0 +1,88 @@
+## Load and extract imputed data, bind with original data to get age groups
+imp_df_long_exp  <- complete(imp_df, action = 'long', include = FALSE) %>% 
+  as_tibble
+
+imp_df_long_exp <- imp_df_long_exp %>% 
+  left_join(uk_not %>% 
+              rowid_to_column() %>% 
+              mutate(.id = as.factor(rowid)) %>% 
+              select(.id, agegrp), 
+            by = ".id")
+
+## Exploratory functions
+## summarise yrsinceBCG as percentage by group
+yrsinceBCG_prop_by_group <- function(df, group = c("agegrp", "ukborn")) {
+  dist_yrsinceBCG <- df %>% 
+    filter(!(bcgvacc %in% c("No"))) %>% 
+    dplyr::group_by(.dots = c(group, "yrsinceBCG")) %>% 
+    dplyr::tally() %>% 
+    group_by(.dots = group) %>% 
+    dplyr::mutate(sum_yrsinceBCG = sum(n),
+                  yrsinceBCG_prop = n / sum_yrsinceBCG) %>%
+    select(-n)
+  
+  return(dist_yrsinceBCG)
+}
+
+## Average over imputed data sets and use above function
+yrsinceBCG_prop_by_group_imp <- function(df, group = c("agegrp", "ukborn", ".imp")) {
+  
+  df %>% 
+    filter(!is.na(ageatvac)) %>% 
+    yrsinceBCG_prop_by_group(group = group) %>% 
+    group_by(.dots = c(group, "yrsinceBCG")) %>% 
+    mutate(yrsinceBCG_prop = mean(yrsinceBCG_prop, na.rm = T)) 
+  
+}
+
+## Plot proportions
+plot_yrsincebcg_prop <- function(df, facet_by_ageatvac = FALSE) {
+  p <- df %>% 
+    ggplot(aes(x = ukborn, y  = yrsinceBCG_prop, fill = yrsinceBCG)) +
+    geom_bar(stat = "identity", position = "dodge") +
+    theme_minimal()
+  
+  if (facet_by_ageatvac) {
+    p + facet_grid(agegrp ~ ageatvac)
+  }else{
+    p +    facet_wrap(~agegrp)
+  }
+}
+
+## Summarise bcg vaccination status, by yrsinceBCG
[…]

### 2018-03-28 sensitivity/params_UK_born.R

@@ -1,0 +1,33 @@
+#Remove spurious data
+file.remove(paste0('../data/', folder, '/exposure_outcome_defs.rda'))
+
+# Define Analysis Variables -----------------------------------------------
+stew(file = paste0('../data/', folder, '/exposure_outcome_defs.rda'),{
+  #Define outcomes to analysis - all outcomes selected are binary
+  outcomes <- c('Mortality', 'TB_mortality', 'prevdiag','pulmextrapulm', 'sputsmear')
+  
+  #Define Variables to analysis- BCG status is binary and Years since BCG is continous
+  variables <- c('bcgvacc', 'yrsinceBCG', 'ageatvac')
+  
+  #Define Covariates
+  covariates <- c('age', 'ethgrp', 'sex', 'natquintile', 'year')
+  
+  #Define variables to fit to splines
+  splines <- ('age')
+  
+  #Define outcome in words
+  outcomes_words <- c('All-cause mortality', 'Death due to TB (in those who died*)', 'Recurrent TB', 'Pulmonary TB', 'Sputum smear status - positive')
+  names(outcomes_words) <- outcomes
+  
+  #Define Variables in words
+  variables_words <- c('BCG vaccination', 'Years since BCG vaccine', 'Age at vaccination')
+  names(variables_words) <- variables
+  
+  #Define Covariates in words
+  covariates_words <- c('Age', 'Ethnic group', 'Sex', 'IMD rank (with 1 as most deprived and 5 as least deprived)', 'Calendar year')
+  names(covariates_words) <- covariates
+  
+  #Define splines in words
+  splines_words <- c('age')
+  names(splines_words) <- splines
+})

### 2018-03-29 output/Paper.Rmd
updated draft for grammar

@@
 Its efficacy in adults is context specific, with estimates ranging between 0% and 78%.[@Mangtani2014a] Efficacy has been shown to be dependent on previous exposure, with unexposed individuals receiving the greatest benefit.
-Unlike in adults, BCG has consistently been shown to be highly protective against TB and TB meningitis in children.[@Rodrigues1993; @Colditz1994] For this reason the majority of countries that use BCG vaccinate at birth.[@Zwerling2011] Adult vaccination is no longer common, in the UK where universal BCG vaccination of adolescents was stopped in 2005 in favour of a targeted neonatal programme aimed at high risk children.
+Unlike in adults, BCG has consistently been shown to be highly protective against TB and TB meningitis in children.[@Rodrigues1993; @Colditz1994] For this reason the majority of countries that use BCG vaccinate at birth.[@Zwerling2011] Adult vaccination is no longer common in the UK, where universal BCG vaccination of adolescents was stopped in 2005 in favour of a targeted neonatal programme aimed at high risk children.
 
@@
 The use of routine surveillance data means that this study would be readily repeatable with new data.
-The surveillance data contained multiple known risk factors, this allowed us to adjust for the confounding present in the univariable analysis, which attenuated the evidence for an association with BCG vaccination for all outcomes.
+The surveillance data contained multiple known risk factors, this allowed us to adjust for these confounders in the multivariable analysis, which attenuated the evidence for an association with BCG vaccination for all outcomes.
 However, there are important limitations to consider.
@@
 sum_imp_yrsinceBCG[2, c(2, 5)] <- '*Insufficient data*'
+sum_imp_yrsinceBCG[2, c(3, 6)] <- '-'
 
@@
                                        label = 'sum_imp_ageatvac', 
-                                       caption =   'Summary of associations between age at vaccination and all outcomes, using pooled imputed data(reference is vaccination at <1 year).', 
+                                       caption =   'Summary of associations between age at vaccination and all outcomes, using pooled imputed data (reference is vaccination at <1 year).', 
                                        cap_fun = SuptabRef)

### 2018-04-05 output/Paper.Rmd
added contributing statement

@@
 
+**Contributors**
+
+SA, HC, and EBP conceived and designed the work.
+SA undertook the analysis with advice from all other authors.
+All authors contributed to the interpretation of the data.
+SA wrote the first draft of the paper and all authors contributed to subsequent drafts.
+All authors approve the work for publication and agree to be accountable for the work.
+
 **Funding**

### 2018-06-13 output/Paper.Rmd
added doi link
updated csl style for vaccines

@@
 author: "Sam Abbott, Hannah Christensen, Maeve K Lalor, Dominik Zenner, Colin Campbell, Mary Ramsay, Ellen Brooks-Pollock"
-csl: ../common/bmj.csl
+csl: ../common/vaccines.csl
 output:
@@
 
-The code for the analysis contained in this paper can be found at: [https://github.com/seabbs/ExploreBCGOnOutcomes](https://github.com/seabbs/ExploreBCGOnOutcomes)
+The code for the analysis contained in this paper can be found at: [doi.org/10.5281/zenodo.1213800](doi.org/10.5281/zenodo.1213800)
 

### 2018-06-13 output/Paper.Rmd
added additional references and discussion of meta-analyses

@@
 
-**WORD COUNT:** **Title** 19 (20), **Abstract** 220 (250), **Paper** 2740 (3500)
+**WORD COUNT:** **Title** 19 (20), **Abstract** 220 (300), **Paper** 2740 (5000)
 
@@
 
-**Keywords:** Tuberculosis, BCG, Surveillance
+**Keywords:** Tuberculosis, BCG, Surveillance, Non-specific, Mortality
 
-**What is the key question?**
+**Highlights**
 
-Is there evidence in the Enhanced Tuberculosis (TB) Surveillance System that BCG vaccination improves TB outcomes?
+- Is there evidence in the Enhanced Tuberculosis (TB) Surveillance System that BCG vaccination improves TB outcomes?
 
-**What is the bottom line?**
+- This analysis suggests that vaccination with BCG is associated with reduced all-cause mortality for TB patients and potentially a reduced chance of having a repeat episode.
+There was little evidence of an association with other TB outcomes.
 
-This analysis suggests that vaccination with BCG is associated with reduced all-cause mortality for TB patients and potentially a reduced chance of having a repeat episode.
-
-**Why read on?**
-
-BCG vaccination could reduce all-cause mortality in TB cases and may decrease the number of repeat episodes in low incidence countries.
+- BCG vaccination could reduce all-cause mortality in TB cases and may decrease the number of repeat episodes in low incidence countries.
 
@@
 BCG’s primary mode of action is to directly prevent the development of active, symptomatic disease.
-Its efficacy in adults is context specific, with estimates ranging between 0% and 78%.[@Mangtani2014a] Efficacy has been shown to be dependent on previous exposure, with unexposed individuals receiving the greatest benefit.
-Unlike in adults, BCG has consistently been shown to be highly protective against TB and TB meningitis in children.[@Rodrigues1993; @Colditz1994] For this reason the majority of countries that use BCG vaccinate at birth.[@Zwerling2011] Adult vaccination is no longer common in the UK, where universal BCG vaccination of adolescents was stopped in 2005 in favour of a targeted neonatal programme aimed at high risk children.
+Its efficacy in adults is context specific, with estimates ranging between 0% and 78%.[@Mangtani2014a] Efficacy has been shown to be dependent on previous exposure, with unexposed individuals receiving the greatest benefit.[@Barreto2014a] Unlike in adults, BCG has consistently been shown to be highly protective against TB and TB meningitis in children.[@Rodrigues1993; @Colditz1994] For this reason the majority of countries that use BCG vaccinate at birth.[@Zwerling2011; @WHOBCG2018] Adult vaccination is no longer common in the UK, where universal BCG vaccination of adolescents was stopped in 2005 in favour of a targeted neonatal programme aimed at high risk children.
 
-Vaccination policy has been primarily based on reducing the incidence of active TB and little attention has been given to any additional effects of BCG.[@Fine2005a; @Teo2006] There is some evidence that BCG vaccination induces innate immune responses which may provide non-specific protection,[@Kleinnijenhuis2012] TB patients with BCG scars were found to respond better to treatment with earlier sputum smear conversion,[@Jeremiah2010] and there is evidence to suggest that BCG vaccination is associated with reduced all-cause neonatal mortality[@Higgins] and both reduced TB[@Abubakar2013] and all-cause[@Rieckmann2016] mortality in the general population.
+Vaccination policy has been primarily based on reducing the incidence of active TB and little attention has been given to any additional effects of BCG.[@Fine2005a; @Teo2006] There is some evidence that BCG vaccination induces innate immune responses which may provide non-specific protection,[@Kleinnijenhuis2012] TB patients with BCG scars were found to respond better to treatment with earlier sputum smear conversion,[@Jeremiah2010] and there is evidence to suggest that BCG vaccination is associated with reduced all-cause neonatal mortality[@Garly2003; @Higgins] and both reduced TB[@Abubakar2013] and all-cause[@Rieckmann2016] mortality in the general population.
 Given that the immunology behind TB immunity is not well understood these findings suggest that BCG may play a more important role in improving TB outcomes than previously thought.
@@
 Little work has been done to assess the overall effect of BCG on outcomes for active TB cases although the possible non-specific effects of BCG are an area of active research.[@Higgins; @Kandasamy2016; @Pollard2017] Whilst multiple studies have investigated BCG's association with all-cause mortality, it has been difficult to assess whether the association continues beyond the first year of life.[@Pollard2017] The effect size of the association we identified between BCG and all-cause mortality in active TB cases was comparable to that found in a Danish case-cohort study in the general population (aHR: 0.58 (95% CI 0.39 to 0.85).[@Rieckmann2016] A recent systematic review also found that BCG vaccination was associated with reduced all-cause mortality in neonates, with an average relative risk of 0.70 (95% CI 0.49 to 1.01) from five clinical trials and 0.47 (95% CI 0.32 to 0.69) from nine observational studies at high risk of bias.[@Higgins] We found some weak evidence that BCG vaccination was associated with reduced deaths due to TB (in those who died), although our point estimate had large confidence intervals.
-Several meta-analyses have found evidence supporting this association,[@Colditz1994; @Abubakar2013] with one meta-analysis estimating a 71% (RR: 0.29 95% CI 0.16 to 0.53) reduction in deaths due to TB.[@Colditz1994] Our study was limited by the relative rarity of deaths due to TB in England, whereas other approaches benefited from the use of meta-analysis.
+Several meta-analyses have found evidence supporting this association,[@Colditz1994; @Abubakar2013] with one meta-analysis estimating a 71% (RR: 0.29 95% CI 0.16 to 0.53) reduction in deaths due to TB in individuals vaccinated with BCG.[@Colditz1994] The meta-analysis performed by Abubakar et al.
+also found consistent evidence for this association but was limited by the number, and quality, of studies available.[@Abubakar2013] In contrast to our study, both of these meta-analyses estimated the protection from TB mortality in BCG vaccinated individuals rather than in BCG vaccinated cases.
+Additionally, neither study explored the association between BCG vaccination and all-cause mortality or recurrent TB.
 This study could not determine the possible causal pathway for the association between BCG vaccination all-cause mortality, and recurrent TB.

### 2018-06-13 output/changes-list/vaccines.Rmd
added additional references and discussion of meta-analyses

---
title: "Vaccines"
author: "Sam Abbott"
date: "13/06/2018"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Text

Highlighted the novel findings over the Abubakar et al. 2013 (10.3310/hta17370) review with the following in the discussion;

- The meta-analysis performed by Abubakar et al. also found consistent evidence for this association but was limited by the number, and quality, of studies available.[@Abubakar2013] In contrast to our study, both of these meta-analyses estimated the protection from TB mortality in BCG vaccinated individuals rather than in BCG vaccinated cases. Additionally, neither study explored the association between BCG vaccination and all-cause mortality or recurrent TB.

# Highlights

Modified previous summary bullets to be highlights added the following sentance to the second bullet;

- There was little evidence of an association with other TB outcomes.

# Keywords

Added the following;

- non-specific
- mortality

# References

I have added 4 new references to the introduction;

- BCG vaccine: WHO position paper (https://doi.org/10.1016/j.vaccine.2018.03.009)
- Causes of variation in BCG vaccine efficacy: Examining evidence from the BCG REVAC cluster randomized trial to explore the masking and the blocking hypotheses (https://doi.org/10.1016/j.vaccine.2014.05.042)
- BCG scar and positive tuberculin reaction associated with reduced child mortality in West Africa: A non-specific beneficial effect of BCG? (https://doi.org/10.1016/S0264-410X(03)00181-6)

### 2018-06-13 output/Paper.Rmd
packrat updates, added new version DOI

@@
 
-**WORD COUNT:** **Title** 19 (20), **Abstract** 220 (300), **Paper** 2740 (5000)
+**WORD COUNT:** **Title** 19 (20), **Abstract** 220 (300), **Paper** 2790 (5000)
 
@@
 - This analysis suggests that vaccination with BCG is associated with reduced all-cause mortality for TB patients and potentially a reduced chance of having a repeat episode.
-There was little evidence of an association with other TB outcomes.
+
+- There was little evidence of an association with other TB outcomes.
 
@@
 
-R was used for all statistical analysis.@R The analysis was conducted in two stages.
+R was used for all statistical analysis.[@R] The analysis was conducted in two stages.
 Firstly, we calculated proportions for all demographic and outcome variables, and compared vaccinated and unvaccinated TB cases using the $\chi ^2$ test.
@@
 Several meta-analyses have found evidence supporting this association,[@Colditz1994; @Abubakar2013] with one meta-analysis estimating a 71% (RR: 0.29 95% CI 0.16 to 0.53) reduction in deaths due to TB in individuals vaccinated with BCG.[@Colditz1994] The meta-analysis performed by Abubakar et al.
-also found consistent evidence for this association but was limited by the number, and quality, of studies available.[@Abubakar2013] In contrast to our study, both of these meta-analyses estimated the protection from TB mortality in BCG vaccinated individuals rather than in BCG vaccinated cases.
+also found consistent evidence for this association but was limited by the number, and quality, of studies available.[@Abubakar2013] In contrast to our study, both of these meta-analyses estimated the protection from TB mortality in BCG vaccinated individuals rather than in BCG vaccinated cases who had died from any cause.
 Additionally, neither study explored the association between BCG vaccination and all-cause mortality or recurrent TB.
@@
 
-The code for the analysis contained in this paper can be found at: [doi.org/10.5281/zenodo.1213800](doi.org/10.5281/zenodo.1213800)
+The code for the analysis contained in this paper can be found at: [doi.org/10.5281/zenodo.1213799](doi.org/10.5281/zenodo.1213799)
 

### 2018-06-13 output/changes-list/vaccines.Rmd
packrat updates, added new version DOI

@@
 - The meta-analysis performed by Abubakar et al.
-also found consistent evidence for this association but was limited by the number, and quality, of studies available.[@Abubakar2013] In contrast to our study, both of these meta-analyses estimated the protection from TB mortality in BCG vaccinated individuals rather than in BCG vaccinated cases.
+also found consistent evidence for this association but was limited by the number, and quality, of studies available.[@Abubakar2013] In contrast to our study, both of these meta-analyses estimated the protection from TB mortality in BCG vaccinated individuals rather than in BCG vaccinated cases who had died from any cause.
 Additionally, neither study explored the association between BCG vaccination and all-cause mortality or recurrent TB.
@@
 
-I have added 4 new references to the introduction;
+I have added 3 new references to the introduction;
 

### 2018-06-13 packrat/lib/x86_64-apple-darwin15.6.0/3.5.0/PKPDmodels/doc/UsersGuide.Rnw
packrat updates, added new version DOI

\documentclass[12pt]{article}
\usepackage{Sweave,amsmath,amsfonts,bm}
\usepackage[authoryear,round]{natbib}
\bibliographystyle{plainnat}
\DefineVerbatimEnvironment{Sinput}{Verbatim}
{formatcom={\vspace{-1ex}},fontshape=sl,
  fontfamily=courier,fontseries=b, fontsize=\footnotesize}
\DefineVerbatimEnvironment{Soutput}{Verbatim}
{formatcom={\vspace{-1ex}},fontfamily=courier,fontseries=b,%
  fontsize=\footnotesize}
%%\VignetteIndexEntry{User's Guide}
%%\VignetteDepends{PKPDmodels,lattice}
\title{User's Guide for the PKPDmodels Package}
\author{
  Douglas Bates\\
  Dept. of Statistics, University of Wisconsin - Madison, U.S.A.\\
  and\\
  France Mentr\'{e}\\
  UMR738, INSERM, University Paris Diderot\\
}
\SweaveOpts{engine=R,eps=FALSE,pdf=TRUE,strip.white=true,keep.source=TRUE}
\SweaveOpts{include=TRUE}
\setkeys{Gin}{width=\textwidth}
%% commands
\makeatletter
\newcommand\code{\bgroup\@makeother\_\@makeother\~\@makeother\$\@codex}
\def\@codex#1{{\normalfont\ttfamily\hyphenchar\font=-1 #1}\egroup}
\makeatother
%%\let\code=\texttt
\let\proglang=\textsf
\newcommand{\pkg}[1]{{\fontseries{b}\selectfont #1}}
\newcommand{\Cl}{\ensuremath{\mathit{Cl}}}
\begin{document}
<<preliminaries,echo=FALSE,results=hide>>=
options(width=69,digits=5,show.signif.stars=FALSE,str=strOptions(strict.width="cut"))
library(PKPDmodels)
library(lattice)
#lattice.options(default.theme = function() standard.theme())
lattice.options(default.theme = function() standard.theme(color=FALSE))
@
\maketitle
\begin{abstract}
  The \pkg{PKPDmodels} package provides a function, \code{PKmod},
  that is used to generate model functions for 1, 2, or 3 compartment
  pharmacokinetic models with linear elimination kinetics.  The form
  of administration can be intravenous bolus, infusion or oral
  administration with first-order absorption.  The model can be for
  single dose, multiple doses or steady-state conditions.
\end{abstract}

\section{Introduction}
\label{sec:Intro}

A pharmacokinetic model describes the concentration of an analyte in
the central compartment of a compartment model as a function of time
and other covariates such as dose.  The \code{PKmod} function
generates model functions for pharmacokinetic models with linear
elimination kinetics.  The mode of administration can be intravenous
bolus, continuous infusion or oral with first-order absorption
kinetics.  The dosage can be a single dose, multiple doses or the
steady-state condition after multiple doses.

The pharmacokinetic parameters can be expressed as an elimination rate
constant, $k$ (units of 1/time), the clearance rate, $\Cl$ (units of
volume per unit time) or volume of distribution, $V$ (units of
volume).  Only two of these three parameters are specified because of
the relationship
\begin{displaymath}
  k = \frac{\Cl}{V}
\end{displaymath}

Some types of models include other parameters such as the absorption
rate constant, $k_a$, for oral doses with first-order absorption.

By default the function that is returned is a byte-compiled function
that evaluates both the expected values of the concentration in the
central compartment and the gradient with respect to the parameters.
There is an option to create a function that will evaluate the
expected value of the response and the gradient and the Hessian.
Although a
default set of parameters is provided, these can be modified by a
set of parameter transformations.  For example, the default
parameterization for a one-compartment model uses the
elimination rate constant, $k$, and the clearance, $\Cl$ but the model
can be expressed in terms of the volume of distribution, $V$, and the
clearance, $\Cl$, by including the transformation \code{k ~ V/Cl}.
For population pharmacokinetics it is often helpful to transform from
parameters such as $\Cl$ and $V$ to their logarithms.

\section{A simple example}
\label{sec:simple}

A PK model with linear elimination for a single-dose bolus injection
can be expressed in terms of the elimination rate constant, $k$, and
the volume of distribution, $V$, as
\begin{displaymath}
  C(t) = \frac{D\exp(-k\,t)}{V}
\end{displaymath}
where $C(t)$ is the concentration in the central compartment at time
$t$ (since administration of the dose) and $D$ is the dose.  The
corresponding formula in \proglang{R} is 
<<bolus_1cpt_sd_Vk_expr>>=
PKexpr("bolus", "sd")
@

If we wish to transform to another set of parameters, we specify the
transformations as a list of formulas.  For example, to express the
model in terms of the volume of distribution, $V$, and the clearance,
$\Cl$,
<<bolus1cptSdVClExpr>>=
PKexpr("bolus", "sd", list(k ~ Cl/V))
@ 
or, to express the model in terms of $\log(V)$ and $\log(\Cl)$,
<<bolus1cptSdlVlClExpr>>=
PKexpr("bolus", "sd", list(k ~ Cl/V, Cl ~ exp(lCl), V ~ exp(lV)))
@
Note that a substitution formula has the parameter name on the
left and the expression to be substituted for that parameter on the
right. When there are multiple substitution formulas they are
evaluated left to right.

The function \code{PK1expr} exists solely so that the user can verify
that the formula being used is what they expect.  In typical usage we
create the model function directly.
<<bolus_1cpt_sd_Vk_expr>>=
(bolus1cptSdVk <- PKmod("bolus", "sd"))
@ 
When this function is evaluated at given values of \code{dose},
\code{time} and the pharmacokinetic parameters, it returns the
function evaluations with an attribute called \code{"gradient"}.
<<bolusSdVkEval>>=
tvals <- 0:10
bolus1cptSdVk(dose=1, t=tvals, k=0.3, V=1)
@

When fitting pharmacokinetic models, especially population
pharmacokinetic models, it is helpful to have this analytic gradient
evaluation rather than calculating numerical gradients.  Because
common sub-expressions in the gradient columns and the model function
are evaluated once only and because the resulting function consists of
rather simple byte-compiled expressions the evaluation of the model
function and gradient is essentially as fast as the evaluation of the
model function itself.

The evaluation of common sub-expressions is more obvious when several
parameter transformations are applied
<<bolusSdlVlCl>>=
(bolus1cptSdlVlCl <-
 PKmod("bolus", "sd", list(k ~ Cl/V, Cl ~ exp(lCl), V ~ exp(lV))))
@
where we can see that the clearance, $\Cl=\exp(\mathit{lCl})$
(\code{.expr1}), the volume of distribution, $V=\exp{\mathit(lV)}$
(\code{.expr2}), and the elimination rate constant, $k=\Cl/V$
(\code{.expr3}) are each evaluated once only.

Expressing the parameters in this formulation we obtain the same
function evaluation as before
<<bolusSdlVlClEval>>=
bolus1cptSdlVlCl(dose=1, t=tvals, lCl=log(0.3), lV=0)
@
Naturally the gradient has changed because we are calculating the
derivatives with respect to a different set of parameters.

\section{Single oral dose}
\label{sec:simple}

In phase 1 and phase 2 clinical trials population pharmacokinetics are
often assessed by sampling at many closely-spaced time points in a
small number of subjects after a single oral dose.  The \code{Theoph}
data set in the \pkg{datasets} package comes from such a study of the
drug theophylline which is used in the treatment of asthma.
<<strTheoph, echo=FALSE>>=
str(Theoph <- data.frame(lapply(Theoph, function(x) x)))
@ 
These data are shown in Figure~\ref{fig:Theophplot}.
\begin{figure}[tb]
  \centering
<<Theophplot,fig=TRUE,echo=FALSE,height=3>>=
print(xyplot(conc ~ Time|reorder(Subject, Theoph$conc, max),
             Theoph, type=c("g","b"),
             xlab="Time since drug administration (hr)",
             ylab="Serum concentration (mg/l)",
             aspect="xy", layout=c(6,2)))
@   
\caption{Concentration of theophylline versus time for 12 subjects
  following a single oral dose of the drug.  The panels have been
  ordered (left to right starting at the bottom row) by increasing
  maximum concentration.}
  \label{fig:Theophplot}
\end{figure}

A model function for a single oral dose with first-order absorption
using the parameters $\log(k_a)$, $\log(V)$ and $\log(\Cl)$ is
<<oral1cptSdlkalVlCl>>=
oral1cptSdlkalVlCl <-
 PKmod("oral", "sd", list(ka ~ exp(lka), k ~ exp(lCl)/V, V ~ exp(lV)))
@ 
corresponding to the formula
<<oral1cptSdlkalVlCl>>=
PKexpr("oral", "sd", list(ka ~ exp(lka), k ~ exp(lCl)/V, V ~ exp(lV)))
@

Initial values for the parameters are taken to be $\log(V)_0=-1$,
$\log(k_a)_0=0.5$ and $\log(\Cl)_0=-4$ producing a fitted model
<<fm1>>=
summary(fm1 <- nls(conc ~ oral1cptSdlkalVlCl(Dose, Time, lV, lka, lCl),
                   Theoph, start=c(lV=-1, lka=0.5, lCl=-4),
                   subset=Subject==1), corr=TRUE)
@

\end{document}

### 2018-06-13 packrat/lib/x86_64-apple-darwin15.6.0/3.5.0/Rvmmin/doc/Rvmmin.Rmd
packrat updates, added new version DOI

---
title: "Rvmmin - an R implementation of the Fletcher(1970)
   variable metric method with bounds and masks"
author: "John Nash"
date: '2017-07-18'
output: pdf_document
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{Rvmmin15}
  %\usepackage[utf8]{inputenc}
bibliography: Rvmmin.bib
---

# Rvmmin description, examples and tests

**Rvmmin** is an all-R version of the Fletcher-Nash variable
metric nonlinear parameter optimization code of @Fletcher70
as modified by @cnm79.

This vignette is intended to show various features of the 
package, so it is rather detailed and "busy". However, it 
is also hopefully helpful in showing how to use the method
for more difficult problems.

## Algorithm implementation

Fletcher's variable metric method attempts to mimic Newton's 
iteration for function minimization approximately.

Newton's method starts with an original set of parameters $x_0$.
At a given iteraion, which could be the first, we want to solve

$$x_{k+1}  =  x_{k}  -   H^{-1} g$$

where $H$ is the Hessian and $g$ is the gradient at $x_k$.

Newton's method is unattractive in general function minimization
situations because

 - evaluating the Hessian is generally time consuming and error
 prone;

 - solving the equation 
         $$H delta = -g$$
  (which is much less computational effort than inverting $H$), 
  is still a lot of work which needs to be carried out every
  iteration.

While the base Newton algorithm is as given, generally we
carry out some sort of line search along the search direction
delta from the current iterate $x_k$. Indeed, many otherwise
highly educated workers try to implement it without paying 
attention to safeguarding the iterations and ensuring appropriate
progress towards a minimum.

## Termination nuances

### Termination variation with control tolerances

Let us use the Chebyquad test problem in $n$=4 parameters with different controls **eps** and
**acctol** and tabulate the results to explore how our results change with different values
of these program control inputs.

```{r}
cyq.f <- function (x) {
  rv<-cyq.res(x)
  f<-sum(rv*rv)
}

cyq.res <- function (x) {
# Fletcher's chebyquad function m = n -- residuals 
   n<-length(x)
   res<-rep(0,n) # initialize
   for (i in 1:n) { #loop over resids
     rr<-0.0
     for (k in 1:n) {
  z7<-1.0
  z2<-2.0*x[k]-1.0
        z8<-z2
        j<-1
        while (j<i) {
            z6<-z7
            z7<-z8
            z8<-2*z2*z7-z6 # recurrence to compute Chebyshev polynomial
            j<-j+1
        } # end recurrence loop
        rr<-rr+z8
      } # end loop on k
      rr<-rr/n
      if (2*trunc(i/2) == i) { rr <- rr + 1.0/(i*i - 1) }
      res[i]<-rr
    } # end loop on i
    res
}

cyq.jac<- function (x) {
#  Chebyquad Jacobian matrix
   n<-length(x)
   cj<-matrix(0.0, n, n)
   for (i in 1:n) { # loop over rows
     for (k in 1:n) { # loop over columns (parameters)
       z5<-0.0
       cj[i,k]<-2.0
       z8<-2.0*x[k]-1.0 
       z2<-z8
       z7<-1.0
       j<- 1
       while (j<i) { # recurrence loop
         z4<-z5
         z5<-cj[i,k]
         cj[i,k]<-4.0*z8+2.0*z2*z5-z4
         z6<-z7
         z7<-z8
         z8<-2.0*z2*z7-z6
         j<- j+1
       } # end recurrence loop
       cj[i,k]<-cj[i,k]/n
     } # end loop on k
   } # end loop on i
   cj
}

cyq.g <- function (x) {
   cj<-cyq.jac(x)
   rv<-cyq.res(x)
   gg<- as.vector(2.0* rv %*% cj)
}

require(Rvmmin)
nn <- 4
xx0 <- 1:nn
xx0 <- xx0 / (nn+1.0) # Initial value suggested by Fletcher

# cat("aed\n")
# aed <- Rvmminu(xx0, cyq.f, cyq.g, control=list(trace=2, checkgrad=FALSE))
# print(aed)
#================================
# Now build a table of results for different values of eps and acc
veps <- c(1e-3, 1e-5, 1e-7, 1e-9, 1e-11)
vacc <- c(.1, .01, .001, .0001, .00001, .000001)
resdf <- data.frame(eps=NA, acctol=NA, nf=NA, ng=NA, fval=NA, gnorm=NA)
for (eps in veps) {
  for (acctol in vacc) {
    ans <- Rvmminu(xx0, cyq.f, cyq.g, 
          control=list(eps=eps, acctol=acctol, trace=0))
    gn <- as.numeric(crossprod(cyq.g(ans$par)))
    resdf <- rbind(resdf, 
              c(eps, acctol, ans$counts[1], ans$counts[2], ans$value, gn))
  }
}
resdf <- resdf[-1,]
# Display the function value found for different tolerances
xtabs(formula = fval ~ acctol + eps, data=resdf)
# Display the gradient norm found for different tolerances
xtabs(formula = gnorm ~ acctol + eps, data=resdf)
# Display the number of function evaluations used for different tolerances
xtabs(formula = nf ~ acctol + eps, data=resdf)
# Display the number of gradient evaluations used for different tolerances
xtabs(formula = ng ~ acctol + eps, data=resdf)

```

Here -- and we caution that this is but a single instance of a single test problem -- the 
differences in results and level of effort to obtain them are regulated by the values
of `eps` only. This control is used to judge the size of the gradient norm and the
gradient projection on the search vector.

### Problems of function scale

One of the more difficult aspects of termination decisions is that
we need to decide when we have a "nearly" zero gradient. However, 
this "zero gradient" is relative to the overall scale of the
function. Let us see what happens when we consider solving a problem
where the function scale is adjustable. Note that we multiply the
constant sequence `yy` by `pi/4` to avoid integer values which may
give results that are fortuitously better than may be normally found.

```{r}
require(Rvmmin)
sq<-function(x, exfs=1){
  nn<-length(x)
  yy<-(1:nn)*pi/4
  f<-(10^exfs)*sum((yy-x)^2)
  f
}
sq.g <- function(x, exfs=1){
  nn<-length(x)
  yy<-(1:nn)*pi/4
  gg<- 2*(x - yy)*(10^exfs)
}
require(Rvmmin)
nn <- 4
xx0 <- rep(pi, nn) # crude start

# Now build a table of results for different values of eps and acc
veps <- c(1e-3, 1e-5, 1e-7, 1e-9, 1e-11)
exfsi <- 1:6
resdf <- data.frame(eps=NA, exfs=NA, nf=NA, ng=NA, fval=NA, gnorm=NA)
for (eps in veps) {
  for (exfs in exfsi) {
    ans <- Rvmminu(xx0, sq, sq.g, 
                   control=list(eps=eps, trace=0), exfs=exfs)
    gn <- as.numeric(crossprod(sq.g(ans$par)))
    resdf <- rbind(resdf, 
                   c(eps, exfs, ans$counts[1], ans$counts[2], ans$value, gn))
  }
}
resdf <- resdf[-1,]
# Display the function value found for different tolerances
xtabs(formula = fval ~ exfs + eps, data=resdf)
# Display the gradient norm found for different tolerances
xtabs(formula = gnorm ~ exfs + eps, data=resdf)
# Display the number of function evaluations used for different tolerances
xtabs(formula = nf ~ exfs + eps, data=resdf)
# Display the number of gradient evaluations used for different tolerances
xtabs(formula = ng ~ exfs + eps, data=resdf)
```

The general tendency here is for the amount of work in terms of function 
evaluations to rise with the function scale and with tighter (smaller) 
test tolerances, while the quality of the solution is poorer with 
larger scale and also larger (looser) tolerances. However, some exceptions
can be seen, though the overall quality of solutions (function and gradient
norm) is very good. Moreover, the number of gradient evaluations does not climb
notably with the scale or inverse tolerance.

### Problems of parameter scale

There are similar issues of parameter scaling. Let us look at very simple
sum of squares function where we scale the parameters in a nasty way.

```{r}
ssq.f<-function(x){
   nn<-length(x)
   yy <- 1:nn
   f<-sum((yy-x/10^yy)^2)
   f
}
ssq.g <- function(x){
   nn<-length(x)
   yy<-1:nn
   gg<- 2*(x/10^yy - yy)*(1/10^yy)
}

xy <- c(1, 1/10, 1/100, 1/1000)
# note: gradient was checked using numDeriv
veps <- c(1e-3, 1e-5, 1e-7, 1e-9, 1e-11)
vacc <- c(.1, .01, .001, .0001, .00001, .000001)
resdf <- data.frame(eps=NA, acctol=NA, nf=NA, ng=NA, fval=NA, gnorm=NA)
for (eps in veps) {
  for (acctol in vacc) {
    ans <- Rvmminu(xy, ssq.f, ssq.g, 
          control=list(eps=eps, acctol=acctol, trace=0))
    gn <- as.numeric(crossprod(ssq.g(ans$par)))
    resdf <- rbind(resdf, 
              c(eps, acctol, ans$counts[1], ans$counts[2], ans$value, gn))
  }
}
resdf <- resdf[-1,]
# Display the function value found for different tolerances
xtabs(formula = fval ~ acctol + eps, data=resdf)
# Display the gradient norm found for different tolerances
xtabs(formula = gnorm ~ acctol + eps, data=resdf)
# Display the number of function evaluations used for different tolerances
xtabs(formula = nf ~ acctol + eps, data=resdf)
# Display the number of gradient evaluations used for different tolerances
xtabs(formula = ng ~ acctol + eps, data=resdf)
```

The results above suggest that parameter scaling is not much of a problem. 
Actually, these are the very best results I have found with any method for
this problem, which is actually rather nasty. I suggest trying this problem
on your favourite optimizer. Alternatively, use the package `optimr` and
run the function `opm()` with `method="ALL"`.

### Weeds problem with random starts

This notorious problem (see @cnm79, page 120, @nlpor14, page 205, for details
under the heading **Hobbs Weeds problem**) is small but generally difficult due
to the possibility of bad scaling of both function and parameters and a 
near-singular Hessian in the original parameterization.

The Fletcher variable metric method can solve this problem
quite well, though default termination settings should be overridden.
It is important to ensure there are enough iterations to
allow the method to "grind" at the problem. If one uses default 
settings for maxit in optim:BFGS, then the success rate
drops to less than 2/3 of cases tried below.

Below we use 100 "random" starting points for both Rvmmin and the
optim:BFGS minimizers (which should be, but are not quite, the same).

```{r}
## hobbstarts.R -- starting points for Hobbs problem 
hobbs.f<- function(x){ # # Hobbs weeds problem -- function
    if (abs(12*x[3]) > 500) { # check computability
       fbad<-.Machine$double.xmax
       return(fbad)
    }
    res<-hobbs.res(x)
    f<-sum(res*res)
##    cat("fval =",f,"\n")
##    f
}
hobbs.res<-function(x){ # Hobbs weeds problem -- residual
# This variant uses looping
    if(length(x) != 3) stop("hobbs.res -- parameter vector n!=3")
    y<-c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
         38.558, 50.156, 62.948, 75.995, 91.972)
    t<-1:12
    if(abs(12*x[3])>50) {
       res<-rep(Inf,12)
    } else {
       res<-x[1]/(1+x[2]*exp(-x[3]*t)) - y
    }
}

hobbs.jac<-function(x){ # Jacobian of Hobbs weeds problem
   jj<-matrix(0.0, 12, 3)
   t<-1:12
    yy<-exp(-x[3]*t)
    zz<-1.0/(1+x[2]*yy)
     jj[t,1] <- zz
     jj[t,2] <- -x[1]*zz*zz*yy
     jj[t,3] <- x[1]*zz*zz*yy*x[2]*t
   return(jj)
}

hobbs.g<-function(x){ # gradient of Hobbs weeds problem
    # NOT EFFICIENT TO CALL AGAIN
    jj<-hobbs.jac(x)
    res<-hobbs.res(x)
    gg<-as.vector(2.*t(jj) %*% res)
    return(gg)
}
require(Rvmmin)
set.seed(12345)
nrun<-100
sstart<-matrix(runif(3*nrun, 0, 5), nrow=nrun, ncol=3)
ustart<-sstart %*% diag(c(100, 10, 0.1))
nsuccR <- 0
nsuccO <- 0
vRvm <- rep(NA, nrun)
voptim <- vRvm
fRvm <- vRvm
gRvm <- vRvm
foptim <- vRvm
goptim <- vRvm

for (irun in 1:nrun) {
  us <- ustart[irun,]
#  print(us)
#  ans <- Rvmminu(us, hobbs.f, hobbs.g, control=list(trace=1))
#  ans <- optim(us, hobbs.f, hobbs.g, method="BFGS")
  ans <- Rvmminu(us, hobbs.f, hobbs.g, control=list(trace=0))
  ao <- optim(us, hobbs.f, hobbs.g, method="BFGS", 
               control=list(maxit=3000))
# ensure does not max function out

# cat(irun,"  Rvmminu value =",ans$value,"  optim:BFGS value=",ao$value,"\n")
  if (ans$value < 2.5879) nsuccR <- nsuccR + 1
  if (ao$value < 2.5879) nsuccO <- nsuccO + 1
#  tmp <- readline()
  vRvm[irun] <- ans$value
  voptim[irun] <- ao$value
  fRvm[irun] <- ans$counts[1]
  gRvm[irun] <- ans$counts[2]
  foptim[irun] <- ao$counts[1]
  goptim[irun] <- ao$counts[2]

}
cat("Rvmminu: number of successes=",nsuccR,"  propn=",nsuccR/nrun,"\n")
cat("optim:BFGS no. of successes=",nsuccO,"  propn=",nsuccO/nrun,"\n")
fgc <- data.frame(fRvm, foptim, gRvm, goptim)
summary(fgc)
```

From this summary, it appears that Rvmmin, on average, uses fewer gradient and
function evaluations to achieve the desired result.

For comparison, we now re-run the example with default settings for maxit in 
optim:BFGS.

```{r}
nsuccR <- 0
nsuccO <- 0
for (irun in 1:nrun) {
  us <- ustart[irun,]
#  print(us)
#  ans <- Rvmminu(us, hobbs.f, hobbs.g, control=list(trace=1))
#  ans <- optim(us, hobbs.f, hobbs.g, method="BFGS")
  ans <- Rvmminu(us, hobbs.f, hobbs.g, control=list(trace=0))
  ao <- optim(us, hobbs.f, hobbs.g, method="BFGS")
# ensure does not max function out

# cat(irun,"  Rvmminu value =",ans$value,"  optim:BFGS value=",ao$value,"\n")
  if (ans$value < 2.5879) nsuccR <- nsuccR + 1
  if (ao$value < 2.5879) nsuccO <- nsuccO + 1
#  tmp <- readline()
  vRvm[irun] <- ans$value
  voptim[irun] <- ao$value
  fRvm[irun] <- ans$counts[1]
  gRvm[irun] <- ans$counts[2]
  foptim[irun] <- ao$counts[1]
  goptim[irun] <- ao$counts[2]

}
cat("Rvmminu: number of successes=",nsuccR,"  propn=",nsuccR/nrun,"\n")
cat("optim:BFGS no. of successes=",nsuccO,"  propn=",nsuccO/nrun,"\n")
fgc <- data.frame(fRvm, foptim, gRvm, goptim)
summary(fgc)
```

## Bounds and masks

Let us make sure that Rvmminb is doing the right thing with 
bounds and masks. (This is actually a test in the package.)

### Bounds

```{r}
bt.f<-function(x){
 sum(x*x)
}

bt.g<-function(x){
  gg<-2.0*x
}

lower <- c(0, 1, 2, 3, 4)
upper <- c(2, 3, 4, 5, 6)
bdmsk <- rep(1,5)
xx <- rep(0,5) # out of bounds
ans <- Rvmmin(xx, bt.f, bt.g, lower=lower, upper=upper, bdmsk=bdmsk)
ans

```

### Masks

Here we fix one or more paramters and minimize over the rest.

```{r}
sq.f<-function(x){
   nn<-length(x)
   yy<-1:nn
   f<-sum((yy-x)^2)
   f
}
sq.g <- function(x){
   nn<-length(x)
   yy<-1:nn
   gg<- 2*(x - yy)
}

xx0 <- rep(pi,3)
bdmsk <- c(1, 0, 1) # Middle parameter fixed at pi
cat("Check final function value (pi-2)^2 = ", (pi-2)^2,"\n")

require(Rvmmin)
ans <- Rvmmin(xx0, sq.f, sq.g, lower=-Inf, upper=Inf, bdmsk=bdmsk,
               control=list(trace=2))
ans
ansnog <- Rvmmin(xx0, sq.f, lower=-Inf, upper=Inf, bdmsk=bdmsk,
               control=list(trace=2))
ansnog
```

## References

### 2018-06-13 output/Paper.Rmd
finalised updates for vaccine

@@
 author: "Sam Abbott, Hannah Christensen, Maeve K Lalor, Dominik Zenner, Colin Campbell, Mary Ramsay, Ellen Brooks-Pollock"
-csl: ../common/vaccines.csl
+csl: ../common/vaccine.csl
 output:
@@
 
-- Is there evidence in the Enhanced Tuberculosis (TB) Surveillance System that BCG vaccination improves TB outcomes?
+- Found evidence of an association between BCG vaccination and reduced all-cause mortality in TB cases.
 
-- This analysis suggests that vaccination with BCG is associated with reduced all-cause mortality for TB patients and potentially a reduced chance of having a repeat episode.
+- Weaker evidence of an association between BCG vaccination and reduced repeat TB episodes in TB cases.
 
@@
 
-- BCG vaccination could reduce all-cause mortality in TB cases and may decrease the number of repeat episodes in low incidence countries.
+- We explored the identified associations by age and time since vaccination.
 

### 2018-06-13 output/cover_letter/cover_letter.Rmd
packrat updates, added new version DOI
finalised updates for vaccine

@@
 ---
-title: "Cover letter"
+title: ""
 output:
@@
 Recognition of such benefits could lead to a re-evaluation of BCG policy, both in low and high incidence countries.
+We investigated, but could not find conclusive evidence that, the association with reduced all-cause mortality was due to an association between BCG vaccination and reduced TB mortality.
 Our study benefits from the use of data from the Enhanced Tuberculosis surveillance system, which contains data on demographics, microbiology and outcomes of active TB cases in England.

### 2018-06-28 output/Paper.Rmd
updated discussion wording

@@
 Several meta-analyses have found evidence supporting this association,[@Colditz1994; @Abubakar2013] with one meta-analysis estimating a 71% (RR: 0.29 95% CI 0.16 to 0.53) reduction in deaths due to TB in individuals vaccinated with BCG.[@Colditz1994] The meta-analysis performed by Abubakar et al.
-also found consistent evidence for this association but was limited by the number, and quality, of studies available.[@Abubakar2013] In contrast to our study, both of these meta-analyses estimated the protection from TB mortality in BCG vaccinated individuals rather than in BCG vaccinated cases who had died from any cause.
+also found consistent evidence for this association, with a Rate Ratio of 0.22 (95% CI 0.15 to 0.33).[@Abubakar2013] In contrast to our study, both of these meta-analyses estimated the protection from TB mortality in BCG vaccinated individuals rather than in BCG vaccinated cases who had died from any cause.
 Additionally, neither study explored the association between BCG vaccination and all-cause mortality or recurrent TB.

### 2018-07-13 README.Rmd
readme updates
added zenodo badge
Added preprint doi to readme

@@
 # Exploring the effects of BCG vaccination in patients diagnosed with tuberculosis: observational study using the Enhanced Tuberculosis Surveillance system
+
+[![Paper](https://img.shields.io/badge/Paper-10.1101/366476-lightgrey.svg)](https://doi.org/10.1101/366476)
+[![DOI](https://zenodo.org/badge/127124135.svg)](https://zenodo.org/badge/latestdoi/127124135)
 
@@
 
-- Load the analysis results by running `output/Paper.Rmd`. The sensitivity analysis results can be reproduced by setting variables in this document, as explained at the top of the script.
+- Load the analysis results by running `output/Paper.Rmd`.
+This complete analysis is not reproducible as the raw data cannot be published, however all interim results are stored in the repository and these can be explored.
+The sensitivity analysis results can be reproduced by setting variables in this document, as explained at the top of the script.
 

### 2019-02-20 README.Rmd
dropped packrat and moved to checkpoint based package loading

@@
 
-- Intall R and Rstudio.
+- Intall R (analysis run with `3.5.0`) and Rstudio.
 
@@
 
-- Install the required packages using `packrat` as follows.
+- Install the required packages using the following script. Versioned to "2018-06-19".
 
 ```{r get-packages, eval = FALSE}
-install.packages("packrat")
-packrat::restore()
-
+source("scripts/packages.R")
 ```

### 2019-02-20 scripts/packages.R

@@ -1,4 +1,3 @@
-
 # packages used in this module --------------------------------------------
 package_names <- c('tidyverse', 'splines', 'devtools', 'parallel', 
                    'pryr', 'cowplot', 'reshape2', 'knitr', 'lme4', 'plotly',
@@ -10,16 +9,17 @@ pkgLoad <- function(x)
 {
   if (!require(x,character.only = TRUE))
   {
-    install.packages(x,dep=TRUE, repos='http://www.stats.bris.ac.uk/R/')
+    install.packages(x, dep=TRUE)
     if(!require(x,character.only = TRUE)) stop("Package not found")
   }
   return(paste0('Loaded ', x, ' successfully'))
 }
 
+## Set up checkpoint versioning
+pkgLoad("checkpoint")
+checkpoint("2018-06-19")
+
+## Install/Load all other packages
 pck <- lapply(package_names, pkgLoad)
 print(pck)
 
-#library install via github
-library(modelr)
-library(purrr)
-library(broom)

### 2019-02-20 output/review/vaccine-review-response.Rmd
added reviewer responses and update package loading

---
title: "Reviewer Response"
output: word_document
---

Dear Dr. Orenstein,

Thank you for sending review reports for our manuscript. The reviewer's comments were very useful, and we have implemented the majority of them.

We enclose a revised version of the manuscript (clean and with tracked changes), which addreses the reviewer's comments and suggestions. The reviewer's comments are also shown below with our response to each comment.

Yours sincerely,

Sam Abbott

## Authors' response to Reviewer #1:

Thank you for your detailed comments on our manuscript. We have addressed your concerns below.

### Summary

1. "The data set used has a very large amount of missing data that has high potential to introduce bias into the results. About 1/3 of people studied had unknown BCG vaccination status and those with unknown vaccination status had much higher mortality than people in whom vaccination status was known (11% vs 3%). Of those that died and had known vaccination status, a further 38% of people did not have a known cause of death. Because this study only analyzed cases with complete data, conclusions are based on a subset of TB cases that may not be representative of the population as a whole."

We agree that missing data was an issue in this study, particularly missing BCG vaccination status. To address this we conducted all analyses in this study using multiple imputation (imputing all outcomes and exposures). We also discussed results from the complete analysis in the context of the results from the analysis with multiply imputed data. However, we were unable to fully account for the impact of a missing not at random mechanism for BCG vaccination. This was therefore listed as a limitation in the discussion:

"Variable data completeness changed with time, with both BCG vaccination status and year of vaccination having a high percentage of missing data, which may not be missing completely at random. We therefore checked the robustness of our results with multiple imputation including regional variability, however an unknown missing not at random mechanism, or unmeasured confounding may still have introduced bias. We found a greatly increased risk of all-cause mortality for those vaccinated more than 10 years ago in the analysis with multiply imputed data, compared to the complete case analysis. This is likely to be driven by a missing not at random mechanism for years since vaccination, with older cases being both more likely to have been vaccinated more than 10 years previously and to also have an unknown year of vaccination."

### Highlights

1. "It would be more helpful if this section listed exactly what was found in the study and how it contributes to the literature rather than make general statements about what was explored."

Based on the requirments laid out on the Vaccine help page (https://www.elsevier.com/authors/journal-authors/highlights), we feel that this section lists the results as presented in this study.

### Introduction

1. "I suggest the introduction section discuss in more precise terms exactly what is known about the efficacy of BCG vaccination and what questions remain unanswered that this study is trying to address."

We acknowledge that the efficacy of BCG vaccination is important to include as context for our study. We updated paragraph two of the introduction to explicitly state that the BCG vaccine has been shown to be highly effective in England, and that its effectivness may be correlated with distance from the equator. This now reads as follows.

  "BCG’s primary mode of action is to directly prevent the development of active, symptomatic disease. Its efficacy in adults is context specific, with estimates ranging between 0% and 78%.[3] It has been shown to highly efficacious in England and there is some evidence that efficacy increases with distance from the equator.[7] Efficacy has been shown to be dependent on previous exposure, with unexposed individuals receiving the greatest benefit.[4] Unlike in adults, BCG has consistently been shown to be highly protective against TB and TB meningitis in children.[5,6] For this reason the majority of countries that use BCG vaccinate at birth.[7,8]"

We feel that the final paragraph of the introduction details which questions remain unanswered - this is included below in response to your next comment.

2. "There is frequent mention of in the introduction (and throughout this manuscript) of the "additional effects" of BCG aside from reducing the incidence of active TB. However, this term is very imprecise, and it makes it difficult to interpret exactly which additional effects the authors are referring to. There are numerous studies showing BCG has many such additional effects, particularly reducing the development of TB meningitis and disseminated TB in young children. The authors need to more clearly describe which "additional effects" of BCG vaccination are not known and how this study can help us understand them better."

We agree that this needed clarifiction. In point 4 below, we have added an explanation of additional effects in the context of TB outcomes. We feel that the following sentences (page 6, lines 19-23 and page 7, lines 1-3) complete this definition:

  "There is some evidence that BCG vaccination induces innate immune responses which may provide non-specific protection,[11] TB patients with BCG scars were found to respond better to treatment with earlier sputum smear conversion,[12] and there is evidence to suggest that BCG vaccination is associated with reduced all-cause neonatal mortality[13,14] and both reduced TB[15] and all-cause[16] mortality in the general population. Given that the immunology behind TB immunity is not well understood these findings suggest that BCG may play a more important role in improving TB outcomes than previously thought."

This was the only use of this phrase in the manuscript.

3."The statement on line 7-9 that the emergence of drug-resistant TB has made BCG the best available vaccine needs further justification. I am unaware of any vaccine candidates that have differential vaccine efficacy based on MDR status."

We agree that this sentance was unclear. We have reworded it to the following:

  "However, the lack of a more effective vaccine and the emergence of drug-resistant TB strains means that BCG vaccination remains an important tool for reducing TB incidence and mortality rates."

4. "I would suggest revising the statement on lines 18-19 that "Vaccination policy has been primarily based on reducing the incidence of active TB and 19 little attention has been given to any additional effects of BCG" be revised. Vaccination policy is based just as much on the attenuation of disease severity (less meningitis and disseminated TB in kids) as it is reducing the overall incidence of TB."

We agree that this was an omission. We have amended this sentence to the following:

  "Vaccination policy has been primarily based on reducing the incidence of active TB, and mitigating disease severity, with little attention having been given to any additional effects of BCG vaccination on TB outcomes."

5. "I would recommend the introduction section end with a clear statement of the study question."

The primary aim of the study is detailed in the penultimate sentence of the introduction:

  "We aimed to quantify the effects of BCG vaccination on outcomes for individuals with notified TB in England using routinely collected surveillance data to provide evidence for appropriate public health action and provision."

The final sentence details the secondary aim of the study:

  "Where we found an association, we additionally explored the role of years since vaccination, and age at vaccination."

### Methods

1. "A much more detailed discussion is needed to understand how the outcome variables differ from one another and how the information was collected. Specifically, a clearer description of how "all-cause mortality" and "TB-related mortality" were collected is needed."

Asked Maeve for details. Alternatively review PHE report.

2. "How long were most people followed to assess for all-cause mortality? The table indicates people were "followed for 12, 24, and 36 months (or until treatment completion)". Given patients with drug susceptible TB are usually only treated for 6 months, it seems few people would have been followed for this amount of time."

Asked Maeve for details. Alternatively review PHE report.

3. "It is also unclear what criteria were used to determine whether the death was due to TB and who made the assessment."

Asked Maeve for details. Alternatively review PHE report.

4. "There is no information included as to how information about BCG vaccination was collected. Was this entirely by self-report? If so, was there any effort made to ensure the information provided was accurate (e.g. cross-referencing vaccination records), particularly the age at vaccination?"

Asked Maeve for details. Alternatively review PHE report.

5. "For the multiple imputation procedure, which variables were imputed? Was BCG vaccination status imputed for this model when missing?"

We updated the methods section to highlight that all outcome and explanatory variables were included in the multple imputation approach:

  "To mitigate the impact of missing data we used multiple imputation, with the MICE package.[27] We imputed 50 data sets (for 20 iterations) using all outcome and explanatory variables included in the analysis as predictors along with Public Health England centre. The model results were pooled using the small sample method,[28] and effect sizes compared with those from the main analysis."

6. "I would consider discussing the outcome variables in the text rather than listing them in table format. As above, the information in the tables does not provide much clarity about the outcome variables."

We have moved the outcome variable definitions from a table into the main text of the manuscript. Including the modifications from comments 1., 2., and 3. above this now reads as follows:

### Results

1. "This manuscript contains a large number of tables and many of them seem to be redundant. I would suggest combining some of the tables so more information can be included in the main manuscript rather than the supplement. For example, Tables S1 and S2 contain information that would normally be a part of "Table 1". I would consider combining these tables to make a new Table 1 so readers can more clearly see how outcomes and covariates differed based on BCG vaccine status."

We have moved both S1 and S2 into the main body of the text. The tables are presented seperately to make it clear that not all outcomes were considered in the same model.

2. "I suggest including the total n used for each analysis in all tables."

We have included both the univariate and multivariate sample size for all analyses that use complete case analysis. For analyses that use mutltiple imputation, all samples were used. We have now highlighted this by adding the following to each table using multiply imputed data (stars indicating data that varies by table): "Sample size: ***".

3. "In supplemental table S9, the post-hoc power calculation is not necessary. This is simply re-stating the primary analysis. By definition those analyses that did not reach statistical significance were not adequately powered to detect statistical significance given the observed effect size."

We agree that this is not required and have removed it from the supplementary information.

4. "Since there were an unusually high number of deaths determined to not be due to TB in this population, it would be helpful to include the other most common causes of death. "

Unfortunately causes of death that were not due to TB, or related to TB were not recorded in our dataset, so we are unable to add this information.

### Discussion

1. "Much of the discussion section is a re-statement of the results. I would suggest the authors focus more on interpreting the results from the study and describing how they relate to the published literature."

We agree that a large part of the discussion is devoted to discussing the results of this study. This mostly relates to the limitations of the study, and the approaches that we took to mitigate them, that we felt needed to be discussed in order to put our results into context. That being said, we feel that paragraphs four and five of the discussion interpret the results from this study and relate to the published literature.

2. "This study had a large amount of missing data for the exposure and outcome variables of interest and this could be a source of significant bias. This is a major limitation of the study and warrants a thorough discussion (usually the 2nd to last paragraph)."

We agree that the proportion of missing data is a significant limitation of this study (as with many studies using routine surveillance data). As outlined above, to explore the impact of this we repeated all analyses using multiple imputation. This is discussed, along with the remaining limitations in the following sections:

  "Variable data completeness changed with time, with both BCG vaccination status and year of vaccination having a high percentage of missing data, which may not be missing completely at random. We therefore checked the robustness of our results with multiple imputation including regional variability, however an unknown missing not at random mechanism, or unmeasured confounding may still have introduced bias. We found a greatly increased risk of all-cause mortality for those vaccinated more than 10 years ago in the analysis with multiply imputed data, compared to the complete case analysis. This is likely to be driven by a missing not at random mechanism for years since vaccination, with older cases being both more likely to have been vaccinated more than 10 years previously and to also have an unknown year of vaccination. The high percentage of missing data also means that we were likely to be underpowered to detect an effect of BCG vaccination on sputum smear status and deaths due to TB (in those who died), with years since vaccination, and age at vaccination likely to be underpowered for all outcomes"

## Authors' response to Reviewer #2:

Thank you for your review of our manuscript. We have addressed your comments below.

1.	"Given the recent revaccination results, it would be good to see a breakdown of the effect of primary vaccination and revaccination on outcome. It is not clear from S4, for example, whether the non-neonatal persons were revaccinated. "

We agree that it would have been interesting to explore the impact of revaccination on our findings. Unfortunately, we were not able to investigate this as we did not have data on revaccination.

2.	"I am a bit baffled by examining the duration of effect data according to a discrete variable: more or less than 10 year since vaccination… why was age since vaccination not used as a continuous variable in analysis?"

We categorised duration of effect for two reasons. Firstly, to minimise recall bias. Our justification for this is that recall bias is likely to have an increasing impact with time since vaccination. This would mean that those vaccinated several decades ago would be more likely to report an incorrect vaccination date with greater error than those vaccinated more recently. Secondly, any relationship between outcomes and years since BCG vaccination is unlikely to be linear over the timescales considered in this study. We have now highlighted this in the manuscript with the following statement in the methods secton:

"Years since BCG vaccination was defined as year of notification minus year of vaccination and categorised into two groups (0 to 10 and 11+ years). This was based on: evidence that the average duration of BCG protection is 10-15 years;[15] concerns that recall bias maybe greater when vaccination took place over a decade ago compared to within the last few years; and any association between years since vaccination and TB outcomes may be non-linear."

3.	"The outcome measures used for this study appear randomly chosen - there could be many additional outcomes of TB disease. Such as extent of pulmonary disease on chest roentgenograms.

We agree that there could be additional outcome variables included in this study. Unfortunately, we could only include outcomes that were included in the routine surveliance dataset on which this study was based. Given the availability of outcome data, we selected outcomes for further evaluation based on  evidence from the literature of prior associations with BCG vaccination, associations with TB infectiousness, or poor consequences for patients. We have now highlighted this in the manuscript with the following statement in the methods section:

"We considered five TB outcomes that were selected based on: their availability in the ETS; evidence from the literature of prior associations with BCG vaccination; associations with increased case infectiousness; or severe outcomes for patients "

4. "Similarly, was a specific analysis done to look at extrapulmonary, disseminated or tuberculous disease?"

This would be an interesting follow up study as it is plausible that some of the effects we have identified would differ by disease type. However, we did not pre-specify this in our analysis plan and therefore did not explore this in the current study. Additionally, given the low sample size for some outcomes, analyses on disease subset are likely to be underpowered.

### 2019-02-20 scripts/packages.R

@@ -3,7 +3,7 @@ package_names <- c('tidyverse', 'splines', 'devtools', 'parallel',
                    'pryr', 'cowplot', 'reshape2', 'knitr', 'lme4', 'plotly',
                    'powerMediation', 'compiler', 'pander', 'mice', 'pomp', 
                    'bookdown', 'matrixStats', 'haven', 'lattice', 'stringr',
-                   'forcats', 'epiR')
+                   'forcats', 'epiR', 'rmarkdown')
 
 pkgLoad <- function(x)
 {
@@ -17,7 +17,7 @@ pkgLoad <- function(x)
 
 ## Set up checkpoint versioning
 pkgLoad("checkpoint")
-checkpoint("2018-06-19")
+checkpoint("2018-06-19", scanForPackages = TRUE)
 
 ## Install/Load all other packages
 pck <- lapply(package_names, pkgLoad)

### 2019-02-21 scripts/packages.R
added checkpoint changes

@@ -17,7 +17,8 @@ pkgLoad <- function(x)
 
 ## Set up checkpoint versioning
 pkgLoad("checkpoint")
-checkpoint("2018-06-19", scanForPackages = TRUE)
+if (!dir.exists("./.checkpoint")) { dir.create("./.checkpoint")}
+checkpoint("2018-06-19", scanForPackages = TRUE, checkpointLocation = "./.checkpoint")
 
 ## Install/Load all other packages
 pck <- lapply(package_names, pkgLoad)

### 2019-02-21 output/review/vaccine-review-response.Rmd
updated reviewer responses

@@
 title: "Reviewer Response"
-output: word_document
+output:
+  word_document: default
+  pdf_document: default
+  html_document:
+    df_print: paged
 ---
@@
 We enclose a revised version of the manuscript (clean and with tracked changes), which addreses the reviewer's comments and suggestions.
-The reviewer's comments are also shown below with our response to each comment.
+The reviewer’s comments are shown below with our response to each comment in italics.
 
@@
 
-Thank you for your detailed comments on our manuscript. We have addressed your concerns below.
-
 ### Summary
 
-1.
-"The data set used has a very large amount of missing data that has high potential to introduce bias into the results.
-About 1/3 of people studied had unknown BCG vaccination status and those with unknown vaccination status had much higher mortality than people in whom vaccination status was known (11% vs 3%).
+The manuscript "Exploring the effects of BCG vaccination in patients diagnosed with tuberculosis: observational study using the Enhanced Tuberculosis Surveillance system" by Abbott et al.
+is a retrospective analysis that examines the association between prior BCG vaccination and all-cause mortality and TB-related mortality in reported TB cases in England.
+Although BCG has been shown to reduce the incidence of TB disease and disease severity in children, there is less data as to its effect on mortality and data on the length of protection are mixed.
+In this study, prior BCG vaccination was associated with lower all-cause and TB-related mortality (though the latter did not reach statistical significance) in patients diagnosed with TB.
+While these findings are interesting, I think there are significant methodologic concerns.
+The data set used has a very large amount of missing data that has high potential to introduce bias into the results.
+About 1/3
+of people studied had unknown BCG vaccination status and those with unknown vaccination status had much higher mortality than people in whom vaccination status was known (11% vs 3%).
 Of those that died and had known vaccination status, a further 38% of people did not have a known cause of death.
-Because this study only analyzed cases with complete data, conclusions are based on a subset of TB cases that may not be representative of the population as a whole."
+Because this study only analyzed cases with complete data, conclusions are based on a subset of TB cases that may not be representative of the population as a whole.
 
-We agree that missing data was an issue in this study, particularly missing BCG vaccination status.
+*Thank you for your detailed comments on our manuscript. We have addressed your concerns below.*
+
+*We agree that missing data was an issue in this study, particularly missing BCG vaccination status.
 To address this we conducted all analyses in this study using multiple imputation (imputing all outcomes and exposures).
@@
 However, we were unable to fully account for the impact of a missing not at random mechanism for BCG vaccination.
-This was therefore listed as a limitation in the discussion:
+This was therefore listed as a limitation in the discussion:*
 
-"Variable data completeness changed with time, with both BCG vaccination status and year of vaccination having a high percentage of missing data, which may not be missing completely at random.
+*"Variable data completeness changed with time, with both BCG vaccination status and year of vaccination having a high percentage of missing data, which may not be missing completely at random.
 We therefore checked the robustness of our results with multiple imputation including regional variability, however an unknown missing not at random mechanism, or unmeasured confounding may still have introduced bias.
 We found a greatly increased risk of all-cause mortality for those vaccinated more than 10 years ago in the analysis with multiply imputed data, compared to the complete case analysis.
-This is likely to be driven by a missing not at random mechanism for years since vaccination, with older cases being both more likely to have been vaccinated more than 10 years previously and to also have an unknown year of vaccination."
+This is likely to be driven by a missing not at random mechanism for years since vaccination, with older cases being both more likely to have been vaccinated more than 10 years previously and to also have an unknown year of vaccination."*
 
@@
 
-1. "It would be more helpful if this section listed exactly what was found in the study and how it contributes to the literature rather than make general statements about what was explored."
+It would be more helpful if this section listed exactly what was found in the study and how it contributes to the literature rather than make general statements about what was explored.
 
-Based on the requirments laid out on the Vaccine help page (https://www.elsevier.com/authors/journal-authors/highlights), we feel that this section lists the results as presented in this study.
+*We agree that this section should focus on the results from this study.
+Based on the requirments laid out on the Vaccine help page (https://www.elsevier.com/authors/journal-authors/highlights), we have adapted the final bullet point as follows:*
+
+*There was weak evidence of associations between TB outcomes and age at, or years since, vaccination.*
 
@@
 
-1.
-"I suggest the introduction section discuss in more precise terms exactly what is known about the efficacy of BCG vaccination and what questions remain unanswered that this study is trying to address."
+I suggest the introduction section discuss in more precise terms exactly what is known about the efficacy of BCG vaccination and what questions remain unanswered that this study is trying to address.
 
-We acknowledge that the efficacy of BCG vaccination is important to include as context for our study.
-We updated paragraph two of the introduction to explicitly state that the BCG vaccine has been shown to be highly effective in England, and that its effectivness may be correlated with distance from the equator.
-This now reads as follows.
+*We acknowledge that information known about the efficacy of BCG vaccination is important to include as context for our study.
+We did not give precise estimates for the efficacy of the BCG vaccine as efficacy has been shown to be context specific.
+However, We updated paragraph two of the introduction to explicitly state that the BCG vaccine has been shown to be highly effective in England, and that its effectivness may be correlated with distance from the equator.
+This now reads as follows.*
 
-  "BCG’s primary mode of action is to directly prevent the development of active, symptomatic disease.
-Its efficacy in adults is context specific, with estimates ranging between 0% and 78%.[3] It has been shown to highly efficacious in England and there is some evidence that efficacy increases with distance from the equator.[7] Efficacy has been shown to be dependent on previous exposure, with unexposed individuals receiving the greatest benefit.[4] Unlike in adults, BCG has consistently been shown to be highly protective against TB and TB meningitis in children.[5,6] For this reason the majority of countries that use BCG vaccinate at birth.[7,8]"
+  *"BCG’s primary mode of action is to directly prevent the development of active, symptomatic disease.
+Its efficacy in adults is context specific, with estimates ranging between 0% and 78%.[3] It has been shown to highly efficacious in England and there is some evidence that efficacy increases with distance from the equator.[7] Efficacy has been shown to be dependent on previous exposure, with unexposed individuals receiving the greatest benefit.[4] Unlike in adults, BCG has consistently been shown to be highly protective against TB and TB meningitis in children.[5,6] For this reason the majority of countries that use BCG vaccinate at birth.[7,8]"*
 
-We feel that the final paragraph of the introduction details which questions remain unanswered - this is included below in response to your next comment.
+*We feel that the final paragraph of the introduction details which questions remain unanswered - this is included below in response to your next comment.*
 
-2.
-"There is frequent mention of in the introduction (and throughout this manuscript) of the "additional effects" of BCG aside from reducing the incidence of active TB.
+There is frequent mention of in the introduction (and throughout this manuscript) of the "additional effects" of BCG aside from reducing the incidence of active TB.
 However, this term is very imprecise, and it makes it difficult to interpret exactly which additional effects the authors are referring to.
 There are numerous studies showing BCG has many such additional effects, particularly reducing the development of TB meningitis and disseminated TB in young children.
-The authors need to more clearly describe which "additional effects" of BCG vaccination are not known and how this study can help us understand them better."
+The authors need to more clearly describe which "additional effects" of BCG vaccination are not known and how this study can help us understand them better.
 
-We agree that this needed clarifiction.
-In point 4 below, we have added an explanation of additional effects in the context of TB outcomes.
-We feel that the following sentences (page 6, lines 19-23 and page 7, lines 1-3) complete this definition:
+*We agree that this needed clarifiction.
+As part of our response to your comment below lines 18-19 we have added an explanation of additional effects in the context of TB outcomes.
+We feel that the following sentences (page 6, lines 19-23 and page 7, lines 1-3) complete this definition:*
 
-  "There is some evidence that BCG vaccination induces innate immune responses which may provide non-specific protection,[11] TB patients with BCG scars were found to respond better to treatment with earlier sputum smear conversion,[12] and there is evidence to suggest that BCG vaccination is associated with reduced all-cause neonatal mortality[13,14] and both reduced TB[15] and all-cause[16] mortality in the general population.
-Given that the immunology behind TB immunity is not well understood these findings suggest that BCG may play a more important role in improving TB outcomes than previously thought."
+  *"There is some evidence that BCG vaccination induces innate immune responses which may provide non-specific protection,[11] TB patients with BCG scars were found to respond better to treatment with earlier sputum smear conversion,[12] and there is evidence to suggest that BCG vaccination is associated with reduced all-cause neonatal mortality[13,14] and both reduced TB[15] and all-cause[16] mortality in the general population.
+Given that the immunology behind TB immunity is not well understood these findings suggest that BCG may play a more important role in improving TB outcomes than previously thought."*
 
-This was the only use of this phrase in the manuscript.
+*This was the only use of this phrase in the manuscript.*
 
-3."The statement on line 7-9 that the emergence of drug-resistant TB has made BCG the best available vaccine needs further justification.
-I am unaware of any vaccine candidates that have differential vaccine efficacy based on MDR status."
+The statement on line 7-9 that the emergence of drug-resistant TB has made BCG the best available vaccine needs further justification.
+I am unaware of any vaccine candidates that have differential vaccine efficacy based on MDR status.
 
-We agree that this sentance was unclear. We have reworded it to the following:
+*We agree that this sentance was unclear. We have reworded it to the following:*
 
-  "However, the lack of a more effective vaccine and the emergence of drug-resistant TB strains means that BCG vaccination remains an important tool for reducing TB incidence and mortality rates."
+*"However, the lack of a more effective vaccine and the emergence of drug-resistant TB strains means that BCG vaccination remains an important tool for reducing TB incidence and mortality rates."*
 
-4.
-"I would suggest revising the statement on lines 18-19 that "Vaccination policy has been primarily based on reducing the incidence of active TB and 19 little attention has been given to any additional effects of BCG" be revised.
-Vaccination policy is based just as much on the attenuation of disease severity (less meningitis and disseminated TB in kids) as it is reducing the overall incidence of TB."
+I would suggest revising the statement on lines 18-19 that "Vaccination policy has been primarily based on reducing the incidence of active TB and 19 little attention has been given to any additional effects of BCG" be revised.
+Vaccination policy is based just as much on the attenuation of disease severity (less meningitis and disseminated TB in kids) as it is reducing the overall incidence of TB.
 
-We agree that this was an omission. We have amended this sentence to the following:
+*We agree that this was an omission. We have amended this sentence to the following:*
 
-  "Vaccination policy has been primarily based on reducing the incidence of active TB, and mitigating disease severity, with little attention having been given to any additional effects of BCG vaccination on TB outcomes."
+*"Vaccination policy has been primarily based on reducing the incidence of TB disease, and mitigating disease severity, with little attention having been given to any additional effects of BCG vaccination on TB outcomes."*
 
-5. "I would recommend the introduction section end with a clear statement of the study question."
+I would recommend the introduction section end with a clear statement of the study question.
 
-The primary aim of the study is detailed in the penultimate sentence of the introduction:
+*The primary aim of the study is detailed in the penultimate sentence of the introduction:*
 
-  "We aimed to quantify the effects of BCG vaccination on outcomes for individuals with notified TB in England using routinely collected surveillance data to provide evidence for appropriate public health action and provision."
+*"We aimed to quantify the effects of BCG vaccination on outcomes for individuals with notified TB in England using routinely collected surveillance data to provide evidence for appropriate public health action and provision."*
 
-The final sentence details the secondary aim of the study:
+*The final sentence details the secondary aim of the study:*
 
-  "Where we found an association, we additionally explored the role of years since vaccination, and age at vaccination."
+*"Where we found an association, we additionally explored the role of years since vaccination, and age at vaccination."*
 
@@
 
-1.
-"A much more detailed discussion is needed to understand how the outcome variables differ from one another and how the information was collected.
-Specifically, a clearer description of how "all-cause mortality" and "TB-related mortality" were collected is needed."
+A much more detailed discussion is needed to understand how the outcome variables differ from one another and how the information was collected.
+Specifically, a clearer description of how "all-cause mortality" and "TB-related mortality" were collected is needed.
+How long were most people followed to assess for all-cause mortality?
+The table indicates people were "followed for 12, 24, and 36 months (or until treatment completion)".
+Given patients with drug susceptible TB are usually only treated for 6 months, it seems few people would have been followed for this amount of time.
+It is also unclear what criteria were used to determine whether the death was due to TB and who made the assessment.
 
-Asked Maeve for details. Alternatively review PHE report.
+*We agree that futher detail was required to help differentiate outcomes.
+We also agree that the expplanation of the approach used to define mortality needed reworking.
+We made these changes below at the same time as moving from a table to describing the outcomes in the text.*
 
-2.
-"How long were most people followed to assess for all-cause mortality?
-The table indicates people were "followed for 12, 24, and 36 months (or until treatment completion)".
-Given patients with drug susceptible TB are usually only treated for 6 months, it seems few people would have been followed for this amount of time."
-
-Asked Maeve for details. Alternatively review PHE report.
-
-3. "It is also unclear what criteria were used to determine whether the death was due to TB and who made the assessment."
-
-Asked Maeve for details. Alternatively review PHE report.
-
-4.
-"There is no information included as to how information about BCG vaccination was collected.
+There is no information included as to how information about BCG vaccination was collected.
 Was this entirely by self-report?
 If so, was there any effort made to ensure the information provided was accurate (e.g.
-cross-referencing vaccination records), particularly the age at vaccination?"
+cross-referencing vaccination records), particularly the age at vaccination?
 
-Asked Maeve for details. Alternatively review PHE report.
+*We agree that this should be included in the methods section. We have amended the explanation of BCG vaccination status to the following:*
 
-5. "For the multiple imputation procedure, which variables were imputed? Was BCG vaccination status imputed for this model when missing?"
+*"BCG status was collected during the assessment of TB cases and then entered in the ETS.
+When available BCG vaccination records are used, although often these are not available and so patient recall or the presence of a scare is relied on.
+When cases are uncertain, and there is no evidence of a scar, no BCG status is given.
+Year of vaccination is collected similarly."*
 
-We updated the methods section to highlight that all outcome and explanatory variables were included in the multple imputation approach:
+*We acknowledge that this is a limitation of our study for which ideally a validation study would be carried out.
+To highlight this we included the following in the discussion (amended to include year of vaccination).*
 
-  "To mitigate the impact of missing data we used multiple imputation, with the MICE package.[27] We imputed 50 data sets (for 20 iterations) using all outcome and explanatory variables included in the analysis as predictors along with Public Health England centre.
-The model results were pooled using the small sample method,[28] and effect sizes compared with those from the main analysis."
+*"Finally, BCG vaccination status, and year of vaccination, may be subject to misclassification due to recall bias; validation studies of the recording of BCG status in the ETS would be required to assess this."*
 
-6.
-"I would consider discussing the outcome variables in the text rather than listing them in table format.
-As above, the information in the tables does not provide much clarity about the outcome variables."
+For the multiple imputation procedure, which variables were imputed? Was BCG vaccination status imputed for this model when missing?
 
-We have moved the outcome variable definitions from a table into the main text of the manuscript. Including the modifications from comments 1., 2., and 3. above this now reads as follows:
+*We updated the methods section to highlight that all outcome and explanatory variables were included in the multple imputation approach:*
+
+*"To mitigate the impact of missing data we used multiple imputation, with the MICE package.[27] We imputed 50 data sets (for 20 iterations) using all outcome and explanatory variables included in the analysis as predictors along with Public Health England centre.
+The model results were pooled using the small sample method,[28] and effect sizes compared with those from the main analysis."*
+
+I would consider discussing the outcome variables in the text rather than listing them in table format.
+As above, the information in the tables does not provide much clarity about the outcome variables.
+
+*We have moved the outcome variable definitions from a table into the main text of the manuscript. Including the modifications from comments 1., 2., and 3. above this now reads as follows:*
+
+*"The outcomes we considered were: all-cause mortality, death due to TB (in those who died), recurrent TB, pulmonary disease, and sputum smear status.
+These outcomes were selected based on: their availability in the ETS; evidence from the literature of prior associations with BCG vaccination; associations with increased case infectiousness; or severe outcomes for patients."*"
+
+*"All-cause mortality was defined using the overall outcome recorded in ETS, this is based on follow to 36 months starting from date of starting treatment.
+Where the treatment start date was not available the notification date was used if appropriate.
+The date of death was validated against Office for National Statistics (ONS) data.
+Those that were lost to follow up, or not evaluated were treated as missing.
+In cases with a known cause of death, death due to TB was defined as those that died from TB, or where TB had contributed to their death.
+This was assessed using the cases death certificates.
+TB cases who had recurrent episodes were identified in the dataset using their NHS number.
+Positive sputum smear status was given to cases that had a sputum sample shown to contain Acid-Fast Bacilli.
+A positive sputum smear status indicates that cases are more likely to be infectious.
+Cases were defined as having pulmonary TB if a positive sputum smear sample was recorded or if a positive culture was grown from a pulmonary laboratory specimen.
+People with laryngeal TB were defined as having pulmonary TB, and people with miliary TB were not excluded."*
 
@@
 
-1.
-"This manuscript contains a large number of tables and many of them seem to be redundant.
+This manuscript contains a large number of tables and many of them seem to be redundant.
 I would suggest combining some of the tables so more information can be included in the main manuscript rather than the supplement.
 For example, Tables S1 and S2 contain information that would normally be a part of "Table 1".
-I would consider combining these tables to make a new Table 1 so readers can more clearly see how outcomes and covariates differed based on BCG vaccine status."
+I would consider combining these tables to make a new Table 1 so readers can more clearly see how outcomes and covariates differed based on BCG vaccine status.
 
-We have moved both S1 and S2 into the main body of the text. The tables are presented seperately to make it clear that not all outcomes were considered in the same model.
+*We have moved both S1 and S2 into the main body of the text.
+The tables are presented seperately to make it clear that not all outcomes were considered in the same model.
+Whilst this is our prefered method of presenting this data we would be happy to combine these tables into a single table if the reviewer felt this would improve interpretability.*
 
-2. "I suggest including the total n used for each analysis in all tables."
+I suggest including the total n used for each analysis in all tables.
 
-We have included both the univariate and multivariate sample size for all analyses that use complete case analysis.
-For analyses that use mutltiple imputation, all samples were used.
-We have now highlighted this by adding the following to each table using multiply imputed data (stars indicating data that varies by table): "Sample size: ***".
+*We have included both the univariate and multivariate sample size for all analyses that use complete case analysis.
+These can be found at the bottom of each table as a footnote.
+For analyses that use mutltiple imputation, all samples with a given outcome were used.
+For BCG vaccination this is all samples.
+For age at vaccination and years at vaccination the sample size varied between multiply imputed datasets as the proportion of the population with imputed BCG vaccination changed - this means we cannot give a definite sample size for these analyses.
+For the sensitivity analysis the sample size was presented in the table caption.
+This has been highlighted by using n = prior to the sample size.*
 
-3.
-"In supplemental table S9, the post-hoc power calculation is not necessary.
+In supplemental table S9, the post-hoc power calculation is not necessary.
 This is simply re-stating the primary analysis.
-By definition those analyses that did not reach statistical significance were not adequately powered to detect statistical significance given the observed effect size."
+By definition those analyses that did not reach statistical significance were not adequately powered to detect statistical significance given the observed effect size.
 
-We agree that this is not required and have removed it from the supplementary information.
+*We agree that this is not required and have removed it from the supplementary information.*
 
-4. "Since there were an unusually high number of deaths determined to not be due to TB in this population, it would be helpful to include the other most common causes of death. "
+Since there were an unusually high number of deaths determined to not be due to TB in this population, it would be helpful to include the other most common causes of death.
 
-Unfortunately causes of death that were not due to TB, or related to TB were not recorded in our dataset, so we are unable to add this information.
+*Unfortunately causes of death that were not due to TB, or related to TB were not recorded in our dataset, so we are unable to add this information.*
 
@@
 
-1.
-"Much of the discussion section is a re-statement of the results.
-I would suggest the authors focus more on interpreting the results from the study and describing how they relate to the published literature."
+Much of the discussion section is a re-statement of the results.
+I would suggest the authors focus more on interpreting the results from the study and describing how they relate to the published literature.
 
-We agree that a large part of the discussion is devoted to discussing the results of this study.
-This mostly relates to the limitations of the study, and the approaches that we took to mitigate them, that we felt needed to be discussed in order to put our results into context.
-That being said, we feel that paragraphs four and five of the discussion interpret the results from this study and relate to the published literature.
+*We agree that the first paragraph of the discussion summarises the results of this study.
+We feel that this is required as this is the only summary of our findings in the paper, with results presented by outcome in the results section.
+A large part of the relates to the limitations of the study, and the approaches that we took to mitigate them, that we felt needed to be discussed in order to put our results into context.
+We feel that paragraphs four and five of the discussion interpret the results from this study and relate to the published literature.
+However, we would be very happy to add any suggested literature references to the paper, as additional context can only improve the interpretation of our findings.
+*
 
-2.
-"This study had a large amount of missing data for the exposure and outcome variables of interest and this could be a source of significant bias.
-This is a major limitation of the study and warrants a thorough discussion (usually the 2nd to last paragraph)."
+This study had a large amount of missing data for the exposure and outcome variables of interest and this could be a source of significant bias.
+This is a major limitation of the study and warrants a thorough discussion (usually the 2nd to last paragraph).
 
-We agree that the proportion of missing data is a significant limitation of this study (as with many studies using routine surveillance data).
+*We agree that the proportion of missing data is a significant limitation of this study (as with many studies using routine surveillance data).
 As outlined above, to explore the impact of this we repeated all analyses using multiple imputation.
-This is discussed, along with the remaining limitations in the following sections:
+This is discussed, along with the remaining limitations in the following sections:*
 
-  "Variable data completeness changed with time, with both BCG vaccination status and year of vaccination having a high percentage of missing data, which may not be missing completely at random.
+*"Variable data completeness changed with time, with both BCG vaccination status and year of vaccination having a high percentage of missing data, which may not be missing completely at random.
 We therefore checked the robustness of our results with multiple imputation including regional variability, however an unknown missing not at random mechanism, or unmeasured confounding may still have introduced bias.
@@
 This is likely to be driven by a missing not at random mechanism for years since vaccination, with older cases being both more likely to have been vaccinated more than 10 years previously and to also have an unknown year of vaccination.
-The high percentage of missing data also means that we were likely to be underpowered to detect an effect of BCG vaccination on sputum smear status and deaths due to TB (in those who died), with years since vaccination, and age at vaccination likely to be underpowered for all outcomes"
+The high percentage of missing data also means that we were likely to be underpowered to detect an effect of BCG vaccination on sputum smear status and deaths due to TB (in those who died), with years since vaccination, and age at vaccination likely to be underpowered for all outcomes"*
 
@@
 
-Thank you for your review of our manuscript. We have addressed your comments below.
+There is renewed interest in BCG vaccination.
+For example, recently published data suggests that BCG revaccination in adolescence may prevent Mtb infection.
+More data are emerging about effects of different strains, doses and routes of vaccination: all variables that may ultimately significantly impact TB outcomes.
+This manuscript is therefore very timely and contributes to the knowledge through a thorough retrospective look at specific outcomes of TB disease, and of mortality.
+There are shortcomings, but the authors have spelled these out.
+The suggestion that BCG may reduce all-come mortality in older age groups, and that recurrent TB may be prevented by BCG, is intriguing and contributes to our knowledge of this vaccine.
 
-1.	"Given the recent revaccination results, it would be good to see a breakdown of the effect of primary vaccination and revaccination on outcome.
+*Thank you for your review of our manuscript. We have addressed your comments below.*
+
+1.	Given the recent revaccination results, it would be good to see a breakdown of the effect of primary vaccination and revaccination on outcome.
 It is not clear from S4, for example, whether the non-neonatal persons were revaccinated.
-"
 
-We agree that it would have been interesting to explore the impact of revaccination on our findings. Unfortunately, we were not able to investigate this as we did not have data on revaccination.
+*We agree that it would have been interesting to explore the impact of revaccination on our findings.
+Unfortunately, we were not able to investigate this as we did not have data on revaccination.
+Given that the rate of revaccination in England is thought to be low we feel that any analyses looking at this would likely be underpowered.
+*
 
-2.	"I am a bit baffled by examining the duration of effect data according to a discrete variable: more or less than 10 year since vaccination… why was age since vaccination not used as a continuous variable in analysis?"
+2.	I am a bit baffled by examining the duration of effect data according to a discrete variable: more or less than 10 year since vaccination… why was age since vaccination not used as a continuous variable in analysis?
 
-We categorised duration of effect for two reasons.
-Firstly, to minimise recall bias.
-Our justification for this is that recall bias is likely to have an increasing impact with time since vaccination.
-This would mean that those vaccinated several decades ago would be more likely to report an incorrect vaccination date with greater error than those vaccinated more recently.
-Secondly, any relationship between outcomes and years since BCG vaccination is unlikely to be linear over the timescales considered in this study.
-We have now highlighted this in the manuscript with the following statement in the methods secton:
+*We initially considered using a continuous varibale but ultimately decided to use a binary categorical variable for years since vaccination.
+This was primarily motivated by concerns over the linearity of any relationship between TB outcomes and years since vaccination.
+Whilst BCG effectiveness appears to decrease in the first 10-15 years after vaccination with a near linear relationship it is unknown whether this continues beyond 15 years.
+The use of a linear relationship in this scenario may have led to a reduction in the effect size.
+10 years was chosen as a breakpoint as this is a conservative estimate of when BCG effectiveness has reduced to near parity with no vaccination.
+We hypothesised that any protection from negative outcomes would also decrease over the same timescale.
+We were also concerned the recall bias of year of vaccination would increase with years since vaccination, categorising reduces the impact of this bias.
+We have now highlighted this in the manuscript with the following statement in the methods secton: *
 
-"Years since BCG vaccination was defined as year of notification minus year of vaccination and categorised into two groups (0 to 10 and 11+ years).
-This was based on: evidence that the average duration of BCG protection is 10-15 years;[15] concerns that recall bias maybe greater when vaccination took place over a decade ago compared to within the last few years; and any association between years since vaccination and TB outcomes may be non-linear."
+*"Years since BCG vaccination was defined as year of notification minus year of vaccination and categorised into two groups (0 to 10 and 11+ years).
+This was based on: evidence that the average duration of BCG protection is 10-15 years;[15] increasing recall bias with time since vaccination, and any association between years since vaccination and TB outcomes may be non-linear."*
 
-3.	"The outcome measures used for this study appear randomly chosen - there could be many additional outcomes of TB disease. Such as extent of pulmonary disease on chest roentgenograms.
+*To highlight that understanding the functional form for the decay of the effectiveness of the BCG vaccine at preventing negative outcomes we have amended future work section of the discussion as follows:*
 
-We agree that there could be additional outcome variables included in this study.
-Unfortunately, we could only include outcomes that were included in the routine surveliance dataset on which this study was based.
+*"Further work is required to determine whether years since vaccination and age at vaccination are associated with TB outcomes as this study was limited by low sample size, missing data for year of vaccination, and the relative rarity of some TB outcomes.
+However, due to the continuous collection of the surveillance data used in this analysis, this study could be repeated once additional data have been collected.
+If this study were to repeated with a larger sample size particular attention should be given to the functional form of any decay in protection from negative TB outcomes.
+The results from this study require validation in independent datasets and the analysis should be reproducible in other low incidence countries that have similarly developed surveillance systems.
+If validated in low incidence countries, similar studies in medium to high incidence countries should be conducted because any effect would have a greater impact in these settings."*
+
+3. The outcome measures used for this study appear randomly chosen - there could be many additional outcomes of TB disease. Such as extent of pulmonary disease on chest roentgenograms.
+
+*We agree that there could be additional outcome variables included in this study.
+However, we could only include outcomes that were included in the routine surveliance dataset on which this study was based.
 Given the availability of outcome data, we selected outcomes for further evaluation based on  evidence from the literature of prior associations with BCG vaccination, associations with TB infectiousness, or poor consequences for patients.
-We have now highlighted this in the manuscript with the following statement in the methods section:
+We have now highlighted this in the manuscript with the following statement in the methods section: *
 
-"We considered five TB outcomes that were selected based on: their availability in the ETS; evidence from the literature of prior associations with BCG vaccination; associations with increased case infectiousness; or severe outcomes for patients "
+*"The outcomes we considered were: all-cause mortality, death due to TB (in those who died), recurrent TB, pulmonary disease, and sputum smear status.
+These outcomes were selected based on: their availability in the ETS; evidence from the literature of prior associations with BCG vaccination; associations with increased case infectiousness; or severe outcomes for patients"*
 
-4. "Similarly, was a specific analysis done to look at extrapulmonary, disseminated or tuberculous disease?"
+4. Similarly, was a specific analysis done to look at extrapulmonary, disseminated or tuberculous disease?
 
-This would be an interesting follow up study as it is plausible that some of the effects we have identified would differ by disease type.
+*This would be an interesting follow up study as it is plausible that some of the effects we have identified would differ by disease type.
 However, we did not pre-specify this in our analysis plan and therefore did not explore this in the current study.
 Additionally, given the low sample size for some outcomes, analyses on disease subset are likely to be underpowered.
+We have added the following to the final paragraph of our discussion highlighting that this is a logical next step.*
+
+*"Additionally, a larger sample size would allow investigation of the associations identified between TB outcomes and BCG vaccination stratified by pulmonary, extrapulmonary, and disseminated TB disease"*

### 2019-03-04 output/review/vaccine-review-response.Rmd
Added updated draft and review responses

@@
 output:
+  pdf_document: default
   word_document: default
-  pdf_document: default
   html_document:
@@
 
-Thank you for sending review reports for our manuscript. The reviewer's comments were very useful, and we have implemented the majority of them.
+Thank you for sending review reports for our manuscript. The reviewer’s comments were useful, and have improved the paper.
 
-We enclose a revised version of the manuscript (clean and with tracked changes), which addreses the reviewer's comments and suggestions.
+We enclose a revised version of the manuscript (clean and with tracked changes), which addreses the reviewer’s comments and suggestions.
 The reviewer’s comments are shown below with our response to each comment in italics.
@@
 *We agree that missing data was an issue in this study, particularly missing BCG vaccination status.
-To address this we conducted all analyses in this study using multiple imputation (imputing all outcomes and exposures).
+To address this, we conducted all analyses in this study using multiple imputation (imputing all outcomes and exposures).
 We also discussed results from the complete analysis in the context of the results from the analysis with multiply imputed data.
-However, we were unable to fully account for the impact of a missing not at random mechanism for BCG vaccination.
+However, as with all studies based on routinely-collected data,  we were unable to fully account for the impact of a missing not at random mechanism for BCG vaccination.
 This was therefore listed as a limitation in the discussion:*
@@
 We did not give precise estimates for the efficacy of the BCG vaccine as efficacy has been shown to be context specific.
-However, We updated paragraph two of the introduction to explicitly state that the BCG vaccine has been shown to be highly effective in England, and that its effectivness may be correlated with distance from the equator.
+However, we updated paragraph two of the introduction to explicitly state that the BCG vaccine has been shown to be highly effective in England, and that its effectivness may be correlated with distance from the equator.
 This now reads as follows.*
@@
 
-*We feel that the final paragraph of the introduction details which questions remain unanswered - this is included below in response to your next comment.*
+*The final paragraph of the introduction details which questions remain unanswered - this is included below in response to your next comment.*
 
@@
 
-*We agree that this was an omission. We have amended this sentence to the following:*
+*This was an omission. We have amended this sentence to the following:*
 
@@
 
-*We agree that futher detail was required to help differentiate outcomes.
-We also agree that the expplanation of the approach used to define mortality needed reworking.
+*We have added futher detail to help differentiate outcomes and reworked the explanation of the approach used to define mortality.
 We made these changes below at the same time as moving from a table to describing the outcomes in the text.*
@@
 *"BCG status was collected during the assessment of TB cases and then entered in the ETS.
-When available BCG vaccination records are used, although often these are not available and so patient recall or the presence of a scare is relied on.
+When available BCG vaccination records are used, although often these are not available and so patient recall, or the presence of a scare is relied on.
 When cases are uncertain, and there is no evidence of a scar, no BCG status is given.
-Year of vaccination is collected similarly."*
+Year of vaccination was collected similarly."*
 
 *We acknowledge that this is a limitation of our study for which ideally a validation study would be carried out.
-To highlight this we included the following in the discussion (amended to include year of vaccination).*
+To highlight this, we included the following in the discussion (amended to include year of vaccination).*
 
@@
 
-*"All-cause mortality was defined using the overall outcome recorded in ETS, this is based on follow to 36 months starting from date of starting treatment.
+*"All-cause mortality was defined using the overall outcome recorded in ETS, this is based on 36 months of folow up starting from date of starting treatment.
 Where the treatment start date was not available the notification date was used if appropriate.
@@
 The tables are presented seperately to make it clear that not all outcomes were considered in the same model.
-Whilst this is our prefered method of presenting this data we would be happy to combine these tables into a single table if the reviewer felt this would improve interpretability.*
+Whilst this is our prefered method of presenting this data we would be happy to combine these tables into a single table if required.*
 
@@
 
-*Unfortunately causes of death that were not due to TB, or related to TB were not recorded in our dataset, so we are unable to add this information.*
+*Unfortunately causes of death that were not due to TB, or related to TB, were not recorded in our the Enhanced TB surveillance system, so we are unable to add this information.*
 
@@
 
-*We agree that the first paragraph of the discussion summarises the results of this study.
-We feel that this is required as this is the only summary of our findings in the paper, with results presented by outcome in the results section.
-A large part of the relates to the limitations of the study, and the approaches that we took to mitigate them, that we felt needed to be discussed in order to put our results into context.
+*We feel that the first paragraph of the discussion is required as this is the only summary of our findings in the paper, with results presented by outcome in the results section.
+A large part of the discussion relates to the limitations of the study, and the approaches that we took to mitigate them, that we felt needed to be discussed in order to put our results into context.
 We feel that paragraphs four and five of the discussion interpret the results from this study and relate to the published literature.
-However, we would be very happy to add any suggested literature references to the paper, as additional context can only improve the interpretation of our findings.
-*
+However, we would be happy to add any specific literature references to the paper, as additional context can only improve the interpretation of our findings.*
 
@@
 
-*We agree that the proportion of missing data is a significant limitation of this study (as with many studies using routine surveillance data).
+*We acknowledge that the proportion of missing data is a significant limitation of this study (as with many studies using routine surveillance data).
 As outlined above, to explore the impact of this we repeated all analyses using multiple imputation.

### 2019-03-04 output/review/vaccine-review-response.Rmd
comments up to date

@@
   *"BCG’s primary mode of action is to directly prevent the development of active, symptomatic disease.
-Its efficacy in adults is context specific, with estimates ranging between 0% and 78%.[3] It has been shown to highly efficacious in England and there is some evidence that efficacy increases with distance from the equator.[7] Efficacy has been shown to be dependent on previous exposure, with unexposed individuals receiving the greatest benefit.[4] Unlike in adults, BCG has consistently been shown to be highly protective against TB and TB meningitis in children.[5,6] For this reason the majority of countries that use BCG vaccinate at birth.[7,8]"*
+Its efficacy in adults is context specific, with estimates ranging between 0% and 78%.[3] It has been shown to highly efficacious in England and there is some evidence that efficacy increases with distance from the equator.
+Efficacy has been shown to be dependent on previous exposure, with unexposed individuals receiving the greatest benefit.[4] Unlike in adults, BCG has consistently been shown to be highly protective against TB and TB meningitis in children.[5,6] For this reason the majority of countries that use BCG vaccinate at birth.[7,8]"*
 

### 2019-03-04 output/Paper.Rmd
comments up to date

@@
 
-**WORD COUNT:** **Title** 19 (20), **Abstract** 220 (300), **Paper** 2790 (5000)
+**WORD COUNT:** **Title** 19 (20), **Abstract** 220 (300), **Paper** 3190 (5000)
 
@@
 
-- We explored the identified associations by age and time since vaccination.
+- There was weak evidence of associations between TB outcomes and age at, or years since, vaccination.
 
@@
 Bacillus Calmette–Guérin (BCG) is one of the mostly widely-used vaccines and the only vaccine that protects against tuberculosis (TB) disease.
-BCG was first used in humans in 1921 and was introduced into the WHO Expanded Program on Immunization in 1974.[@The2004] BCG vaccination has been controversial due to its variable efficacy and possibility of causing a false positive result with the standard skin test for TB.[@Zwerling2011a] However, the lack of a more effective vaccine and the emergence of drug-resistant TB strains means that BCG remains the best available vaccination for TB.
+BCG was first used in humans in 1921 and was introduced into the WHO Expanded Program on Immunization in 1974.[@The2004] BCG vaccination has been controversial due to its variable efficacy and possibility of causing a false positive result with the standard skin test for TB.[@Zwerling2011a] However, the lack of a more effective vaccine and the emergence of drug-resistant TB strains means that BCG vaccination remains an important tool for reducing TB incidence and mortality rates.
 
 BCG’s primary mode of action is to directly prevent the development of active, symptomatic disease.
-Its efficacy in adults is context specific, with estimates ranging between 0% and 78%.[@Mangtani2014a] Efficacy has been shown to be dependent on previous exposure, with unexposed individuals receiving the greatest benefit.[@Barreto2014a] Unlike in adults, BCG has consistently been shown to be highly protective against TB and TB meningitis in children.[@Rodrigues1993; @Colditz1994] For this reason the majority of countries that use BCG vaccinate at birth.[@Zwerling2011; @WHOBCG2018] Adult vaccination is no longer common in the UK, where universal BCG vaccination of adolescents was stopped in 2005 in favour of a targeted neonatal programme aimed at high risk children.
+Its efficacy in adults is context specific, with estimates ranging between 0% and 78%.[@Mangtani2014a] It has been shown to highly efficacious in England and there is some evidence that efficacy increases with distance from the equator.
+Efficacy has been shown to be dependent on previous exposure, with unexposed individuals receiving the greatest benefit.[@Barreto2014a] Unlike in adults, BCG has consistently been shown to be highly protective against TB and TB meningitis in children.[@Rodrigues1993; @Colditz1994] For this reason the majority of countries that use BCG vaccinate at birth.[@Zwerling2011; @WHOBCG2018] Adult vaccination is no longer common in the UK, where universal BCG vaccination of adolescents was stopped in 2005 in favour of a targeted neonatal programme aimed at high risk children.
 
-Vaccination policy has been primarily based on reducing the incidence of active TB and little attention has been given to any additional effects of BCG.[@Fine2005a; @Teo2006] There is some evidence that BCG vaccination induces innate immune responses which may provide non-specific protection,[@Kleinnijenhuis2012] TB patients with BCG scars were found to respond better to treatment with earlier sputum smear conversion,[@Jeremiah2010] and there is evidence to suggest that BCG vaccination is associated with reduced all-cause neonatal mortality[@Garly2003; @Higgins] and both reduced TB[@Abubakar2013] and all-cause[@Rieckmann2016] mortality in the general population.
+Vaccination policy has been primarily based on reducing the incidence of TB disease, and mitigating disease severity, with little attention having been given to any additional effects of BCG vaccination on TB outcomes.[@Fine2005a; @Teo2006] There is some evidence that BCG vaccination induces innate immune responses which may provide non-specific protection,[@Kleinnijenhuis2012] TB patients with BCG scars were found to respond better to treatment with earlier sputum smear conversion,[@Jeremiah2010] and there is evidence to suggest that BCG vaccination is associated with reduced all-cause neonatal mortality[@Garly2003; @Higgins] and both reduced TB[@Abubakar2013] and all-cause[@Rieckmann2016] mortality in the general population.
 Given that the immunology behind TB immunity is not well understood these findings suggest that BCG may play a more important role in improving TB outcomes than previously thought.
@@
 BCG vaccination status and year of vaccination have been collected since 2008.
-We considered five TB outcomes which were selected due to their association with increased case infectiousness or poor consequences for patients (`r tabRef('OutcomeDef')`).
+The outcomes we considered were: all-cause mortality, death due to TB (in those who died), recurrent TB, pulmonary disease, and sputum smear status.
+These outcomes were selected based on: their availability in the ETS; evidence from the literature of prior associations with BCG vaccination; associations with increased case infectiousness; or severe outcomes for patients.
 
-```{r outcome-defs, warning = FALSE, message = FALSE, echo = FALSE}
-OutcomeDefsTab <- data_frame(Outcome = c(
-                                         'All-cause mortality',
-                                         'Death due to TB (in those who died)',
-                                         'Recurrent TB',
-                                         'Pulmonary disease', 
-                                         'Sputum smear status - positive'
-                                         ),
-                             Definition = c('All-cause mortality was defined using the overall outcome recorded in ETS, this is based on follow up at 12, 24, and 36 months (or until treatment completion).
-If a case dies within 12 months of completing treatment, from TB or from a cause related to TB then their overall outcome will be updated in the ETS.
-Those that were lost to follow up, or not evaluated were treated as missing.',
-                                            'For cases with a known cause of death, death due to TB is defined as those that died from TB, or where TB had contributed to their death.',
-                                           'TB cases who had recurrent episodes were identified in the dataset.',
-                                            'Cases that present with pulmonary TB (regardless of extra-pulmonary TB status).',
-                                            'Status of sputum sample tested for Acid-Fast Bacilli.'))
-
-OutcomeDefsTab %>% 
-  kable(align = 'll', caption = tabRef('OutcomeDef', 'Summary of outcome definitions and rationale for inclusion '))
-```
+All-cause mortality was defined using the overall outcome recorded in ETS, this is based on 36 months of follow up starting from date of starting treatment.
+Where the treatment start date was not available the notification date was used if appropriate.
+The date of death was validated against Office for National Statistics (ONS) data.
+Those that were lost to follow up, or not evaluated were treated as missing.
+In cases with a known cause of death, death due to TB was defined as those that died from TB, or where TB had contributed to their death.
+This was assessed using the cases death certificates.
+TB cases who had recurrent episodes were identified in the dataset using their NHS number.
+Positive sputum smear status was given to cases that had a sputum sample shown to contain Acid-Fast Bacilli.
+A positive sputum smear status indicates that cases are more likely to be infectious.
+Cases were defined as having pulmonary TB if a positive sputum smear sample was recorded or if a positive culture was grown from a pulmonary laboratory specimen.
+People with laryngeal TB were defined as having pulmonary TB, and people with miliary TB were not excluded.
 
@@
 
-BCG status was taken directly from the ETS.
-Years since BCG vaccination was defined as year of notification minus year of vaccination and categorised into two groups (0 to 10 and 11+ years), based on evidence that the average duration of BCG protection is 10-15 years.[@Abubakar2013] Age at vaccination is defined in the online supplementary information.
+BCG status was collected during the assessment of TB cases and then entered in the ETS.
+When available BCG vaccination records are used, although often these are not available and so patient recall, or the presence of a scare is relied on.
+When cases are uncertain, and there is no evidence of a scar, no BCG status is given.
+Year of vaccination was collected similarly.
+Years since BCG vaccination was defined as year of notification minus year of vaccination and categorised into two groups (0 to 10 and 11+ years).
+This was based on: evidence that the average duration of BCG protection is 10-15 years;[@Abubakar2013] increasing recall bias with time since vaccination, and any association between years since vaccination and TB outcomes may be non-linear.
+Age at vaccination is defined in the online supplementary information.
 
@@
 We repeated the analysis excluding duplicate recurrent cases, and restricting the study population to those eligible for the BCG schools scheme (defined as UK born cases that were aged 14 or over in 2004) to assess the comparability of the BCG vaccinated and unvaccinated populations.
-To mitigate the impact of missing data we used multiple imputation, with the MICE package.[@VanBuuren2011] We imputed 50 data sets (for 20 iterations) using all variables included in the analysis as predictors along with Public Health England centre.
+To mitigate the impact of missing data we used multiple imputation, with the MICE package.[@VanBuuren2011] We imputed 50 data sets (for 20 iterations) using ) using all outcome and explanatory variables included in the analysis as predictors along with Public Health England centre.
 The model results were pooled using the small sample method,[@Barnard1999] and effect sizes compared with those from the main analysis.
@@
 `r ukborn_prop_vcbirth[1]` respectively, P: `r ukborn_test_vcbirth`).
-See online `r SuptabRef('outcomesum')` for the breakdown of outcome variables and `r SuptabRef('ConfounderStrat')` for the breakdown of confounding variables.
+See online `r tabRef('outcomesum')` for the breakdown of outcome variables and `r tabRef('ConfounderStrat')` for the breakdown of confounding variables.
 
@@
 }
+```
+
+```{r outcome summary, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
+bake(file  = paste0('../data/', folder, '/outcome_demo_tab_bcg.rds'), {
+#Compile summary table
+var <- c(outcomes)
+var_words <- c(outcomes_words)
+categories <- list(c('Yes', 'No', NA))
+names(categories[[1]]) <- c('Vaccinated', 'Unvaccinated', 'Unknown vaccine status')
+names(categories) <- 'bcgvacc'
+
+  outcome_sum <- bind_cols(demographic_summary(var, var_words, uk_not), variable_summary(categories, var, var_words, uk_not))
+}) -> outcome_sum
+
+## Pretty results table
+outcome_sum %>% 
+    pretty_sum_results_table(col_names = c("Outcome", colnames(outcome_sum)[-1]),
+                             footer = paste0('{% all cases}(% complete within vaccine status)[% *complete within category*], * ',  tb_death_footer), 
+                             label = 'outcomesum', 
+                             caption = 'Outcomes for individuals in England notified with tuberculosis between 2009-2015,  stratified by BCG vaccination status.', 
+                             format.args = (list(strip.white = FALSE)),
+                             cap_fun = tabRef)
+```
+
+```{r demographic summary, echo=FALSE, message=FALSE}
+bake(file = paste0('../data/', folder, '/demographic_tab_bcg.rds'), {
+  #Compile summary table
+var <- c(covariates)
+var_words <- c(covariates_words)
+categories <- list(c('Yes', 'No', NA))
+names(categories[[1]]) <- c('Vaccinated', 'Unvaccinated', 'Unknown vaccine status')
+names(categories) <- 'bcgvacc'
+
+demo_sum <- bind_cols(demographic_summary(var, var_words, uk_not), variable_summary(categories, var, var_words, uk_not))
+}) -> demo_sum_bcg
+
+demo_sum_bcg %>% 
+              pretty_sum_results_table(col_names = c("Confounder", colnames(demo_sum_bcg)[-1]),
+                                       footer = paste0('{% all cases}(% complete within vaccine status)[% *complete within category*], * ',  tb_death_footer), 
+                                       label = 'ConfounderStrat', 
+                                       caption = 'Confounders for individuals in England notified with tuberculosis between 2009-2015,  stratified by BCG vaccination status.',
+                                       cap_fun = tabRef)
 ```
@@
 We were not able to adjust for either tuberculin skin test (TST) stringency, or the latitude effect, although we were able to adjust for UK birth status.[@Roy2014b] However, the bias induced by these confounders is likely to be towards the null, meaning that our effect estimates are likely to be conservative.
-Finally, BCG vaccination status may be subject to misclassification due to recall bias; validation studies of the recording of BCG status in the ETS would be required to assess this.
+Finally, BCG vaccination status, and year of vaccination, may be subject to misclassification due to recall bias; validation studies of the recording of BCG status in the ETS would be required to assess this.
 
@@
 However, due to the continuous collection of the surveillance data used in this analysis, this study could be repeated once additional data have been collected.
+If this study were to be repeated with a larger sample size particular attention should be given to the functional form of any decay in protection from negative TB outcomes.
+Additionally, a larger sample size would allow investigation of the associations identified between TB outcomes and BCG vaccination stratified by pulmonary, extrapulmonary, and disseminated TB disease.
 The results from this study require validation in independent datasets and the analysis should be reproducible in other low incidence countries that have similarly developed surveillance systems.
@@
 ```
-
-**Outcome variables stratified by BCG vaccination status**
-
-```{r outcome summary, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
-bake(file  = paste0('../data/', folder, '/outcome_demo_tab_bcg.rds'), {
-#Compile summary table
-var <- c(outcomes)
-var_words <- c(outcomes_words)
-categories <- list(c('Yes', 'No', NA))
-names(categories[[1]]) <- c('Vaccinated', 'Unvaccinated', 'Unknown vaccine status')
-names(categories) <- 'bcgvacc'
-
-  outcome_sum <- bind_cols(demographic_summary(var, var_words, uk_not), variable_summary(categories, var, var_words, uk_not))
-}) -> outcome_sum
-
-## Pretty results table
-outcome_sum %>% 
-    pretty_sum_results_table(col_names = c("Outcome", colnames(outcome_sum)[-1]),
-                             footer = paste0('{% all cases}(% complete within vaccine status)[% *complete within category*], * ',  tb_death_footer), 
-                             label = 'outcomesum', 
-                             caption = 'Outcomes for individuals in England notified with tuberculosis between 2009-2015,  stratified by BCG vaccination status.', 
-                             format.args = (list(strip.white = FALSE)),
-                             cap_fun = SuptabRef)
-```
-
-**Confounding variables stratified by BCG vaccination status**
-
-```{r demographic summary, echo=FALSE, message=FALSE}
-bake(file = paste0('../data/', folder, '/demographic_tab_bcg.rds'), {
-  #Compile summary table
-var <- c(covariates)
-var_words <- c(covariates_words)
-categories <- list(c('Yes', 'No', NA))
-names(categories[[1]]) <- c('Vaccinated', 'Unvaccinated', 'Unknown vaccine status')
-names(categories) <- 'bcgvacc'
-
-demo_sum <- bind_cols(demographic_summary(var, var_words, uk_not), variable_summary(categories, var, var_words, uk_not))
-}) -> demo_sum_bcg
-
-demo_sum_bcg %>% 
-              pretty_sum_results_table(col_names = c("Confounder", colnames(demo_sum_bcg)[-1]),
-                                       footer = paste0('{% all cases}(% complete within vaccine status)[% *complete within category*], * ',  tb_death_footer), 
-                                       label = 'ConfounderStrat', 
-                                       caption = 'Confounders for individuals in England notified with tuberculosis between 2009-2015,  stratified by BCG vaccination status.',
-                                       cap_fun = SuptabRef)
-```
-
-##### PAGEBREAK
 
@@
                               label = 'BCGoutsumsens', 
-                              caption =   paste0('Summary of associations between BCG vaccination and all outcomes; cases that have no recurrent flag in the ETS (', nrow(uk_not %>% filter(is.na(recurrent))), '), and cases that would have been eligible for the BCG schools scheme (', elg_school_stats_sum['cases eligible for the schools scheme'] %>% as.numeric, ').
+                              caption =   paste0('Summary of associations between BCG vaccination and all outcomes; cases that have no recurrent flag in the ETS (n = ', nrow(uk_not %>% filter(is.na(recurrent))), '), and cases that would have been eligible for the BCG schools scheme (n = ', elg_school_stats_sum['cases eligible for the schools scheme'] %>% as.numeric, ').
 Those defined to be eligible for the schools scheme are the UK born, that were aged 14 or over in 2004'),
@@
 }
-```
-
-#####PAGEBREAK
-
-**Estimated power**
-
-Power estimates were calculated in both the univariable and multivariable datasets for all outcomes.
-This represents an over estimate of the statistical power, as only a single exposure variable was accounted for, and age at vaccination has been simplified to a binary vaccinated at birth variable (`r SuptabRef('power_table')`).
-
-```{r Estimate stat power of study, include=FALSE}
-bake(file = paste0('../data/', folder, '/estimate_study_power.rds'), {
-#source paper calc
-source('../functions/Power_calculation.R')
-
-#power_table
-power_table = function(var, variables_words, outcomes, covariates, outcomes_words,data)
- {
-
-if (var == 'VaccatBirth') 
-   {
-        data$VaccatBirth <- ifelse(!is.na(data$AgeatBCG), ifelse(data$AgeatBCG < 1, 'Yes', 'No'), NA) %>% factor
-        var_words <- "Age since vaccination (vaccination at birth vs. older)"
-   }else{
-       var_words <- variables_words[var]
-    }
-
-  uni <- outcomes %>% map_df( function(.) power_calc(data, ., var, cov = '', outcomes_words)) %>% select(Outcome, Power)
-
-  #Run for adjusted data
-  adj <- outcomes %>% map_df( function(.) power_calc(data, ., var, cov = covariates, outcomes_words)) %>% select(Power)
-
-  #Construct table
-  power_table <- paste0(uni$Power,'%') %>% as.matrix %>% t
-  power_table <- rbind(power_table, paste0(adj$Power, '%') %>% as.matrix %>% t)
-  power_table <- bind_cols(as_data_frame(c('&nbsp;&nbsp;Univariable', '&nbsp;&nbsp;Multivariable')), as_data_frame(power_table))
-  colnames(power_table) <- c('Analysis', outcomes_words)
-
-  power_table[3,] <- c(var_words, rep('', length(outcomes_words)))
-  power_table <- bind_rows(power_table[3,], power_table[-3,])
-  return(power_table)
-  }
-
-  # Power for BCG vaccination -----------------------------------------------
-  power_table_out <- power_table('bcgvacc', variables_words, outcomes, covariates, outcomes_words, uk_not)
-
-  #Power for Years since vaccination
-
-  power_table_out <- bind_rows(power_table_out, power_table('yrsinceBCG', variables_words, outcomes, covariates, outcomes_words, uk_not))
-
-  # Power for Vaccination at birth -----------------------------------------------
-
-  power_table_out <- bind_rows(power_table_out, power_table('VaccatBirth', variables_words, outcomes, covariates, outcomes_words, uk_not))
-
-}) -> power_table
-
-```
-
-```{r power table, echo=FALSE, message=FALSE, eval = TRUE}
-power_table %>% 
- pretty_sum_results_table(label = 'power_table', 
-                           col_names = c('Exposure', colnames(.)[-1]),
-                           caption = 'Summary of the estimated power for each analysis, for both the univariable and multivariable data sets, with alpha set as 0.05.
-Power estimates assume a single exposure variable, and age at vaccination has been simplified into a binary vaccinated at birth variable',
-                           cap_fun = SuptabRef
- )
 ```

### 2019-03-19 output/review/vaccine-review-response.Rmd
initial colab changes

@@
 
-We enclose a revised version of the manuscript (clean and with tracked changes), which addreses the reviewer’s comments and suggestions.
+We enclose a revised version of the manuscript (clean and with tracked changes), which addresses the reviewer’s comments and suggestions.
 The reviewer’s comments are shown below with our response to each comment in italics.
@@
 Of those that died and had known vaccination status, a further 38% of people did not have a known cause of death.
-Because this study only analyzed cases with complete data, conclusions are based on a subset of TB cases that may not be representative of the population as a whole.
+Because this study only analysed cases with complete data, conclusions are based on a subset of TB cases that may not be representative of the population as a whole.
 
@@
 *We agree that this section should focus on the results from this study.
-Based on the requirments laid out on the Vaccine help page (https://www.elsevier.com/authors/journal-authors/highlights), we have adapted the final bullet point as follows:*
+Based on the requirements laid out on the Vaccine help page (https://www.elsevier.com/authors/journal-authors/highlights), we have adapted the final bullet point as follows:*
 
@@
 We did not give precise estimates for the efficacy of the BCG vaccine as efficacy has been shown to be context specific.
-However, we updated paragraph two of the introduction to explicitly state that the BCG vaccine has been shown to be highly effective in England, and that its effectivness may be correlated with distance from the equator.
+However, we updated paragraph two of the introduction to explicitly state that the BCG vaccine has been shown to be highly effective in England, and that its effectiveness may be correlated with distance from the equator.
 This now reads as follows.*
@@
 Its efficacy in adults is context specific, with estimates ranging between 0% and 78%.[3] It has been shown to highly efficacious in England and there is some evidence that efficacy increases with distance from the equator.
-Efficacy has been shown to be dependent on previous exposure, with unexposed individuals receiving the greatest benefit.[4] Unlike in adults, BCG has consistently been shown to be highly protective against TB and TB meningitis in children.[5,6] For this reason the majority of countries that use BCG vaccinate at birth.[7,8]"*
+Efficacy has been shown to be dependent on previous exposure, with unexposed individuals receiving the greatest benefit.[4] Unlike in adults, BCG has consistently been shown to be highly protective against TB and TB meningitis in children.[5,6] For this reason the majority of countries that use BCG, vaccinate at birth.[7,8]"*
 
@@
 
-*We agree that this needed clarifiction.
+*We agree that this needed clarification.
 As part of our response to your comment below lines 18-19 we have added an explanation of additional effects in the context of TB outcomes.
@@
 
-*We agree that this sentance was unclear. We have reworded it to the following:*
+*We agree that this sentence was unclear. We have reworded it to the following:*
 
@@
 
-*"BCG status was collected during the assessment of TB cases and then entered in the ETS.
-When available BCG vaccination records are used, although often these are not available and so patient recall, or the presence of a scare is relied on.
+*"BCG status was collected and recorded in ETS by case managers.
+Information on BCG vaccination status may have come from vaccination records, patient recall or the presence of a scar.
 When cases are uncertain, and there is no evidence of a scar, no BCG status is given.
@@
 
-*We updated the methods section to highlight that all outcome and explanatory variables were included in the multple imputation approach:*
+*We updated the methods section to highlight that all outcome and explanatory variables were included in the multiple imputation approach:*
 
@@
 *"The outcomes we considered were: all-cause mortality, death due to TB (in those who died), recurrent TB, pulmonary disease, and sputum smear status.
-These outcomes were selected based on: their availability in the ETS; evidence from the literature of prior associations with BCG vaccination; associations with increased case infectiousness; or severe outcomes for patients."*"
+These outcomes were selected based on: their availability in the ETS; evidence from the literature of prior associations with BCG vaccination; associations with increased case infectiousness; or severe outcomes for patients."*
 
-*"All-cause mortality was defined using the overall outcome recorded in ETS, this is based on 36 months of folow up starting from date of starting treatment.
+*"All-cause mortality was defined using the overall outcome recorded in ETS, this is based on up to 36 months of follow up starting from date of starting treatment.
+Follow up ends when a case is recorded as completing treatment, with treatment status evaluated at 12, 24, and 36 months from starting treatment.
 Where the treatment start date was not available the notification date was used if appropriate.
@@
 In cases with a known cause of death, death due to TB was defined as those that died from TB, or where TB had contributed to their death.
-This was assessed using the cases death certificates.
-TB cases who had recurrent episodes were identified in the dataset using their NHS number.
+Cause of death was recorded by case managers.
+TB cases who had recurrent episodes were identified using probabilistic matching.
 Positive sputum smear status was given to cases that had a sputum sample shown to contain Acid-Fast Bacilli.
 A positive sputum smear status indicates that cases are more likely to be infectious.
-Cases were defined as having pulmonary TB if a positive sputum smear sample was recorded or if a positive culture was grown from a pulmonary laboratory specimen.
-People with laryngeal TB were defined as having pulmonary TB, and people with miliary TB were not excluded."*
+Cases were defined as having pulmonary TB if a positive sputum smear sample was recorded, if a positive culture was grown from a pulmonary laboratory specimen, or if they were clinically assessed as having pulmonary TB."*
 
@@
 *We have moved both S1 and S2 into the main body of the text.
-The tables are presented seperately to make it clear that not all outcomes were considered in the same model.
-Whilst this is our prefered method of presenting this data we would be happy to combine these tables into a single table if required.*
+The tables are presented separately to make it clear that not all outcomes were considered in the same model.
+Whilst this is our preferred method of presenting this data we would be happy to combine these tables into a single table if required.*
 
@@
 These can be found at the bottom of each table as a footnote.
-For analyses that use mutltiple imputation, all samples with a given outcome were used.
+For analyses that use multiple imputation, all samples with a given outcome were used.
 For BCG vaccination this is all samples.
@@
 
-*We initially considered using a continuous varibale but ultimately decided to use a binary categorical variable for years since vaccination.
+*We initially considered using a continuous variable but ultimately decided to use a binary categorical variable for years since vaccination.
 This was primarily motivated by concerns over the linearity of any relationship between TB outcomes and years since vaccination.
@@
 We were also concerned the recall bias of year of vaccination would increase with years since vaccination, categorising reduces the impact of this bias.
-We have now highlighted this in the manuscript with the following statement in the methods secton: *
+We have now highlighted this in the manuscript with the following statement in the methods section: *
 
 *"Years since BCG vaccination was defined as year of notification minus year of vaccination and categorised into two groups (0 to 10 and 11+ years).
-This was based on: evidence that the average duration of BCG protection is 10-15 years;[15] increasing recall bias with time since vaccination, and any association between years since vaccination and TB outcomes may be non-linear."*
+This was based on: evidence that the average duration of BCG protection is at least 10-15 years;[15] increasing recall bias with time since vaccination, and any association between years since vaccination and TB outcomes may be non-linear."*
 
@@
 *We agree that there could be additional outcome variables included in this study.
-However, we could only include outcomes that were included in the routine surveliance dataset on which this study was based.
+However, we could only include outcomes that were included in the routine surveillance dataset on which this study was based.
 Given the availability of outcome data, we selected outcomes for further evaluation based on  evidence from the literature of prior associations with BCG vaccination, associations with TB infectiousness, or poor consequences for patients.

### 2019-03-19 output/Paper.Rmd
initial colab changes

@@
 
-Maeve K Lalor, TB section, National Infection Service, Public Health England, London, UK
+Maeve K Lalor, TB Unit, T.A.R.G.E.T, National Infection Service (NIS), Public Health England, London, UK
 
-Dominik Zenner, Respiratory Diseases Department, Centre for Infectious Disease Surveillance and Control (CIDSC), National Infections Service, Public Health England, London, UK
+Dominik Zenner, Institute for Global Health, University College London, London, UK
 
-Colin Campbell, National Infection Service, Public Health England, London, UK
+Colin Campbell, TB Unit, T.A.R.G.E.T, National Infection Service (NIS), Public Health England, London, UK
 
@@
 
-**WORD COUNT:** **Title** 19 (20), **Abstract** 220 (300), **Paper** 3190 (5000)
+**WORD COUNT:** **Title** 19 (20), **Abstract** 220 (300), **Paper** 3196 (5000)
 

### 2019-03-21 output/review/vaccine-review-response.Rmd
added paper changes based on review

@@
 
-Thank you for sending review reports for our manuscript. The reviewer’s comments were useful, and have improved the paper.
+Thank you for sending review reports for our manuscript. The reviewer’s comments were useful and have improved the paper.
 
@@
 The tables are presented separately to make it clear that not all outcomes were considered in the same model.
-Whilst this is our preferred method of presenting this data we would be happy to combine these tables into a single table if required.*
+Whilst this is our preferred method of presenting this data, we would be happy to combine these tables into a single table if required.*
 

### 2019-03-21 output/Paper.Rmd
added paper changes based on review

@@
 Its efficacy in adults is context specific, with estimates ranging between 0% and 78%.[@Mangtani2014a] It has been shown to highly efficacious in England and there is some evidence that efficacy increases with distance from the equator.
-Efficacy has been shown to be dependent on previous exposure, with unexposed individuals receiving the greatest benefit.[@Barreto2014a] Unlike in adults, BCG has consistently been shown to be highly protective against TB and TB meningitis in children.[@Rodrigues1993; @Colditz1994] For this reason the majority of countries that use BCG vaccinate at birth.[@Zwerling2011; @WHOBCG2018] Adult vaccination is no longer common in the UK, where universal BCG vaccination of adolescents was stopped in 2005 in favour of a targeted neonatal programme aimed at high risk children.
+Efficacy has been shown to be dependent on previous exposure, with unexposed individuals receiving the greatest benefit.[@Barreto2014a] Unlike in adults, BCG has consistently been shown to be highly protective against TB and TB meningitis in children.[@Rodrigues1993; @Colditz1994] For this reason the majority of countries that use BCG, vaccinate at birth.[@Zwerling2011; @WHOBCG2018] Adult vaccination is no longer common in the UK, where universal BCG vaccination of adolescents was stopped in 2005 in favour of a targeted neonatal programme aimed at high risk children.
 
 Vaccination policy has been primarily based on reducing the incidence of TB disease, and mitigating disease severity, with little attention having been given to any additional effects of BCG vaccination on TB outcomes.[@Fine2005a; @Teo2006] There is some evidence that BCG vaccination induces innate immune responses which may provide non-specific protection,[@Kleinnijenhuis2012] TB patients with BCG scars were found to respond better to treatment with earlier sputum smear conversion,[@Jeremiah2010] and there is evidence to suggest that BCG vaccination is associated with reduced all-cause neonatal mortality[@Garly2003; @Higgins] and both reduced TB[@Abubakar2013] and all-cause[@Rieckmann2016] mortality in the general population.
-Given that the immunology behind TB immunity is not well understood these findings suggest that BCG may play a more important role in improving TB outcomes than previously thought.
+Given that the immunology behind TB immunity is not fully understood these findings suggest that BCG may play a more important role in improving TB outcomes than previously thought.
 We aimed to quantify the effects of BCG vaccination on outcomes for individuals with notified TB in England using routinely collected surveillance data to provide evidence for appropriate public health action and provision.
@@
 
-All-cause mortality was defined using the overall outcome recorded in ETS, this is based on 36 months of follow up starting from date of starting treatment.
+All-cause mortality was defined using the overall outcome recorded in ETS, this is based on up to 36 months of follow up starting from date of starting treatment.
+Follow up ends when a case is recorded as completing treatment, with treatment status evaluated at 12, 24, and 36 months from starting treatment.
 Where the treatment start date was not available the notification date was used if appropriate.
@@
 In cases with a known cause of death, death due to TB was defined as those that died from TB, or where TB had contributed to their death.
-This was assessed using the cases death certificates.
-TB cases who had recurrent episodes were identified in the dataset using their NHS number.
+Cause of death was recorded by case managers.
+TB cases who had recurrent episodes were identified using probabilistic matching.
 Positive sputum smear status was given to cases that had a sputum sample shown to contain Acid-Fast Bacilli.
 A positive sputum smear status indicates that cases are more likely to be infectious.
-Cases were defined as having pulmonary TB if a positive sputum smear sample was recorded or if a positive culture was grown from a pulmonary laboratory specimen.
-People with laryngeal TB were defined as having pulmonary TB, and people with miliary TB were not excluded.
+Cases were defined as having pulmonary TB if a positive sputum smear sample was recorded, if a positive culture was grown from a pulmonary laboratory specimen, or if they were clinically assessed as having pulmonary TB.
 
@@
 
-BCG status was collected during the assessment of TB cases and then entered in the ETS.
-When available BCG vaccination records are used, although often these are not available and so patient recall, or the presence of a scare is relied on.
+BCG status was collected and recorded in ETS by case managers.
+Information on BCG vaccination status may have come from vaccination records, patient recall or the presence of a scar.
 When cases are uncertain, and there is no evidence of a scar, no BCG status is given.
@@
 Years since BCG vaccination was defined as year of notification minus year of vaccination and categorised into two groups (0 to 10 and 11+ years).
-This was based on: evidence that the average duration of BCG protection is 10-15 years;[@Abubakar2013] increasing recall bias with time since vaccination, and any association between years since vaccination and TB outcomes may be non-linear.
+This was based on: evidence that the average duration of BCG protection is at least 10-15 years;[@Abubakar2013] increasing recall bias with time since vaccination, and any association between years since vaccination and TB outcomes may be non-linear.
 Age at vaccination is defined in the online supplementary information.

### 2019-03-26 scripts/packages.R
added sensitivity analysis saving

@@ -17,8 +17,8 @@ pkgLoad <- function(x)
 
 ## Set up checkpoint versioning
 pkgLoad("checkpoint")
-if (!dir.exists("./.checkpoint")) { dir.create("./.checkpoint")}
-checkpoint("2018-06-19", scanForPackages = TRUE, checkpointLocation = "./.checkpoint")
+if (!dir.exists("../.checkpoint")) { dir.create("../.checkpoint")}
+checkpoint("2018-06-19", scanForPackages = TRUE, checkpointLocation = "../.checkpoint")
 
 ## Install/Load all other packages
 pck <- lapply(package_names, pkgLoad)

### 2019-05-10 README.Rmd
added reviewer responses and update package loading
added binder

@@
 
+[![badge](https://img.shields.io/badge/Launch-Analysis-lightblue.svg)](https://mybinder.org/v2/gh/seabbs/ExploreBCGOnOutcomes/master?urlpath=rstudio)
 [![Paper](https://img.shields.io/badge/Paper-10.1101/366476-lightgrey.svg)](https://doi.org/10.1101/366476)
@@
 
-- Install the required packages using the following script. Versioned to "2018-06-19".
+- Install the required packages using the following script. Versioned to "2018-06-19" using the `checkpoint` package.
 
@@
 
-This analysis can be run in a docker container based on the tidyverse docker image. To run the docker image run:
+To ensure full reproducibility, this analysis can be run in a docker container based on the tidyverse docker image. To run the docker image run:
 

### 2019-05-10 output/review/vaccine-review-response-2.Rmd
proposed changes for second round review

---
title: "Reviewer Response"
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
---

Dear Dr. Orenstein,

Thank you for sending review reports for our manuscript. The reviewer’s comments were again useful and have improved the paper.

We enclose a revised version of the manuscript (clean and with tracked changes), which addresses the reviewer’s comments and suggestions. The reviewer’s comments are shown below with our response to each comment in italics.

Yours sincerely,

Sam Abbott

## Authors' response to Reviewer #1:

The authors have addressed all of my comments.  The additional detail about the study methodology and revisions in the presentation of results represent significant improvements to the manuscript.

*Thank you again for your detailed comments on our manuscript. We are pleased that we have been able to address your concerns and agree that the revisions have improved the paper.*

## Authors' response to Reviewer #3:

The authors have done responded to the previous reviewers' comments as well as can be expected. Despite the use of multiple imputation techniques using MICE, the sheer amount of missing data that are so vital to the core questions being asked is worrisome in terms of outcomes. As the authors note, this is the nature of "real world" studies and there likely is no better way to get better answers.

*Thank you for your helpful comments on our manuscript.*

I have some issues that were not addressed in the original reviews:

1. This is a "pet peeve" of mine. Study authors act and speak as if there has been only one BCG vaccine. I know of no other vaccine that exists in multiple preparations that is spoken of or analyzed as a single vaccine. In fact, there have been dozens of BCG vaccines over the decades and detailed analyses and studies have shown that they vary greatly in many physical and immunological characteristics. The authors of this study also "speak" as if there is only one BCG strain. In fact, just within the U.K. there have been several different strains used over time and given at different ages. In addition, a large proportion of the subjects of this study received their BCG vaccine in another country. At a bare minimum, the authors need to recognize this fact and address it as a study limitation. I could not tell if any analyses were performed comparing country of origin; for several countries the "N" is quite large and the analyses would have some significant power to detect BCG strain differences. I realize that it is not possible to identify the exact strain of BCG vaccine used at different times in different countries, but this might give some insight into possible BCG strain-related differences.

*We agree that this needed clarification as it is rarely discussed. We have addressed this as a study limitation by adding the following to the discussion:*

*"We could also not adjust for the BCG strain each individual may have received, the BCG strain used may vary both temporally and geographically."*

*In this study we did not conduct any analyses comparing country of origin. However, it would be interesting to do a follow up analysis looking at between country variation as a proxy of BCG strain. We have added the following to the final paragraph of our discussion highlighting that this is a potential next step.*

*"Between country variations in the strength of associations could also be explored as a proxy to BCG strain using sub-group analysis."*

2. I like to "follow the data" within the paper but found the tables to be nearly impossible to follow; in fact, I still don't really understand them because of how things are labeled and presented. I apologize if I am just missing something, but I have no idea to what the "Yes" and "No" refer. I have reviewed a lot of papers and if I have to spend this much time trying to decipher the tables, then there is likely a better way to present the data. For example, in Table 1, there is a heading "Death due to TB (in those who died)"; below that are the rows "Yes" and "No". I thought that the yes-no referred to whether they died of TB, but then I got to Table 3 where the numbers next to "Death due to TB (in those who died)" are used. Sorry but I just cannot follow it - makes no sense to me. Perhaps the problem is how things are labeled and formatted but please try to make this easier to understand. In addition the format {% all cases}(% complete within vaccine status)[complete within category] is also very confusing.

*We are sorry that the table formatting was difficult to interpret. We have added the following to the table captions for tables 3 and 4 to clarify the cases and cases with outcome columns:*

*"Cases represents all notifications with complete data and a given BCG status, regardless of outcome. Cases with outcome is similarly defined but includes only cases with the specified outcome."*

*Whilst we agree that the use of "{% all cases}(% complete within vaccine status)[complete within category]" makes the table more complex to interpret we feel that this structure allows us to more clearly detail the missing data present. This is an important consideration as the volume of missing data is important to understand when interpreting the study findings.*

3. In Table 2, please explain to what "IMD rank" refers. This should be presented as part of the table.

*We agree that this was missing from the table. We have added the following definition to the table footer:*

*"Index of Multiple Deprivation (2010) categorised into five groups for England"*

### 2019-05-10 output/Paper.Rmd
added sensitivity analysis saving
proposed changes for second round review

@@
 
+demo_sum_bcg$`Demographic Characteristics` <- stringr::str_replace(demo_sum_bcg$`Demographic Characteristics`, "deprived\\)", "deprived\\)\\*")
+
 demo_sum_bcg %>% 
               pretty_sum_results_table(col_names = c("Confounder", colnames(demo_sum_bcg)[-1]),
-                                       footer = paste0('{% all cases}(% complete within vaccine status)[% *complete within category*], * ',  tb_death_footer), 
+                                       footer = paste0('{% all cases}(% complete within vaccine status)[% *complete within category*], * ', "Index of Multiple Deprivation (2010) categorised into five groups for England"),
                                        label = 'ConfounderStrat', 
@@
                                      label = 'BCGoutsum', 
-                                     caption = 'Summary of associations between BCG vaccination and all outcomes')
+                                     caption = 'Summary of associations between BCG vaccination and all outcomes.
+Cases represents all notifications with complete data and a given BCG status, regardless of outcome.
+Cases with outcome is similarly defined but includes only cases with the specified outcome.')
 ```
@@
 The baseline exposure is vaccination $\u2264 10$ years before diagnosis compared to vaccination $11+$ years before diagnosis.
-Deaths due to TB (in those who died) had insufficient data for effect sizes to be estimated in both the univariable and multivariable analysis.'
+Deaths due to TB (in those who died) had insufficient data for effect sizes to be estimated in both the univariable and multivariable analysis.
+Cases represents all notifications with complete data and a given BCG status, regardless of outcome.
+Cases with outcome is similarly defined but includes only cases with the specified outcome.'
                                      )
@@
 Additionally, a larger sample size would allow investigation of the associations identified between TB outcomes and BCG vaccination stratified by pulmonary, extrapulmonary, and disseminated TB disease.
+Between country variations in the strength of associations could also be explored as a proxy to BCG strain using sub-group analysis.
 The results from this study require validation in independent datasets and the analysis should be reproducible in other low incidence countries that have similarly developed surveillance systems.
@@
                                      label = 'ageatvacoutsum', 
-                                     caption = 'Summary of associations between age at vaccination and all outcomes in individuals who were vaccinated - the baseline exposure is vaccination at birth compared to vaccination from 1 to < 12, 12 to < 16, and 16+ years of age.',
+                                     caption = 'Summary of associations between age at vaccination and all outcomes in individuals who were vaccinated - the baseline exposure is vaccination at birth compared to vaccination from 1 to < 12, 12 to < 16, and 16+ years of age.
+Cases represents all notifications with complete data and a given BCG status, regardless of outcome.
+Cases with outcome is similarly defined but includes only cases with the specified outcome.',
                                      cap_fun = SuptabRef
@@
 
+  saveRDS(bcg_ass_main, file.path('../data', folder, "sensitivity_analysis.rds"))
+
   ## Make table

### 2019-05-28 output/review/vaccine-review-response-2.Rmd
second vaccine review changes

@@
 
-Thank you for sending review reports for our manuscript. The reviewer’s comments were again useful and have improved the paper.
+Thank you for sending review reports for our manuscript.
 
-We enclose a revised version of the manuscript (clean and with tracked changes), which addresses the reviewer’s comments and suggestions.
-The reviewer’s comments are shown below with our response to each comment in italics.
+We have revised the paper in light of the reviewers’ comments. In particular, we have revised the tables for clarity, and improved their captioning.
+
+We enclose a revised version of the manuscript (clean and with tracked changes), which addresses the reviewers' comments and suggestions.
+The reviewers' comments are shown below with our response to each comment in italics.
+
+We hope that the manuscript is now suitable for publication in Vaccine.
 
@@
 
-*Whilst we agree that the use of "{% all cases}(% complete within vaccine status)[complete within category]" makes the table more complex to interpret we feel that this structure allows us to more clearly detail the missing data present.
-This is an important consideration as the volume of missing data is important to understand when interpreting the study findings.*
+*We have also clarified the definition of sputum spear status in Table 1 and UK birth status in Table 2.*
+
+*We agree that the use of the {% all cases}(% complete within vaccine status)[complete within category] structure in Table 1 and Table 2 adds complexity but feel that this is justified as missing data is an important consideration for this study.
+We would welcome suggestions for improvements.*
 
@@
 
-*We agree that this was missing from the table. We have added the following definition to the table footer:*
+*Thank you. We have added the following definition to the table footer:*
 

### 2019-05-28 output/Paper.Rmd
second vaccine review changes

@@
 
+outcome_sum$`Demographic Characteristics` <- stringr::str_replace(outcome_sum$`Demographic Characteristics`, " - positive", "")
+
 ## Pretty results table
@@
 demo_sum_bcg$`Demographic Characteristics` <- stringr::str_replace(demo_sum_bcg$`Demographic Characteristics`, "deprived\\)", "deprived\\)\\*")
+demo_sum_bcg$`Demographic Characteristics` <- stringr::str_replace(demo_sum_bcg$`Demographic Characteristics`, "UK born", "UK birth status")
 

### 2019-07-04 README.Rmd
Update README.Rmd

@@
 [![badge](https://img.shields.io/badge/Launch-Analysis-lightblue.svg)](https://mybinder.org/v2/gh/seabbs/ExploreBCGOnOutcomes/master?urlpath=rstudio)
-[![Paper](https://img.shields.io/badge/Paper-10.1101/366476-lightgrey.svg)](https://doi.org/10.1101/366476)
+[![Preprint](https://img.shields.io/badge/Preprint-10.1101/366476-lightgrey.svg)](https://doi.org/10.1101/366476)
 [![DOI](https://zenodo.org/badge/127124135.svg)](https://zenodo.org/badge/latestdoi/127124135)

### 2019-07-09 README.Rmd
Update README.Rmd

--- 2019-05-10 README.Rmd
@@
 [![badge](https://img.shields.io/badge/Launch-Analysis-lightblue.svg)](https://mybinder.org/v2/gh/seabbs/ExploreBCGOnOutcomes/master?urlpath=rstudio)
-[![Paper](https://img.shields.io/badge/Paper-10.1101/366476-lightgrey.svg)](https://doi.org/10.1101/366476)
+[![Paper](https://img.shields.io/badge/Paper-10.1016/j.vaccine.2019.06.056-lightgreen.svg)](https://doi.org/10.1016/j.vaccine.2019.06.056)
+[![Preprint](https://img.shields.io/badge/Preprint-10.1101/366476-lightgrey.svg)](https://doi.org/10.1101/366476)
 [![DOI](https://zenodo.org/badge/127124135.svg)](https://zenodo.org/badge/latestdoi/127124135)
